### 2014-09-19 paper/include/introduction.tex
Wrote a little

\section{Introduction}
\label{sec:int}
Why is it important to reduce the refreshes?
Briefly mention the methods.

### 2014-09-19 paper/include/techniques.tex
Wrote a little

\section{Intelligent Refresh Techniques} 
\label{sec:int}

Say a little about every technique.
\subsection{Access Recency}

\subsection{Retention}

\subsection{Tolerance}

\subsection{Validity}

### 2014-09-19 paper/include/retention.tex
Fixed authors and tilte

\section{Retention}
\label{sec:ret}
\cite{ACM:dtail}

### 2014-09-19 paper/include/techniques.tex
added text and excluded pdf from git

@@
 
-Say a little about every technique.
 \subsection{Access Recency}
+DRAM accesses imply refresh operations, and so a subsequent refresh to the same row can be postponed.
+The number of rows affected depends on how many different rows are accessed within the maximum refresh period, which may be few for many workloads.
+Ghosh et al.
+[9] propose Smart Refresh, which divides the refresh period into phases and maintains a per-row timeout counter in the MC.
+The MC decrements the counters each phase and issues a RAS-only refresh when a counter hits zero.
 
 \subsection{Retention}
+Retention time refers to the period during which cells hold valid values as charge gradually leaks.
+Process variation [10][16] causes the retention time of DRAM cells to vary across the chip.
+Note that approaches that ex-
+ploit R information are insensitive to system workloads and global memory usage, which makes them attractive ompo
+nents for a refresh-optimized memory subsystem.
 
 \subsection{Tolerance}
+Applications like games, media processing, machine learning, and unstructured information analysis tolerate errors in portions of their data and still produce acceptably accurate results.
+Approximate computation [31, 8] exploits this approximate data to realize tradeoffs among performance, energy, and accuracy.
+Cells containing such error-tolerant T data need not be refreshed as often as those containing critical data.
+How many cells fall into this category depends on application characteristics.
 
 \subsection{Validity}
+If the OS has not allocated the page frame, its data are meaningless, and refreshes to it are wasteful. Severa
+l software approaches thus attempt to trigger refreshes only for rows with valid data.
+The effectiveness of schemes using such V refresh information is sensitive to the total memory usage of the system.

### 2014-09-19 paper/include/accessrecency.tex

\section{Access Recency}
\label{sec:acc}

Papers that use this technique:

smart
refrint
P. Emma et al.

### 2014-09-19 paper/include/accessrecency.tex
cite added
explained techniques

@@
 
-smart
-refrint
+smart\cite{smartrefresh}
+
+By employing a time-out counter for each memory row of a DRAM module, all the unnecessary periodic refresh operations can be eliminated.
+The basic concept behind our scheme is that a DRAM row that was recently read or written to by the processor (or other devices that share the same DRAM) does not need to be refreshed again by the periodic refresh operation, thereby eliminating excessive refreshes and the energy dissipated.
+
+refrint\cite{refrint}
+
 P. Emma et al.

### 2014-09-19 paper/include/retention.tex
added text and excluded pdf from git
cite added
explained techniques

@@
 \label{sec:ret}
-\cite{ACM:dtail}
+Papers that use this technique:
+
+DTail\cite{dtail}
+
+VRA
+
+RAIDR\cite{raidr}, from 2006...
+Software approach that can exploit off-the-shelf DRAMs to reduce refresh power to vanishingly
+small levels approaching non-volatile memory.
+The key idea is to favor longer-retention pages over shorter-retention pages when allocating DRAM pages.
+This allows selecting a single refresh period that depends on the shortest-retention page among populated pages,
+instead of the shortest-reten tion page overall.
+
+RIO\cite{rioparis}
+
+SECRET\cite{secret}
+Previous works adopt conventional ECC (Error Correcting Code) to correct retention errors.
+These approaches introduce significant area and energy overheads.
+In this paper, we propose a novel error correction framework for retention errors in DRAMs, called SECRET (Selective Error Correction for Refresh Energy reducTion).
+
+Energy Optimization in 3D MPSoCs with Wide-I/O DRAM Using Temperature Variation Aware Bank-wise Refresh
+
+We improve DRAM refresh power by considering the lateral and vertical temperature
+variations in the 3D structure and adapting the per-DRAM-bank refresh period accordingly.

### 2014-09-19 paper/include/tolerance.tex
added text and excluded pdf from git
explained techniques

--- 2014-09-19 paper/include/retention.tex
@@
-\section{Retention}
-\label{sec:ret}
-\cite{ACM:dtail}
+\section{Tolerance} 
+\label{sec:tol}
+
+Papers that use this technique:
+
+Flikker\cite{flikker}
+
+Sparkk\cite{sparkk}
+Multiple authors have suggested techniques where programmers can specify which data is critical and can not tolerate any bit errors and which data can be stored approximately.
+However, in these approaches all bits in the approximate area are treated as equally important.
+We show that this produces suboptimal results and higher energy savings or better quality can be achieved, if a more fine-grained approach is used.
+Our proposal is able to save more refresh power and enables a more effective storage of non-critical data by utilizing a non-uniform refresh of multiple DRAM chips and a permutation of the bits to the DRAM chips.

### 2014-09-21 paper/include/accessrecency.tex
started with A

@@
 
-smart\cite{smartrefresh}
+\subsection{smart\cite{smartrefresh} - from 2007}
+
+The basic concept behind our scheme is that a memory row that has been recently read out or written to does not need to be refreshed again by the periodic refresh mechanism.
+
+Uses a time-out counter for each memory row.
+The counter is set on a read/write/refresh and dicards refreshes until the timer hits zero.
+The counter is associated to a bank/row pair.
+The counters are stored in the memory controller.
+One counter consists of 2 bits.
+
+This scheme will be most effective if all rows are accessed before a refresh is needed.
+
+Implemented so only one row per each logical segments are deremented at the same time, which leads to that the refreshes are evenly distributed over time. (I.e. no burst refresh.)
+
+Employed its solution to 2 GB ram.
+What about todays sizes?
+How effective is the technique now?
+- Only works effective if large parts the memory is acccessed, i.e.
+large part of the memory has to be valid to begin with.
+Is it so today?
 
@@
 
-refrint\cite{refrint}
+\subsection{refrint\cite{refrint} - from 2013}
 

### 2014-09-21 paper/include/retention.tex
added techniques

--- 2014-09-19 paper/include/retention.tex
@@
 \label{sec:ret}
-\cite{ACM:dtail}
+
+Zehan Cui et al. \cite{dtail}
+
+RAIDR  by Jinghui Liu et al. \cite{raidr}
+
+RIO is a technique proposed by Seungjae Baek et al. \cite{rioparis}
+
+With SECRET Chung-Hsiang Lin et al.
+\cite{secret} masks bad cells using Error Correction Pointers (ECP).
+The bad cells are maped with a one time profiling done before first system use and all cells that dont meet the retention time requirements gets a ECP.
+A ECP is basiclly a pointer to another cell that is healthy.
+The framwork which is implemented inside the memory controller checks wheter there is any valid ECP for every address accessed.
+If there is, it replaces the bad cell(s) data with the data from ECP cell instead.
+
+This method gives a high energy saving of up to 87.2\% in refresh power and a reduction of 18.57\% in total DRAM power consumption.
+
+RAIDR
+Software approach that can exploit off-the-shelf DRAMs to reduce refresh power to vanishingly
+small levels approaching non-volatile memory.
+The key idea is to favor longer-retention pages over shorter-retention pages when allocating DRAM pages.
+This allows selecting a single refresh period that depends on the shortest-retention page among populated pages,
+instead of the shortest-reten tion page overall.

### 2014-09-21 paper/include/retention.tex
Written on retention. Basic outline

--- 2014-09-19 paper/include/retention.tex
@@
 \label{sec:ret}
-\cite{ACM:dtail}
+
+Zehan Cui et al.
+\cite{dtail} propses DTail-R in which data for row retention time is stored in the DRAM itself instead of storing it in the memory controller.
+To counter the added latency of a slower memory they utilizes the spatial locality of the retention data and does prefetching.
+Zehan Cui et al.
+also proposes DTail-V which can be combined with DTail-R and is further discussed in \ref{sec:val}.
+The retention time data consists of a 4 bit integer which is multiplied with 64ms to get the rows refresh period.
+They does not mention any compensation for temperature neither how they propose to aquire the retentio time data.
+DTail manages to decrease the amount of refreshes with 87.9\% and only have 16.7\% energy overhead comapred to a ideal 32GB DRAM (no refreshes).
+
+RAIDR by Jinghui Liu et al.
+\cite{raidr} divides all memory rows into different bins based on their retention time.
+They then refresh rows which have bad cells in them at normal interval while the other rows with healthier cells gets refreshed less frequent.
+The implementation is done inside the memory controller and the storage of the retention time in the bins is done using Bloom filters.
+To acomodate for temperature variations they scale all bins refresh times which is suboptimal.
+With RAIDR they achieve a decrease in power consumption of 16.1\% and a performance improvment of 8.6\% with only 1.25 KB storage overhead (in a 32 GB system).
+
+RIO is a technique proposed by Seungjae Baek et al.
+\cite{rioparis} in which they identify cells with low retention time (bad cells) and `deletes'.
+This is done by changing the memory allocating modules of the OS to not allocate pages that have bad cells in them.
+They implement RIO in a Linux kernel which is run and tested on a real hardware system and is able to achieve 87.5\% reduction in refresh count with a performance increas of 4.5\% on average.
+As the implementaion only relies on changes in the kernel and there is no need to modify the hardware this technique would be easy to adopt.
+A maximum of only 0.1\% of total memory pages gets `deleted' and it is temperature aware.
+Seungjae Baek et al.
+also propose PARIS discussed in \ref{sec:val} which can be combined with RIO for further improvments.
+
+With SECRET Chung-Hsiang Lin et al.
+\cite{secret} masks bad cells using Error Correction Pointers (ECP).
+The bad cells are maped with a one time profiling done before first system use and all cells that dont meet the retention time requirements gets a ECP.
+A ECP is basiclly a pointer to another cell that is healthy.
+The framwork which is implemented inside the memory controller checks whether there is any valid ECP for every address accessed.
+If there is, it replaces the bad cell(s) data with the data from ECP cell(s) instead.
+This method gives a high energy saving of up to 87.2\% in refresh power and a reduction of 18.57\% in total DRAM power consumption with the RCP cach overhead taken into account.
+The performance impact ranges from 1.3\% overhad att the worst test case to 1.4\% improvment at its best.
+The degredation of performance is due to the added memory access needed and the improvment is for the decreased congestion due to refreshes.

### 2014-09-21 paper/include/accessrecency.tex
wrote about A

@@
 
-\subsection{smart\cite{smartrefresh} - from 2007}
-
+\subsection{Smart Refresh\cite{smartrefresh} - from 2007}
 The basic concept behind our scheme is that a memory row that has been recently read out or written to does not need to be refreshed again by the periodic refresh mechanism.
@@
 The counters are stored in the memory controller.
-One counter consists of 2 bits.
+One counter consists of 2 SRAM bits.
+More bits give higher granularity and thus also better performance, the refresh is made less often.
 
 This scheme will be most effective if all rows are accessed before a refresh is needed.
+If no accesses are made = the entire working set fits into cache, Smart Refresh becomes CBR (CAS before RAS).
+Only refresh one row at a time = inefficient, so they have made it possible to turn the technique off.
 
 Implemented so only one row per each logical segments are deremented at the same time, which leads to that the refreshes are evenly distributed over time. (I.e. no burst refresh.)
+
+Overhead: 48KB for 2GB and 768 KB for 32GB = 0,0024\% overhead.
+
+Average reduction of refreshes by 59.3\% for 2GB and 52.3\% energy saving regarding the DRAM refresh. No performance degradation.
 
@@
 
-By employing a time-out counter for each memory row of a DRAM module, all the unnecessary periodic refresh operations can be eliminated.
-The basic concept behind our scheme is that a DRAM row that was recently read or written to by the processor (or other devices that share the same DRAM) does not need to be refreshed again by the periodic refresh operation, thereby eliminating excessive refreshes and the energy dissipated.
+\subsection{Refrint\cite{refrint} - from 2013}
+Tageted towards caches and eDRAM.
 
-\subsection{refrint\cite{refrint} - from 2013}
+Our goal is to refresh only the data that will be used in the near future, and only refresh it if it is really needed. The other data is invalidated and/ or written back to main memory.
 
-P. Emma et al.
+The technique target to minimize the refresh for the so called hot and cold lines.
+Hot lines are those being accessed and thus unnesssesary refreshed by the refresh mechanism, whereas the cold lines are those that hasn't been used in a while but still gets refreshed.
+
+To minimize the refresh for hot lines time-based policies are used.
+Normally a counter - Polyphase - is used, which practicly works in the same way as the time-out counters for Smart Refresh.
+It consists of two bits.
+They also has a basic periodic refresh scheme (why?).
+
+For the cold lines four different (data-based) policies can be applied: all, valid, dirty or WB(n,m).
+"All" refreshes all lines.
+Valid refreshes the valid lines.
+Dirty refreshes the dirty lines.The WB policy is associated with a tuple (n, m).
+WB refreshes a Dirty line that is not being accessed for n times before writing it back and changing its state to Valid Clean; moreover, it refreshes a Valid Clean line that is not being accessed for m times before invalidating it.
+WB retains a Dirty line in the cache Ion ger because evicting it has the additional cost of writing the line back to lower-Ievel memory.
+To implement WB, we maintain a per-Iine Count.
+W hen the line is read or written, Count is set to n (if Dirty) or m (if Valid Clean).
+In addition, when the line is refreshed, Count is decremented.
+W hen Count reaches zero, the line is either written back or invalidated.
+Note that the Dirty policy is equivalent to WB(oo,O), while Valid is equivalent to WB(oo,oo).
+Finally, every policy refreshes cache Iines in transient states as well.
+
+They have application categorization for the data-based policies (Cold ones). TODO
+
+Implementation: Hot lines.
+They use a global counter with M bits (M = 1 or 2).
+When a line is accessed, the global counter is copied to N bits associated to the line and those N bits are stored with the valid bit in the memory controller.
+If N = 2, the overhead becomes 0.6\%.
+When the global counter steps to a new phase, all normal accesses are stalled and the logic controlls the valid lines and its local phase bits to see if it matches the global phase.
+If there is a match, a refresh is scheduled.
+Otherwise the memory goes back to normal operation after the scan.
+
+Implementation: Cold lines. TODO
+
+\subsection{Differences and similarities}
+
+Refrint is somewhat a extension of Smart Refresh.
+Smart Refresh's features can be laid somewhat equal to Refrint's time-based polices, wheras the data-based polices have no eqvivalence in Smart Refresh.
+
+They both modify the memory controller, and almost in the same way.
+But Refrint has no timers that is getting decremented the whole time, and instead copies a global "timer" to the line associated bits upon access.
+Of cource Refrint's data-based polices also adds logic to the memory controller.
+
+Discuss energy savings.

### 2014-09-23 paper/include/retention.tex

@@
 Zehan Cui et al.
-\cite{dtail} propses DTail-R in which data for row retention time is stored in the DRAM itself instead of storing it in the memory controller.
+\cite{dtail} propses DTail-R in which data for row retention time is stored in the DRAM itself instead of storing it in the memory controller which many other solutions do.
 To counter the added latency of a slower memory they utilizes the spatial locality of the retention data and does prefetching.
@@
 also proposes DTail-V which can be combined with DTail-R and is further discussed in \ref{sec:val}.
-The retention time data consists of a 4 bit integer which is multiplied with 64ms to get the rows refresh period.
-They does not mention any compensation for temperature neither how they propose to aquire the retentio time data.
-DTail manages to decrease the amount of refreshes with 87.9\% and only have 16.7\% energy overhead comapred to a ideal 32GB DRAM (no refreshes).
+The retention time data consists of 4 bits and is interpreted in such way that the refresh period becomes \(64ms \times 2^n, n = [0...7]\).
+With this structure the amount of memory required is weighing in on a negliable amount of 0.006\% of the DRAM capacity (1.92MB in a 32GB DRAM).
+They does not however mention how compensation for temperature will work neither how they propose to aquire the retention time data for the rows.
+DTail manages to decrease the amount of refreshes with 87.9\% and only have 16.7\% energy overhead compared to a refresh less 32GB DRAM.
 

### 2014-09-23 paper/include/techniques.tex
wrote about A
worked on A

@@
-\section{Intelligent Refresh Techniques} 
-\label{sec:int}
-
+\section{Methods for decreasing the refresh rate} 
+\label{sec:met}
+There are several methods that can be applied to decrease the refresh rate and these can be categorized depending on with sort of information the scheduling decisions are based upon.
+The four sort of information are: access recency, cell retention times, error tolerance of the data, and line validity.
+This categorization follows the taxonamy proposed by authors \cite{dtail}.
 \subsection{Access Recency}

### 2014-09-23 paper/include/retention.tex
Finputs

--- 2014-09-21 paper/include/retention.tex
@@
 
-Zehan Cui et al. \cite{dtail}
+Zehan Cui et al.
+\cite{dtail} propses DTail-R in which data for line refresh rate is stored in the DRAM itself instead of storing it in the memory controller which many other solutions do \cite{raidr}\cite{smartrefresh}\cite{refrint}.
+To counter the added latency of a slower memory they utilizes spatial locality of the refresh data and performs prefetching into a FIFO buffer.
+The retention time data consists of 4 bits and is interpreted in such way that the refresh period becomes \(64ms \times 2^n, n = [0...7]\).
+With this structure the amount of memory required is weighing in on a negliable amount of 0.006\% DRAM capacity (i.e.
+1.92MB in a 32GB DRAM).
+One thing that DTail-R misses is refreshrate compensation for temperature variations.
+As the refresh rate is set per line DTail-R gets a higher resolution than techniques that set refreshrate on a higher level.
+Therefore DTail-R manages to decrease the amount of refreshes with as much as 87.9\% and thereby only have 16.7\% energy overhead compared to a refresh less 32GB DRAM.
+Zehan Cui et al.
+also proposes DTail-V \cite{dtail} which can be combined with DTail-R and is further discussed in \ref{sec:val}.
 
-RAIDR  by Jinghui Liu et al. \cite{raidr}
+RAIDR by Jinghui Liu et al.
+\cite{raidr} divides all memory lines into different groups, called bins, based on the cell with the shortest retention time in the line.
+In the basic configuration there are two bins, one with a refreshrate of \(64ms - 128ms\) and one with \(128ms - 256ms\).
+Lines in the first bin gets refreshed every \(64ms\), lines in the second bin every \(128ms\), and lines not in any bin gets refreshed at the default refresh rate which is set to \(256ms\).
+The implementation is done inside the memory controller and to minimize the storage needed for keeping track of bin members they use Bloom filters.
+Thanks to this only 1.25 KB storage is needed for a 32 GB DRAM with the default configuration.
+The storage requirement can however vary depending the amount of bins used.
+RAIDR also acomodate for temperature variations by scaling all bins refresh rates using a timer.
+With RAIDR a performance improvment of 4.1\% and a decreas in powerconsumption by 8.3\% is achieved in a system with 32GB DRAM compared to auto-refresh.
 
-RIO is a technique proposed by Seungjae Baek et al. \cite{rioparis}
+RIO is a technique proposed by Seungjae Baek et al.
+\cite{rioparis} in which they identify cells with low retention time (bad cells) and `deletes' them.
+This is done by changing the memory allocating modules in the OS to not allocate pages that have bad cells in them.
+They implement RIO in a Linux kernel which is run and tested on a real hardware system and is able to achieve 87.5\% reduction in refresh count with a performance increas of 4.5\% on average.
+As the implementaion only relies on changes in the kernel and there is no need to modify the hardware this technique would be easy to adopt in systems today.
+To keep fragmentation of the memory low a maximum of 0.1\% of total memory pages gets `deleted'.
+Even though deletion of a larger amount of pages could decrease the amount of refreshes further it would cause trouble for the kernel which relies on being able to allocate large contiguous address spaces.
+BLASDLBASDAHSUDGASJHDGJHASGDJHASGDJHAGSDJHGASJHDGAJSHDand it is temperature aware.
+Seungjae Baek et al.
+also propose PARIS discussed in \ref{sec:val} which can be combined with RIO for further improvments.
 
@@
 A ECP is basiclly a pointer to another cell that is healthy.
-The framwork which is implemented inside the memory controller checks wheter there is any valid ECP for every address accessed.
-If there is, it replaces the bad cell(s) data with the data from ECP cell instead.
-
-This method gives a high energy saving of up to 87.2\% in refresh power and a reduction of 18.57\% in total DRAM power consumption.
-
-RAIDR
-Software approach that can exploit off-the-shelf DRAMs to reduce refresh power to vanishingly
-small levels approaching non-volatile memory.
-The key idea is to favor longer-retention pages over shorter-retention pages when allocating DRAM pages.
-This allows selecting a single refresh period that depends on the shortest-retention page among populated pages,
-instead of the shortest-reten tion page overall.
+The framwork which is implemented inside the memory controller checks whether there is any valid ECP for every address accessed.
+If there is, it replaces the bad cell(s) data with the data from ECP cell(s) instead.
+This method gives a high energy saving of up to 87.2\% in refresh power and a reduction of 18.57\% in total DRAM power consumption with the RCP cach overhead taken into account.
+The performance impact ranges from 1.3\% overhad att the worst test case to 1.4\% improvment at its best.
+The degredation of performance is due to the added memory access needed and the improvment is for the decreased congestion due to refreshes.

### 2014-09-23 paper/include/accessrecency.tex

@@
 \subsection{Smart Refresh\cite{smartrefresh} - from 2007}
-The basic concept behind our scheme is that a memory row that has been recently read out or written to does not need to be refreshed again by the periodic refresh mechanism.
+The basic concept behind our scheme is that a memory line that has been recently read out or written to does not need to be refreshed again by the periodic refresh mechanism.
 
-Uses a time-out counter for each memory row.
+Uses a time-out counter for each memory line.
 The counter is set on a read/write/refresh and dicards refreshes until the timer hits zero.
-The counter is associated to a bank/row pair.
+The counter is associated to a bank/line pair.
 The counters are stored in the memory controller.
 One counter consists of 2 SRAM bits.
-More bits give higher granularity and thus also better performance, the refresh is made less often.
+More bits give higher granularity and thus also better performance because the refresh is made less often.
 
-This scheme will be most effective if all rows are accessed before a refresh is needed.
+This scheme will be most effective if all lines are accessed before a refresh is needed.
 If no accesses are made = the entire working set fits into cache, Smart Refresh becomes CBR (CAS before RAS).
-Only refresh one row at a time = inefficient, so they have made it possible to turn the technique off.
+Only refresh one line at a time = inefficient, so they have made it possible to turn the technique off.
 
-Implemented so only one row per each logical segments are deremented at the same time, which leads to that the refreshes are evenly distributed over time. (I.e. no burst refresh.)
+Implemented so only one line per each logical segments are deremented at the same time, which leads to that the refreshes are evenly distributed over time. (I.e. no burst refresh.)
 
@@
 
-Our goal is to refresh only the data that will be used in the near future, and only refresh it if it is really needed. The other data is invalidated and/ or written back to main memory.
+Our goal is to refresh only the data that will be used in the near future, and only refresh if it is really needed. The other data is invalidated and/ or written back to main memory.
 
@@
 
-For the cold lines four different (data-based) policies can be applied: all, valid, dirty or WB(n,m).
-"All" refreshes all lines.
-Valid refreshes the valid lines.
-Dirty refreshes the dirty lines.The WB policy is associated with a tuple (n, m).
+For the cold lines four different (data-based) policies can be applied: \textit{all}, \textit{valid}, \textit{dirty} or WB(n,m).
+\textit{All} refreshes all lines.
+\textit{Valid} refreshes the valid lines.
+\textit{Dirty} refreshes the dirty lines.
+The WB policy is associated with a tuple (n, m).
 WB refreshes a Dirty line that is not being accessed for n times before writing it back and changing its state to Valid Clean; moreover, it refreshes a Valid Clean line that is not being accessed for m times before invalidating it.
@@
 Note that the Dirty policy is equivalent to WB(oo,O), while Valid is equivalent to WB(oo,oo).
-Finally, every policy refreshes cache Iines in transient states as well.
+Finally, every policy refreshes cache lines in transient states as well.
 
-They have application categorization for the data-based policies (Cold ones). TODO
+They categorize applications based on how they are affected by the data-based policies.
+Fig.
+\ref{fig:app_cat} presents an application categorization based on two axes: application footprint and visibility.
+It is likely........
+TODO
+
+\begin{figure}[t!]
+	\includegraphics[width=\textwidth/2]{app_cat}
+	\caption{Application categorization according to the data-based policy \cite{refrint}.}
+	\label{fig:app_cat}
+\end{figure}
 
@@
 
-Implementation: Cold lines. TODO
+Implementation: Cold lines. For each line there is a 5 bit counter and the line's state bits. Depending on the data policy  for each line, but they are somehow neglectible.  ...
+
+Remember that this technique is targeted towards caches and eDRAM and only is tested in that enviroment.
+This means that we can not really use the results shown in their paper, but the techniques can be applied to DRAM.
 
@@
 
-Discuss energy savings.
+Discuss energy savings. Hard to do eDRAM vs DRAM...

### 2014-09-29 lab/pcod_benchmarks/microbench1.c
fu

@@ -1,0 +1,81 @@
+#include <pthread.h>
+#include <stdlib.h>
+#include <stdio.h>
+
+#define min(a,b) (((a)<(b))?(a):(b))
+#define CSIZE 16
+
+static pthread_mutex_t lock;
+static pthread_barrier_t barrier;
+
+int* counterArray;
+
+int nthreads;
+int ncounters;
+
+void* chain(void* arglist)
+{
+    int i, j, summation;
+    long tid = (long) arglist;
+
+    for (i = 0; i < nthreads; i++) { 
+        pthread_mutex_lock(&lock);
+
+        if (tid == nthreads-1) {
+            for (j = 0; j < ncounters; j++) 
+                counterArray[j*CSIZE]++;
+        } else {
+            for (j = 0; j < ncounters; j++) 
+                summation += counterArray[j*CSIZE];
+        }
+       
+        pthread_mutex_unlock(&lock);
+        pthread_barrier_wait(&barrier);
+    }
+
+    return NULL;
+}
+
+
+int main(int argc, const char** const argv)
+{
+    if (argc != 3) {
+       printf("Usage: ./microbench1 <nthreads> <ncounters>\n");
+       exit(1);
+    }
+    
+    nthreads = atoi(argv[1]);
+    ncounters = atoi(argv[2]);
+
+    if (nthreads < 2 || ncounters < 1) {
+        printf("This test requires at least 2 CPUs, 1 counter variable\n");
+        exit(1);
+    }
+    
+    long i;
+    pthread_t pth[nthreads];
+    pthread_attr_t attr;
+    
+    pthread_attr_init(&attr);
+    pthread_attr_setscope(&attr, PTHREAD_SCOPE_SYSTEM);
+    
+    pthread_mutex_init(&lock, NULL);
+    printf("Init done\n");
+    
+    
+    pthread_barrier_init(&barrier, NULL, nthreads);
+    
+    counterArray = (int*) calloc(ncounters*CSIZE, sizeof(int));
+    
+    for (i = 0; i < nthreads; i++) {
+        pthread_create(&pth[i], &attr, chain, (void*) i);
+    }
+    
+    for (i = 0; i < nthreads; i++) { 
+        pthread_join(pth[i], NULL);
+    }
+    
+    printf("Val: %d\n", counterArray[0]);
+    printf("PASSED :-)\n");
+    return 0;
+}

### 2014-09-29 lab/pcod_benchmarks/microbench2.c

@@ -1,0 +1,75 @@
+#include <pthread.h>
+#include <stdlib.h>
+#include <stdio.h>
+
+#define min(a,b) (((a)<(b))?(a):(b))
+#define CSIZE 16
+
+static pthread_mutex_t lock;
+static pthread_barrier_t barrier;
+
+int* counterArray;
+
+int nthreads;
+int ncounters;
+
+void* chain(void* arglist)
+{
+    int i, j;
+    long tid = (long) arglist;
+ 
+    for (i = 0; i < nthreads; i++) { 
+
+        for (j = 0; j < ncounters; j++)
+	    if(j%nthreads == tid)
+                counterArray[j*CSIZE]++;
+       
+        pthread_barrier_wait(&barrier);
+    }
+        
+    return NULL;
+}
+
+
+int main(int argc, const char** const argv)
+{
+    if (argc != 3) {
+       printf("Usage: ./microbench2 <nthreads> <ncounters>\n");
+       exit(1);
+    }
+    
+    nthreads = atoi(argv[1]);
+    ncounters = atoi(argv[2]);
+
+    if (nthreads < 2 || ncounters < 1) {
+        printf("This test requires at least 2 CPUs, 1 counter variable\n");
+        exit(1);
+    }
+    
+    long i;
+    pthread_t pth[nthreads];
+    pthread_attr_t attr;
+    
+    pthread_attr_init(&attr);
+    pthread_attr_setscope(&attr, PTHREAD_SCOPE_SYSTEM);
+    
+    pthread_mutex_init(&lock, NULL);
+    printf("Init done\n");
+    
+    
+    pthread_barrier_init(&barrier, NULL, nthreads);
+    
+    counterArray = (int*) calloc(ncounters*CSIZE, sizeof(int));
+    
+    for (i = 0; i < nthreads; i++) {
+        pthread_create(&pth[i], &attr, chain, (void*)i);
+    }
+    
+    for (i = 0; i < nthreads; i++) { 
+        pthread_join(pth[i], NULL);
+    }
+    
+    printf("Val: %d\n", counterArray[0]);
+    printf("PASSED :-)\n");
+    return 0;
+}

### 2014-09-29 lab/pcod_benchmarks/microbench3.c

@@ -1,0 +1,75 @@
+#include <pthread.h>
+#include <stdlib.h>
+#include <stdio.h>
+
+#define min(a,b) (((a)<(b))?(a):(b))
+#define CSIZE 16
+
+static pthread_mutex_t lock;
+static pthread_barrier_t barrier;
+
+int* counterArray;
+
+int nthreads;
+int ncounters;
+
+void* chain(void* arglist)
+{
+    int i, j;
+
+    for (i = 0; i < nthreads; i++) { 
+        pthread_mutex_lock(&lock);
+
+        for (j = 0; j < ncounters; j++) 
+            counterArray[j*CSIZE]++;
+       
+        pthread_mutex_unlock(&lock);
+        pthread_barrier_wait(&barrier);
+    }
+        
+    return NULL;
+}
+
+
+int main(int argc, const char** const argv)
+{
+    if (argc != 3) {
+       printf("Usage: ./microbench2 <nthreads> <ncounters>\n");
+       exit(1);
+    }
+    
+    nthreads = atoi(argv[1]);
+    ncounters = atoi(argv[2]);
+
+    if (nthreads < 2 || ncounters < 1) {
+        printf("This test requires at least 2 CPUs, 1 counter variable\n");
+        exit(1);
+    }
+    
+    int i;
+    pthread_t pth[nthreads];
+    pthread_attr_t attr;
+    
+    pthread_attr_init(&attr);
+    pthread_attr_setscope(&attr, PTHREAD_SCOPE_SYSTEM);
+    
+    pthread_mutex_init(&lock, NULL);
+    printf("Init done\n");
+    
+    
+    pthread_barrier_init(&barrier, NULL, nthreads);
+    
+    counterArray = (int*) calloc(ncounters*CSIZE, sizeof(int));
+    
+    for (i = 0; i < nthreads; i++) {
+        pthread_create(&pth[i], &attr, chain, NULL);
+    }
+    
+    for (i = 0; i < nthreads; i++) { 
+        pthread_join(pth[i], NULL);
+    }
+    
+    printf("Val: %d\n", counterArray[0]);
+    printf("PASSED :-)\n");
+    return 0;
+}

### 2014-09-30 paper/include/background.tex
changed paper structure

\section{Background} 
\label{sec:bg}

Explain the DRAM structure. Use picture.

Explain basic operations: load and store.

Explain different refresh modes: Burst refresh vs distributed refresh and RAS-only refresh vs CAS before RAS refresh.

Trends - Storage array rectangular
            cant go wider (word lines length = timing = power)
            gets taller = more refreshes
            over time .... projektions (see Intel)

### 2014-09-30 paper/include/discussion.tex
changed paper structure

\section{Discussion} 
\label{sec:disc}

Talk about scheduling real accesses togheter with refresh ops

How do we cope with 3D DRAM?

### 2014-09-30 paper/include/introduction.tex
started with A
wrote about A
changed paper structure

@@
 \label{sec:int}
-Why is it important to reduce the refreshes?
-Briefly mention the methods.
+Whats the problem?
+energy is an inc. system design constraint
+
+Why is it important?
+Main mem consumes a groving portion of system power and this will continue to increase (details \& refs in background)
+
+Mention 3D stacked DRAM: higher tempratures, which gives a higher refresh rate that also increases temratures.
+
+Most main memories include dramtwo ssources of powerconsumption in DRAM static/dynamic (blahah/driving data)
+static leakage
+
+What is the shape of the solution?
+DRAM refresh ops account for a groing portion of dynamic power, yet they perform no usefiul accesses (from cpu perspective)
+
+Why will it work? (dont go into detail, this is a survey paper)
+Finding ways to reduce power spent on frefresh is a straightfroward way to mitigate DRAM power
+
+Briefly mention the methods to lower the refresh rate.

### 2014-09-30 paper/include/reducing_refresh_power.tex
changed paper structure

--- 2014-09-23 paper/include/techniques.tex
@@
-\section{Methods for decreasing the refresh rate} 
-\label{sec:met}
+\section{Reducing refresh power} 
+\label{sec:red}
+Roadmap 
+Worth mentioning scheduling real acces togheter with refresh ops
+Write an intro to the ARTV approches.
 There are several methods that can be applied to decrease the refresh rate and these can be categorized depending on with sort of information the scheduling decisions are based upon.
@@
 This categorization follows the taxonamy proposed by authors \cite{dtail}.
-\subsection{Access Recency}
+
+\subsection{Types of refreshing}
+
+\subsubsection*{\textbf{Access Recency}}
 DRAM accesses imply refresh operations, and so a subsequent refresh to the same row can be postponed.
@@
 
-\subsection{Retention}
+\subsubsection*{\textbf{Retention Time}}
 Retention time refers to the period during which cells hold valid values as charge gradually leaks.
@@
 
-\subsection{Tolerance}
+\subsubsection*{\textbf{Data Tolerance}}
 Applications like games, media processing, machine learning, and unstructured information analysis tolerate errors in portions of their data and still produce acceptably accurate results.
@@
 
-\subsection{Validity}
+\subsubsection*{\textbf{Validity}}
 If the OS has not allocated the page frame, its data are meaningless, and refreshes to it are wasteful. Severa
@@
 The effectiveness of schemes using such V refresh information is sensitive to the total memory usage of the system.
+
+\subsection{Reducing refresh power approches}
+
+\subsection{Access Recency}

### 2014-09-30 paper/include/dtail.tex
Splited retention into parts

--- 2014-09-23 paper/include/retention.tex
@@
-\section{Retention}
-\label{sec:ret}
-
 Zehan Cui et al.
@@
 also proposes DTail-V \cite{dtail} which can be combined with DTail-R and is further discussed in \ref{sec:val}.
-
-RAIDR by Jinghui Liu et al.
-\cite{raidr} divides all memory lines into different groups, called bins, based on the cell with the shortest retention time in the line.
-In the basic configuration there are two bins, one with a refreshrate of \(64ms - 128ms\) and one with \(128ms - 256ms\).
-Lines in the first bin gets refreshed every \(64ms\), lines in the second bin every \(128ms\), and lines not in any bin gets refreshed at the default refresh rate which is set to \(256ms\).
-The implementation is done inside the memory controller and to minimize the storage needed for keeping track of bin members they use Bloom filters.
-Thanks to this only 1.25 KB storage is needed for a 32 GB DRAM with the default configuration.
-The storage requirement can however vary depending the amount of bins used.
-RAIDR also acomodate for temperature variations by scaling all bins refresh rates using a timer.
-With RAIDR a performance improvment of 4.1\% and a decreas in powerconsumption by 8.3\% is achieved in a system with 32GB DRAM compared to auto-refresh.
-
-RIO is a technique proposed by Seungjae Baek et al.
-\cite{rioparis} in which they identify cells with low retention time (bad cells) and `deletes' them.
-This is done by changing the memory allocating modules in the OS to not allocate pages that have bad cells in them.
-They implement RIO in a Linux kernel which is run and tested on a real hardware system and is able to achieve 87.5\% reduction in refresh count with a performance increas of 4.5\% on average.
-As the implementaion only relies on changes in the kernel and there is no need to modify the hardware this technique would be easy to adopt in systems today.
-To keep fragmentation of the memory low a maximum of 0.1\% of total memory pages gets `deleted'.
-Even though deletion of a larger amount of pages could decrease the amount of refreshes further it would cause trouble for the kernel which relies on being able to allocate large contiguous address spaces.
-BLASDLBASDAHSUDGASJHDGJHASGDJHASGDJHAGSDJHGASJHDGAJSHDand it is temperature aware.
-Seungjae Baek et al.
-also propose PARIS discussed in \ref{sec:val} which can be combined with RIO for further improvments.
-
-With SECRET Chung-Hsiang Lin et al.
-\cite{secret} masks bad cells using Error Correction Pointers (ECP).
-The bad cells are maped with a one time profiling done before first system use and all cells that dont meet the retention time requirements gets a ECP.
-A ECP is basiclly a pointer to another cell that is healthy.
-The framwork which is implemented inside the memory controller checks whether there is any valid ECP for every address accessed.
-If there is, it replaces the bad cell(s) data with the data from ECP cell(s) instead.
-This method gives a high energy saving of up to 87.2\% in refresh power and a reduction of 18.57\% in total DRAM power consumption with the RCP cach overhead taken into account.
-The performance impact ranges from 1.3\% overhad att the worst test case to 1.4\% improvment at its best.
-The degredation of performance is due to the added memory access needed and the improvment is for the decreased congestion due to refreshes.

### 2014-09-30 paper/include/raidr.tex
Splited retention into parts

--- 2014-09-23 paper/include/retention.tex
@@
-\section{Retention}
-\label{sec:ret}
-
-Zehan Cui et al.
-\cite{dtail} propses DTail-R in which data for line refresh rate is stored in the DRAM itself instead of storing it in the memory controller which many other solutions do \cite{raidr}\cite{smartrefresh}\cite{refrint}.
-To counter the added latency of a slower memory they utilizes spatial locality of the refresh data and performs prefetching into a FIFO buffer.
-The retention time data consists of 4 bits and is interpreted in such way that the refresh period becomes \(64ms \times 2^n, n = [0...7]\).
-With this structure the amount of memory required is weighing in on a negliable amount of 0.006\% DRAM capacity (i.e.
-1.92MB in a 32GB DRAM).
-One thing that DTail-R misses is refreshrate compensation for temperature variations.
-As the refresh rate is set per line DTail-R gets a higher resolution than techniques that set refreshrate on a higher level.
-Therefore DTail-R manages to decrease the amount of refreshes with as much as 87.9\% and thereby only have 16.7\% energy overhead compared to a refresh less 32GB DRAM.
-Zehan Cui et al.
-also proposes DTail-V \cite{dtail} which can be combined with DTail-R and is further discussed in \ref{sec:val}.
-
 RAIDR by Jinghui Liu et al.
@@
 With RAIDR a performance improvment of 4.1\% and a decreas in powerconsumption by 8.3\% is achieved in a system with 32GB DRAM compared to auto-refresh.
-
-RIO is a technique proposed by Seungjae Baek et al.
-\cite{rioparis} in which they identify cells with low retention time (bad cells) and `deletes' them.
-This is done by changing the memory allocating modules in the OS to not allocate pages that have bad cells in them.
-They implement RIO in a Linux kernel which is run and tested on a real hardware system and is able to achieve 87.5\% reduction in refresh count with a performance increas of 4.5\% on average.
-As the implementaion only relies on changes in the kernel and there is no need to modify the hardware this technique would be easy to adopt in systems today.
-To keep fragmentation of the memory low a maximum of 0.1\% of total memory pages gets `deleted'.
-Even though deletion of a larger amount of pages could decrease the amount of refreshes further it would cause trouble for the kernel which relies on being able to allocate large contiguous address spaces.
-BLASDLBASDAHSUDGASJHDGJHASGDJHASGDJHAGSDJHGASJHDGAJSHDand it is temperature aware.
-Seungjae Baek et al.
-also propose PARIS discussed in \ref{sec:val} which can be combined with RIO for further improvments.
-
-With SECRET Chung-Hsiang Lin et al.
-\cite{secret} masks bad cells using Error Correction Pointers (ECP).
-The bad cells are maped with a one time profiling done before first system use and all cells that dont meet the retention time requirements gets a ECP.
-A ECP is basiclly a pointer to another cell that is healthy.
-The framwork which is implemented inside the memory controller checks whether there is any valid ECP for every address accessed.
-If there is, it replaces the bad cell(s) data with the data from ECP cell(s) instead.
-This method gives a high energy saving of up to 87.2\% in refresh power and a reduction of 18.57\% in total DRAM power consumption with the RCP cach overhead taken into account.
-The performance impact ranges from 1.3\% overhad att the worst test case to 1.4\% improvment at its best.
-The degredation of performance is due to the added memory access needed and the improvment is for the decreased congestion due to refreshes.

### 2014-09-30 paper/include/rioparis.tex
Splited retention into parts

--- 2014-09-23 paper/include/retention.tex
@@
-\section{Retention}
-\label{sec:ret}
-
-Zehan Cui et al.
-\cite{dtail} propses DTail-R in which data for line refresh rate is stored in the DRAM itself instead of storing it in the memory controller which many other solutions do \cite{raidr}\cite{smartrefresh}\cite{refrint}.
-To counter the added latency of a slower memory they utilizes spatial locality of the refresh data and performs prefetching into a FIFO buffer.
-The retention time data consists of 4 bits and is interpreted in such way that the refresh period becomes \(64ms \times 2^n, n = [0...7]\).
-With this structure the amount of memory required is weighing in on a negliable amount of 0.006\% DRAM capacity (i.e.
-1.92MB in a 32GB DRAM).
-One thing that DTail-R misses is refreshrate compensation for temperature variations.
-As the refresh rate is set per line DTail-R gets a higher resolution than techniques that set refreshrate on a higher level.
-Therefore DTail-R manages to decrease the amount of refreshes with as much as 87.9\% and thereby only have 16.7\% energy overhead compared to a refresh less 32GB DRAM.
-Zehan Cui et al.
-also proposes DTail-V \cite{dtail} which can be combined with DTail-R and is further discussed in \ref{sec:val}.
-
-RAIDR by Jinghui Liu et al.
-\cite{raidr} divides all memory lines into different groups, called bins, based on the cell with the shortest retention time in the line.
-In the basic configuration there are two bins, one with a refreshrate of \(64ms - 128ms\) and one with \(128ms - 256ms\).
-Lines in the first bin gets refreshed every \(64ms\), lines in the second bin every \(128ms\), and lines not in any bin gets refreshed at the default refresh rate which is set to \(256ms\).
-The implementation is done inside the memory controller and to minimize the storage needed for keeping track of bin members they use Bloom filters.
-Thanks to this only 1.25 KB storage is needed for a 32 GB DRAM with the default configuration.
-The storage requirement can however vary depending the amount of bins used.
-RAIDR also acomodate for temperature variations by scaling all bins refresh rates using a timer.
-With RAIDR a performance improvment of 4.1\% and a decreas in powerconsumption by 8.3\% is achieved in a system with 32GB DRAM compared to auto-refresh.
-
 RIO is a technique proposed by Seungjae Baek et al.
@@
 also propose PARIS discussed in \ref{sec:val} which can be combined with RIO for further improvments.
-
-With SECRET Chung-Hsiang Lin et al.
-\cite{secret} masks bad cells using Error Correction Pointers (ECP).
-The bad cells are maped with a one time profiling done before first system use and all cells that dont meet the retention time requirements gets a ECP.
-A ECP is basiclly a pointer to another cell that is healthy.
-The framwork which is implemented inside the memory controller checks whether there is any valid ECP for every address accessed.
-If there is, it replaces the bad cell(s) data with the data from ECP cell(s) instead.
-This method gives a high energy saving of up to 87.2\% in refresh power and a reduction of 18.57\% in total DRAM power consumption with the RCP cach overhead taken into account.
-The performance impact ranges from 1.3\% overhad att the worst test case to 1.4\% improvment at its best.
-The degredation of performance is due to the added memory access needed and the improvment is for the decreased congestion due to refreshes.

### 2014-09-30 paper/include/secret.tex
Splited retention into parts

--- 2014-09-21 paper/include/retention.tex
@@
-\section{Retention}
-\label{sec:ret}
-
-Zehan Cui et al.
-\cite{dtail} propses DTail-R in which data for row retention time is stored in the DRAM itself instead of storing it in the memory controller.
-To counter the added latency of a slower memory they utilizes the spatial locality of the retention data and does prefetching.
-Zehan Cui et al.
-also proposes DTail-V which can be combined with DTail-R and is further discussed in \ref{sec:val}.
-The retention time data consists of a 4 bit integer which is multiplied with 64ms to get the rows refresh period.
-They does not mention any compensation for temperature neither how they propose to aquire the retentio time data.
-DTail manages to decrease the amount of refreshes with 87.9\% and only have 16.7\% energy overhead comapred to a ideal 32GB DRAM (no refreshes).
-
-RAIDR by Jinghui Liu et al.
-\cite{raidr} divides all memory rows into different bins based on their retention time.
-They then refresh rows which have bad cells in them at normal interval while the other rows with healthier cells gets refreshed less frequent.
-The implementation is done inside the memory controller and the storage of the retention time in the bins is done using Bloom filters.
-To acomodate for temperature variations they scale all bins refresh times which is suboptimal.
-With RAIDR they achieve a decrease in power consumption of 16.1\% and a performance improvment of 8.6\% with only 1.25 KB storage overhead (in a 32 GB system).
-
-RIO is a technique proposed by Seungjae Baek et al.
-\cite{rioparis} in which they identify cells with low retention time (bad cells) and `deletes'.
-This is done by changing the memory allocating modules of the OS to not allocate pages that have bad cells in them.
-They implement RIO in a Linux kernel which is run and tested on a real hardware system and is able to achieve 87.5\% reduction in refresh count with a performance increas of 4.5\% on average.
-As the implementaion only relies on changes in the kernel and there is no need to modify the hardware this technique would be easy to adopt.
-A maximum of only 0.1\% of total memory pages gets `deleted' and it is temperature aware.
-Seungjae Baek et al.
-also propose PARIS discussed in \ref{sec:val} which can be combined with RIO for further improvments.
-
 With SECRET Chung-Hsiang Lin et al.

### 2014-09-30 paper/include/refrint.tex
cleaned

--- 2014-09-23 paper/include/accessrecency.tex
@@
-\section{Access Recency}
-\label{sec:acc}
-
-Papers that use this technique:
-
-\subsection{Smart Refresh\cite{smartrefresh} - from 2007}
-The basic concept behind our scheme is that a memory line that has been recently read out or written to does not need to be refreshed again by the periodic refresh mechanism.
-
-Uses a time-out counter for each memory line.
-The counter is set on a read/write/refresh and dicards refreshes until the timer hits zero.
-The counter is associated to a bank/line pair.
-The counters are stored in the memory controller.
-One counter consists of 2 SRAM bits.
-More bits give higher granularity and thus also better performance because the refresh is made less often.
-
-This scheme will be most effective if all lines are accessed before a refresh is needed.
-If no accesses are made = the entire working set fits into cache, Smart Refresh becomes CBR (CAS before RAS).
-Only refresh one line at a time = inefficient, so they have made it possible to turn the technique off.
-
-Implemented so only one line per each logical segments are deremented at the same time, which leads to that the refreshes are evenly distributed over time. (I.e. no burst refresh.)
-
-Overhead: 48KB for 2GB and 768 KB for 32GB = 0,0024\% overhead.
-
-Average reduction of refreshes by 59.3\% for 2GB and 52.3\% energy saving regarding the DRAM refresh. No performance degradation.
-
-Employed its solution to 2 GB ram.
-What about todays sizes?
-How effective is the technique now?
-- Only works effective if large parts the memory is acccessed, i.e.
-large part of the memory has to be valid to begin with.
-Is it so today?
-
-\subsection{Refrint\cite{refrint} - from 2013}
 Tageted towards caches and eDRAM.

### 2014-09-30 paper/include/smart_refresh.tex
cleaned

--- 2014-09-21 paper/include/accessrecency.tex
@@
-\section{Access Recency}
-\label{sec:acc}
+The basic concept behind our scheme is that a memory line that has been recently read out or written to does not need to be refreshed again by the periodic refresh mechanism.
 
-Papers that use this technique:
+Uses a time-out counter for each memory line.
+The counter is set on a read/write/refresh and dicards refreshes until the timer hits zero.
+The counter is associated to a bank/line pair.
+The counters are stored in the memory controller.
+One counter consists of 2 SRAM bits.
+More bits give higher granularity and thus also better performance because the refresh is made less often.
 
-\subsection{smart\cite{smartrefresh} - from 2007}
+This scheme will be most effective if all lines are accessed before a refresh is needed.
+If no accesses are made = the entire working set fits into cache, Smart Refresh becomes CBR (CAS before RAS).
+Only refresh one line at a time = inefficient, so they have made it possible to turn the technique off.
 
-The basic concept behind our scheme is that a memory row that has been recently read out or written to does not need to be refreshed again by the periodic refresh mechanism.
+Implemented so only one line per each logical segments are deremented at the same time, which leads to that the refreshes are evenly distributed over time. (I.e. no burst refresh.)
 
-Uses a time-out counter for each memory row.
-The counter is set on a read/write/refresh and dicards refreshes until the timer hits zero.
-The counter is associated to a bank/row pair.
-The counters are stored in the memory controller.
-One counter consists of 2 bits.
+Overhead: 48KB for 2GB and 768 KB for 32GB = 0,0024\% overhead.
 
-This scheme will be most effective if all rows are accessed before a refresh is needed.
-
-Implemented so only one row per each logical segments are deremented at the same time, which leads to that the refreshes are evenly distributed over time. (I.e. no burst refresh.)
+Average reduction of refreshes by 59.3\% for 2GB and 52.3\% energy saving regarding the DRAM refresh. No performance degradation.
 
@@
 Is it so today?
-
-By employing a time-out counter for each memory row of a DRAM module, all the unnecessary periodic refresh operations can be eliminated.
-The basic concept behind our scheme is that a DRAM row that was recently read or written to by the processor (or other devices that share the same DRAM) does not need to be refreshed again by the periodic refresh operation, thereby eliminating excessive refreshes and the energy dissipated.
-
-\subsection{refrint\cite{refrint} - from 2013}
-
-P. Emma et al.

### 2014-09-30 paper/include/introduction.tex
more reorganization

--- 2014-09-19 paper/include/introduction.tex
@@
 \label{sec:int}
-Why is it important to reduce the refreshes?
-Briefly mention the methods.
+\textbf{Whats the problem?}
+energy is an inc. system design constraint
+
+\textbf{Why is it important?}
+Main mem consumes a groving portion of system power and this will continue to increase (details \& refs in background)
+
+Mention 3D stacked DRAM: higher tempratures, which gives a higher refresh rate that also increases temratures.
+
+Most main memories include dramtwo ssources of powerconsumption in DRAM static/dynamic (blahah/driving data)
+static leakage
+
+\textbf{What is the shape of the solution?}
+DRAM refresh ops account for a groing portion of dynamic power, yet they perform no usefiul accesses (from cpu perspective)
+
+\textbf{Why will it work? (dont go into detail, this is a survey paper)}
+Finding ways to reduce power spent on frefresh is a straightfroward way to mitigate DRAM power
+
+Briefly mention the methods to lower the refresh rate.

### 2014-09-30 paper/include/introduction.tex
written everything!

--- 2014-09-19 paper/include/introduction.tex
@@
 \label{sec:int}
-Why is it important to reduce the refreshes?
-Briefly mention the methods.
+%\textbf{Whats the problem?}
+Technology is ever advancing and computer systems of today are more powerfull and energy effecient than ever before.
+These improvments have to a large part the continuos decrease in integrated circuit feature size to thank but as we now are getting close to the physiological limit on how small transistors can become possible improvments have to be looked for elsewhere.
+The fact that battery powered mobile systems are getting more and more common and that power consupmtion correspons to a large part of the cost in driving the data centers of today makes energy consumption an increasingly big system design constraint.
+%\textbf{Why is it important?}
+
+DRAM is the most commonly used memory technique as main memory in computer systems today and represents a groving portion of system power consumption.
+This portion will continue to increase if nothing is done about it and a large part of the DRAM power consumption comes from refreshing all memory-cells in the DRAM.
+This refreshing has to be done as DRAM store data as a charge in a capacitor and given todays silicon, this charge leak over time.
+To increase DRAM performance and lower access latencies 3D stacked DRAM has been proposed in which the working temperature will be higher than in regular DRAM.
+This poses a problem as the rate at which the charge leaks increases with temperature and the DRAM will have to be refreshed more often.
+
+Power consumption in DRAM consists of two parts, static power and dynamic power.
+The static power is caused by leakage and the dynamic power is the result of all accesse to the memory.
+As a refresh basically is the same as a regular read access execpt that the data is not used, and thereby consumes the same amount of energy as a usefull access.
+As such, a straightforward way to mitigate dynamic DRAM power consumption is to try and reduce the amount of refreshes made.
+Several different approches on how this can be done exists and some of them will be presented in this paper.
+Another way to increase power efficency is to do smart scheduling of all DRAM accesses which will give higher performance for the same power consumption.
+However, this will not be examined in any deeper detail in this paper.

### 2014-09-30 paper/include/background.tex
more reorganization
derp me

@@
 \label{sec:bg}
+In this section we give a brief overview of DRAM, covering the organization, basic operation, refresh modes, and trends together with projections for the DRAM structure.
 
-Explain the DRAM structure. Use picture.
+DRAM is hierarichally organized as \reffig{fig:dram_orga} shows.
+At the top of the hierarchy are the channels that are connected to one or more ranks.
+Each channel has individual address, data, and command buses that makes it possible for the channels to operate cuncurrently.
+All ranks on one channel can operate in parallel constrained by the channel bandwith which is shared among the ranks.
+A rank comprizes of several banks where one bank holds a DRAM cell array.
+The banks are thus also able to work in parallel, but are constrianed by both the shared channel bandwith and resources shared among the banks.
 
-Explain basic operations: load and store.
+The DRAM cell array in each bank is structured as \reffig{fig:dram_orgb} shows.
+One cell consists of a transistor and a capacitor, where the transistor is controlled by the wordline wire and connects the capacitor to the bitline wire.
+The wordline connects a number of cells which forms a row.
+Each bitline is connected to a sens amplifier and the row of sens amplifier constitutes the row buffer.
+Data is stored as a charge on the cell capacitor.
+
+\begin{figure*}[t]
+    \centering
+    \subfigure[DRAM hierarcy]{
+        \centering
+        \includegraphics[width=0.4\textwidth]{dram_orga}
+		\label{fig:dram_orga}
+    }
+    \subfigure[DRAM bank structure]{
+        \centering
+        \includegraphics[width=0.4\textwidth]{dram_orgb}
+		\label{fig:dram_orgb}
+    }%
+	\caption{DRAM system organization \cite{raidr}.}
+	\label{fig:dram_org}
+\end{figure*}
+
+To read or write to a DRAM cell the wordline connected to the targeted transistor must first be activated.
+When the wordline is activated, the transistor connects the capacitor to the bitline.
+For this to work, the bitlines has to be precharged to \(V_{DD}\)/2.
+When the capacitor connects to the bitline, charge will either flow from or to the capacitor depending on if the capacitor was charged or not, respectively.
+The sens amplifier will detect the change of voltage and drive the bitline either to \(V_{DD}\) or 0 V.
+Data can then be read from the bitline.
+If data is to be written the corresponding voltage is applied to the bitline, which stores the data in the capacitor.
+Finally, if another row on the bank chall be accessed the open row has to be closed, which disconnects the capacitors from the bitlines.
+
+Over time the capacitors leaks charge and the capacitor has to be recharged in order to preserve the data.
+The operation to recharge a row is called refresh and is performed by opening a row to let the sense amplifiers amplify the detected voltage change and thus recharge the cells to either \(V_{DD}\) or 0 V.
+This follows the same procedure as a read operation and thus is the row refreshed when a read is performed.
+
+The memory controller (MC) normally issues the refresh operation, the auto-refresh command, periodically at an interval of 64 ms for each row, an interval that has been constant for the latests DRAM specifications.
+The interval is halved if the DRAM exeeds the specified working temprature of 85$^{\circ}$C.
+
+The simplest approach to refresh one rank is to refresh all banks in parallel row-by-row at the start of the 64ms interval.
+This approch is called Burst refresh and has some downsides.
+The approch blocks normal accesses and form a peak of power consumption during the burst time.
+A better alternative is to distribute the refreshes across the interval, resulting in a lower average latency for the processor and a more stable power consumption.
 

### 2014-10-01 paper/include/introduction.tex
stuff

@@
 
-DRAM is the most commonly used memory technique as main memory in computer systems today and represents a groving portion of system power consumption.
+DRAM is the most commonly used memory technique as main memory in computer systems today and represents a groving portion of system power consumption \cite{exascale}.
 This portion will continue to increase if nothing is done about it and a large part of the DRAM power consumption comes from refreshing all memory-cells in the DRAM.
@@
 However, this will not be examined in any deeper detail in this paper.
+
+\begin{figure*}[t]
+    \centering
+    \subfigure[DRAM throughput loss]{
+        \centering
+        \includegraphics[width=0.45\textwidth]{dram_throughput}
+		\label{fig:dram_throughput}
+    }
+    \subfigure[DRAM power consumption]{
+        \centering
+        \includegraphics[width=0.45\textwidth]{dram_power}
+		\label{fig:dram_power}
+    }%
+	\caption{Projections on DRAM throughput loss and power consumption \cite{raidr}.}
+	\label{fig:dram_data_proj}
+\end{figure*}

### 2014-10-01 paper/include/rioparis.tex
stuff

@@
-RIO is a technique proposed by Seungjae Baek et al.
+RIO is a technique proposed by S.
+Baek et al.
 \cite{rioparis} in which they identify cells with low retention time (bad cells) and `deletes' them.
@@
 They implement RIO in a Linux kernel which is run and tested on a real hardware system and is able to achieve 87.5\% reduction in refresh count with a performance increas of 4.5\% on average.
-As the implementaion only relies on changes in the kernel and there is no need to modify the hardware this technique would be easy to adopt in systems today.
+It is shown that going for a refresh period longer than \(256ms\) only yields a small performance and power improvment and therefore they does not try to push it higher.
+Their implementaion almost only relies on changes in the kernel and there is no need to modify the hardware so this technique would be easy to adopt in systems today.
 To keep fragmentation of the memory low a maximum of 0.1\% of total memory pages gets `deleted'.
 Even though deletion of a larger amount of pages could decrease the amount of refreshes further it would cause trouble for the kernel which relies on being able to allocate large contiguous address spaces.
-BLASDLBASDAHSUDGASJHDGJHASGDJHASGDJHAGSDJHGASJHDGAJSHDand it is temperature aware.
-Seungjae Baek et al.
-also propose PARIS discussed in \ref{sec:val} which can be combined with RIO for further improvments.
+As retentiontime varies hugely with temperature RIO continously measures the system temperature and compensate the refresh persiod for it.
+S.
+Baek et al.
+also propose PARIS which can be combined with RIO for further improvments.
+\todo[inline]{Write about paris}

### 2014-10-01 paper/include/background.tex
bg work

@@
 		\label{fig:dram_orgb}
-    }%
+    }
 	\caption{DRAM system organization \cite{raidr}.}
@@
 The interval is halved if the DRAM exeeds the specified working temprature of 85$^{\circ}$C.
+The average time between refresh commands are issued by the MC is denoted \(t_{REFI}\) and the time it takes to perform the refresh is denoted \(t_{RFC}\), i.e the time where no normal access requests can be serviced.
 
-The simplest approach to refresh one rank is to refresh all banks in parallel row-by-row at the start of the 64ms interval.
+To refresh one rank, the simplest approach is to refresh all banks in parallel row-by-row at the start of the 64ms interval.
 This approch is called Burst refresh and has some downsides.
-The approch blocks normal accesses and form a peak of power consumption during the burst time.
+The approch blocks normal accesses for a longer period and form a peak of power consumption during the burst time.
 A better alternative is to distribute the refreshes across the interval, resulting in a lower average latency for the processor and a more stable power consumption.
 
-Explain different refresh modes: Burst refresh vs distributed refresh and RAS-only refresh vs CAS before RAS refresh.
+The implementation of a DRAM refresh cycle can be made in two different ways, to either affect one row or many.
+The first is called RAS-only refresh (ROR) where refresh cycle refreshes one targeted row.
+The address for the targeted row has to be put on the bus and it is the MC that makes sure that a row are refreshed within the 64ms interval.
+The second is called CAS before RAS refresh (CBR).
+In this scheme the memory module has an internal address counter that displays which row that shall be refreshed.
+On a refresh cycle, the row stated by the internal address counter is refreshed and the counter is then incremented.
+The advantage with this method is that the address is never put on the bus, which gives a lower power consumption and less bus usage than ROR.
+CBR is the most used method and is the one normally implemented in the auto-refresh command.
+
+When using auto-refresh (CBR), all banks in a rank are blocked for normal accesses.
+All banks are also blocked when ROR is used due to DRAM chip design, however some DRAM chips support per-bank refresh commands that makes it possible to only block the specific bank which is getting refreshed.
+
+As m
+
+\begin{figure}[t]
+    \centering
+    \subfigure[DRAM throughput loss]{
+        \centering
+        \includegraphics[width=0.33\textwidth]{dram_throughput}
+        \label{fig:dram_throughput}
+    }
+    \subfigure[DRAM power consumption]{
+        \centering
+        \includegraphics[width=0.33\textwidth]{dram_power}
+        \label{fig:dram_power}
+    }%
+    \caption{Projections on DRAM throughput loss and power consumption \cite{raidr}.}
+    \label{fig:dram_data_proj}
+\end{figure}
 

### 2014-10-01 paper/include/background.tex
written on bg

@@
 
-As m
-
 \begin{figure}[t]
@@
         \label{fig:dram_power}
-    }%
+    }
     \caption{Projections on DRAM throughput loss and power consumption \cite{raidr}.}
@@
 \end{figure}
+
+The trends of DRAM chips is that they become larger and larger.
+In practice, this is realized to a high degree by adding additional rows in the banks.
+The rows could instead - or also - been made longer by adding DRAM cells, but this solution is not favourable as it increases the power consumption per row access.
 

### 2014-10-02 paper/include/background.tex
made some placeholders for new subsections

@@
 
+\todo[inline]{Drafty stuff below}
+
 The trends of DRAM chips is that they become larger and larger.
@@
 The rows could instead - or also - been made longer by adding DRAM cells, but this solution is not favourable as it increases the power consumption per row access.
+MENTION TIMING ISSUES IF ROWS GET LONGER.
 
-Trends - Storage array rectangular
-            cant go wider (word lines length = timing = power)
-            gets taller = more refreshes
-            over time .... projektions (see Intel)
+When the DRAM chips become larger the throughput loss increases exponentially as \reffig{fig:dram_throughput} shows.
+Also when additional rows are added to the chip, more refresh commands has to be sent and of course gives this many more actual refreshes, especially if CBR is used.
+More refreshes means larger power consumption and is is also so that the refresh bocomes a growing portion of the total DRAM energy consumption.
+In \reffig{fig:dram_power} is is seen that when the device capacity reaches 64 GB the refresh power will nearly constitute half of the total consumption, an increase from 15\% if the device capacity is 8 GB.
+
+In the next years it is projected that ...
+blablabla [source] so it is important and will be more important in the future to lower the total DRAM power consumption and an relatively easy way to do this is to optimizing the DRAM refresh.
+Next.
+we present a bunch of techniques that target DRAM refresh optimization.

### 2014-10-02 paper/include/cream.tex
made some placeholders for new subsections

%CREAM
Do not know where to place it. (RTAV)

\todo[inline]{Ccopy-paste:}
To mitigate the refresh penalty, a Concurrent-REfresh-Aware Memory sys-tem (CREAM) is proposed in this work so that memory access and refresh can be served in parallel. The proposed CREAM architecture distinguishes itself with the following key contributions: (1) Under a given DRAM power budget,
sub-rank-level refresh (SRLR) is developed to reduce refresh power and the saved power is used to enable concurrent memory access; (2) sub-array-level refresh (SALR) is also devised to effectively lower the probability of the conflict between memory access and refresh; (3) In addition, novel sub-array level refresh scheduling schemes, such as sub-array round-robin and dynamic scheduling, are designed to further improve the performance

### 2014-10-02 paper/include/reducing_refresh_power.tex
cleaned
added retention techs to sec 3
more reorganization
made some placeholders for new subsections

@@
 \label{sec:red}
-Roadmap 
-Worth mentioning scheduling real acces togheter with refresh ops
-Write an intro to the ARTV approches.
+\todo[inline]{Do an intro that includes a roadmap, mention the refresh scheduling approach that schedules real acces togheter with refresh ops. Make an intro to ARTV (somewhat done) }
 There are several methods that can be applied to decrease the refresh rate and these can be categorized depending on with sort of information the scheduling decisions are based upon.
@@
 
-\subsection{Types of refreshing}
+\subsection{Approaches to reduce refresh power}
+\todo[inline]{Replace the copy-paste texts}
 
@@
 Process variation [10][16] causes the retention time of DRAM cells to vary across the chip.
-Note that approaches that ex-
-ploit R information are insensitive to system workloads and global memory usage, which makes them attractive ompo
-nents for a refresh-optimized memory subsystem.
+Note that approaches that ex-ploit R information are insensitive to system workloads and global memory usage, which makes them attractive omponents for a refresh-optimized memory subsystem.
 
@@
 
-\subsection{Reducing refresh power approches}
+\subsection{Techniquies to reducing refresh power}
 
-\subsection{Access Recency}
+\subsubsection*{\textbf{Smart Refresh}}
+\label{par:smartrefresh}
+\input{include/smart_refresh}
+
+\subsubsection*{\textbf{Refrint}}
+\label{par:refrint}
+\input{include/refrint}
+
+\subsubsection*{\textbf{DTail-RV}}
+\label{par:dtail}
+\input{include/dtail}
+
+\subsubsection*{\textbf{RAIDR}}
+\label{par:raidr}
+\input{include/raidr}
+
+\subsubsection*{\textbf{RIO and PARIS}}
+\label{par:rioparis}
+\input{include/rioparis}
+
+\subsubsection*{\textbf{SECRET}}
+\label{par:secret}
+\input{include/secret}
+
+\subsubsection*{\textbf{ESKIMO}}
+\label{par:eskimo}
+\input{include/eskimo}
+
+\subsubsection*{\textbf{Sparkk}}
+\label{par:sparkk}
+\input{include/sparkk}
+
+\subsubsection*{\textbf{Flikker}}
+\label{par:flikker}
+\input{include/flikker}
+
+\subsubsection*{\textbf{CREAM}}
+\label{par:cream}
+\input{include/cream}
+
+\subsubsection*{\textbf{Energy Optimization in 3D MPSoCs with Wide-I/O DRAM
+Using Temperature Variation Aware Bank-wise Refresh}}
+\label{par:tempaware}
+\input{include/tempaware}
+
+\subsection{Differences and similarities of the approaches}
+
+Differences and similarities for ARTV approaches.
+
+What in the DRAM devices is modified? / Where is data placed.
+
+How much knowledge does the techniques have about the running software?

### 2014-10-02 paper/include/smart_refresh.tex
made some placeholders for new subsections

@@
+\todo[inline]{Make the text flow}
 The basic concept behind our scheme is that a memory line that has been recently read out or written to does not need to be refreshed again by the periodic refresh mechanism.

### 2014-10-02 paper/include/sparkk.tex
made some placeholders for new subsections

--- 2014-09-19 paper/include/tolerance.tex
@@
-\section{Tolerance} 
-\label{sec:tol}
+%Sparkk
+Tolerance!
 
-Papers that use this technique:
-
-Flikker\cite{flikker}
-
-Sparkk\cite{sparkk}
+\todo[inline]{Ccopy-paste:}
 Multiple authors have suggested techniques where programmers can specify which data is critical and can not tolerate any bit errors and which data can be stored approximately.

### 2014-10-03 paper/include/background.tex
modified bg

@@
 
-DRAM is hierarichally organized as \reffig{fig:dram_orga} shows.
-At the top of the hierarchy are the channels that are connected to one or more ranks.
+DRAM is hierarichally organized as in \reffig{fig:dram_orga}.
+At the top of the hierarchy are the channels that are connected to one or more ranks, each composed of multiple storage arrays called banks.
 Each channel has individual address, data, and command buses that makes it possible for the channels to operate cuncurrently.
-All ranks on one channel can operate in parallel constrained by the channel bandwith which is shared among the ranks.
-A rank comprizes of several banks where one bank holds a DRAM cell array.
-The banks are thus also able to work in parallel, but are constrianed by both the shared channel bandwith and resources shared among the banks.
+All ranks on one channel can also operate in parallel constrained by the channel bandwith which is shared among the ranks.
+Further, the banks within the ranks are also able to operate in parallel, but are constrianed by both the shared channel bandwith and resources shared among the banks.
 
-The DRAM cell array in each bank is structured as \reffig{fig:dram_orgb} shows.
+The DRAM cell array in each bank is structured as in \reffig{fig:dram_orgb}.
 One cell consists of a transistor and a capacitor, where the transistor is controlled by the wordline wire and connects the capacitor to the bitline wire.
 The wordline connects a number of cells which forms a row.
-Each bitline is connected to a sens amplifier and the row of sens amplifier constitutes the row buffer.
+Each bitline is connected to a sens amplifier and the row of sense amplifier constitutes the row buffer.
 Data is stored as a charge on the cell capacitor.
@@
 
-To read or write to a DRAM cell the wordline connected to the targeted transistor must first be activated.
-When the wordline is activated, the transistor connects the capacitor to the bitline.
+To read a row the wordline connected to the rows transistors must first be activated.
+When the wordline is activated, the transistors connects the capacitors to the bitlines.
+When the capacitors connects to the bitlines, charge will either flow from or to the capacitors depending on if the capacitors were charged or not, respectively.
 For this to work, the bitlines has to be precharged to \(V_{DD}\)/2.
-When the capacitor connects to the bitline, charge will either flow from or to the capacitor depending on if the capacitor was charged or not, respectively.
-The sens amplifier will detect the change of voltage and drive the bitline either to \(V_{DD}\) or 0 V.
-Data can then be read from the bitline.
-If data is to be written the corresponding voltage is applied to the bitline, which stores the data in the capacitor.
-Finally, if another row on the bank chall be accessed the open row has to be closed, which disconnects the capacitors from the bitlines.
+The sense amplifiers then detect the change of voltage, which makes it possible to interpret the stored data.
+The charge stored on the capacitors has, as mentioned, left the capacitor and a read operation is thus destructive.
+To maintain the data the capacitors are recarged by the bitlines, which are driven to either either \(V_{DD}\) or 0 V by the sense amplifiers depending on the former detected voltage change.
+Finally, if another row on the bank shall be accessed the open row has to be closed, which disconnects the capacitors from the bitlines.
+
+To write to a row the corresponding wordline is first activated.
+As during a read, the transistors then connects the capacitors to the bitlines and charge will either flow from or to the capacitor.
+The row buffer are then loaded with the data that are driven to the bitlines by the sense amplifiers, thus storing the information on the capacitors.
+The row then remains open, as by a read, until another row shall be accessed, then the row is closed which disconnects the capacitors from the bitlines.
 

### 2014-10-03 paper/include/introduction.tex

--- 2014-09-30 paper/include/introduction.tex
@@
 \label{sec:int}
-%\textbf{Whats the problem?}
-Technology is ever advancing and computer systems of today are more powerfull and energy effecient than ever before.
-These improvments have to a large part the continuos decrease in integrated circuit feature size to thank but as we now are getting close to the physiological limit on how small transistors can become possible improvments have to be looked for elsewhere.
-The fact that battery powered mobile systems are getting more and more common and that power consupmtion correspons to a large part of the cost in driving the data centers of today makes energy consumption an increasingly big system design constraint.
-%\textbf{Why is it important?}
+Computer systems today are more powerfull and energy effecient than ever before.
+Battery powered mobile systems are getting more and more common and power consupmtion correspons to a large part of the cost in driving large data centers.
+This makes energy consumption an increasingly big system design constraint.
 
-DRAM is the most commonly used memory technique as main memory in computer systems today and represents a groving portion of system power consumption.
-This portion will continue to increase if nothing is done about it and a large part of the DRAM power consumption comes from refreshing all memory-cells in the DRAM.
-This refreshing has to be done as DRAM store data as a charge in a capacitor and given todays silicon, this charge leak over time.
+DRAM is the most commonly used main memory technology in computer systems today and represents a groving portion of system power consumption \cite{exascale}.
+This portion will continue to increase as we strive for higher and higher memory capacity.
+A large part of the DRAM power consumption comes from refreshing all memory-cells in the DRAM.
+This refreshing has to be done as DRAM store data as a charge in a capacitor and charges leak over time.
 To increase DRAM performance and lower access latencies 3D stacked DRAM has been proposed in which the working temperature will be higher than in regular DRAM.
-This poses a problem as the rate at which the charge leaks increases with temperature and the DRAM will have to be refreshed more often.
+This increases the leakage and the DRAM will have to be refreshed more often.
 
-Power consumption in DRAM consists of two parts, static power and dynamic power.
-The static power is caused by leakage and the dynamic power is the result of all accesse to the memory.
-As a refresh basically is the same as a regular read access execpt that the data is not used, and thereby consumes the same amount of energy as a usefull access.
-As such, a straightforward way to mitigate dynamic DRAM power consumption is to try and reduce the amount of refreshes made.
-Several different approches on how this can be done exists and some of them will be presented in this paper.
-Another way to increase power efficency is to do smart scheduling of all DRAM accesses which will give higher performance for the same power consumption.
-However, this will not be examined in any deeper detail in this paper.
+There are two main ways to decrease power consumption in DRAM.
+Minimizing unnecessary operations and scheduling the needed ones smarter.
+As such, one straightforward way to mitigate DRAM power consumption is to try and reduce the amount of refreshes which are operations that yield no data transfer and can therefore be seen as unnecessary.
+Several different approches on how this can be done exists and some of them will be surveyed in this paper.
+The other way to decrease power consupmtion, scheduling of all needed DRAM accesses smarter, will not be examined in any deeper detail in this paper.

### 2014-10-03 paper/include/introduction.tex
mod intro

@@
 
-DRAM is the most commonly used main memory technology in computer systems today and represents a groving portion of system power consumption \cite{exascale}.
+DRAM is the most commonly used main memory technology in modern computer systems and represents a growing portion of system power consumption \cite{exascale}.
 This portion will continue to increase as we strive for higher and higher memory capacity.
@@
 Minimizing unnecessary operations and scheduling the needed ones smarter.
-As such, one straightforward way to mitigate DRAM power consumption is to try and reduce the amount of refreshes which are operations that yield no data transfer and can therefore be seen as unnecessary.
-Several different approches on how this can be done exists and some of them will be surveyed in this paper.
+As such, one straightforward way to mitigate DRAM power consumption is to reduce the amount of unnecessary refreshes operations.
+Several different approaches on how this can be done exists and some of them are surveyed in this paper.
 The other way to decrease power consupmtion, scheduling of all needed DRAM accesses smarter, will not be examined in any deeper detail in this paper.

### 2014-10-03 paper/include/background.tex
workn

@@
 The row buffer are then loaded with the data that are driven to the bitlines by the sense amplifiers, thus storing the information on the capacitors.
-The row then remains open, as by a read, until another row shall be accessed, then the row is closed which disconnects the capacitors from the bitlines.
+The row will be closed when another row shall be accessed.
 
-Over time the capacitors leaks charge and the capacitor has to be recharged in order to preserve the data.
-The operation to recharge a row is called refresh and is performed by opening a row to let the sense amplifiers amplify the detected voltage change and thus recharge the cells to either \(V_{DD}\) or 0 V.
+As mentioned, the capacitors leaks charge over time and has to be recharged in order to preserve the data.
+The refresh is performed by opening a row to let the sense amplifiers amplify the detected voltage change and thus recharge the cells to either \(V_{DD}\) or 0 V.
 This follows the same procedure as a read operation and thus is the row refreshed when a read is performed.
 
-The memory controller (MC) normally issues the refresh operation, the auto-refresh command, periodically at an interval of 64 ms for each row, an interval that has been constant for the latests DRAM specifications.
-The interval is halved if the DRAM exeeds the specified working temprature of 85$^{\circ}$C.
-The average time between refresh commands are issued by the MC is denoted \(t_{REFI}\) and the time it takes to perform the refresh is denoted \(t_{RFC}\), i.e the time where no normal access requests can be serviced.
+The memory controller (MC) normally issues the refresh operation periodically at an interval of 64 ms for each row, an interval that has been constant for the latests DRAM specifications [source].
+This interval is halved if the DRAM exeeds the specified working temprature of 85$^{\circ}$C.
+%The average time between refresh commands are issued by the MC is denoted \(t_{REFI}\) and the time it takes to perform the refresh is denoted \(t_{RFC}\), i.e the time where no normal access requests can be serviced.
+% Maybe not needed, we shall see.
 
@@
 The advantage with this method is that the address is never put on the bus, which gives a lower power consumption and less bus usage than ROR.
-CBR is the most used method and is the one normally implemented in the auto-refresh command.
 
-When using auto-refresh (CBR), all banks in a rank are blocked for normal accesses.
+When using CBR, all banks in a rank are blocked for normal accesses.
 All banks are also blocked when ROR is used due to DRAM chip design, however some DRAM chips support per-bank refresh commands that makes it possible to only block the specific bank which is getting refreshed.
@@
     }
-    \caption{Projections on DRAM throughput loss and power consumption \cite{raidr}.}
+    \caption{Projections on DRAM throughput loss and power consumption in extended-temprature operation \cite{raidr}.}
     \label{fig:dram_data_proj}
 \end{figure}
-
-\todo[inline]{Drafty stuff below}
 
@@
 The rows could instead - or also - been made longer by adding DRAM cells, but this solution is not favourable as it increases the power consumption per row access.
-MENTION TIMING ISSUES IF ROWS GET LONGER.
 
-When the DRAM chips become larger the throughput loss increases exponentially as \reffig{fig:dram_throughput} shows.
-Also when additional rows are added to the chip, more refresh commands has to be sent and of course gives this many more actual refreshes, especially if CBR is used.
-More refreshes means larger power consumption and is is also so that the refresh bocomes a growing portion of the total DRAM energy consumption.
-In \reffig{fig:dram_power} is is seen that when the device capacity reaches 64 GB the refresh power will nearly constitute half of the total consumption, an increase from 15\% if the device capacity is 8 GB.
+As refresh operations are made by row, the refresh operations will increase liniary with the device capacity.
+In turn, the throughput loss increases exponentially as in \reffig{fig:dram_throughput}.
+Additional refresh operations gives larger power consumption and the refreshes becomes a growing portion of the total DRAM energy consumption as seen in \reffig{fig:dram_power}.
+When the device capacity reaches 64 GB the refresh power nearly constitute half of the total consumption, an increase from 15\% if the device capacity is 8 GB.
+
+\todo[inline]{Drafty stuff below}
 

### 2014-10-03 paper/include/reducing_refresh_power.tex
workn

@@
 \subsection{Approaches to reduce refresh power}
-\todo[inline]{Replace the copy-paste texts}
 
 \subsubsection*{\textbf{Access Recency}}
-DRAM accesses imply refresh operations, and so a subsequent refresh to the same row can be postponed.
-The number of rows affected depends on how many different rows are accessed within the maximum refresh period, which may be few for many workloads.
-Ghosh et al.
-[9] propose Smart Refresh, which divides the refresh period into phases and maintains a per-row timeout counter in the MC.
-The MC decrements the counters each phase and issues a RAS-only refresh when a counter hits zero.
+As mentioned, a DRAM access is performed in the same way as a refresh, which makes it possible to delay the next refresh operation to an accessed row without endangering the stored data.
+All accessed rows within the maximum refresh period result in that a refresh can be postponed.
+This implies that if the system seldom accesses the DRAM data, or only a small portion of it, the approach's gain diminishes.
 
 \subsubsection*{\textbf{Retention Time}}
+.
+
+\todo[inline]{Replace the copy-paste texts}
 Retention time refers to the period during which cells hold valid values as charge gradually leaks.
@@
 \subsubsection*{\textbf{Data Tolerance}}
+.
+
+\todo[inline]{Replace the copy-paste texts}
 Applications like games, media processing, machine learning, and unstructured information analysis tolerate errors in portions of their data and still produce acceptably accurate results.
@@
 \subsubsection*{\textbf{Validity}}
+.
+
+\todo[inline]{Replace the copy-paste texts}
 If the OS has not allocated the page frame, its data are meaningless, and refreshes to it are wasteful. Severa

### 2014-10-03 paper/include/rioparis.tex
writewrite
u

@@
-RIO is a technique proposed by S.
+RIO is a \textbf{R} based technique proposed by S.
 Baek et al.
@@
 They implement RIO in a Linux kernel which is run and tested on a real hardware system and is able to achieve 87.5\% reduction in refresh count with a performance increas of 4.5\% on average.
-It is shown that going for a refresh period longer than \(256ms\) only yields a small performance and power improvment and therefore they does not try to push it higher.
+It is shown that going for a refresh period longer than \(256ms\) only yields a small performance and power improvment and therefore they does not push it higher.
 Their implementaion almost only relies on changes in the kernel and there is no need to modify the hardware so this technique would be easy to adopt in systems today.
@@
 As retentiontime varies hugely with temperature RIO continously measures the system temperature and compensate the refresh persiod for it.
+
 S.
@@
 also propose PARIS which can be combined with RIO for further improvments.
-\todo[inline]{Write about paris}
+In PARIS the goal is not to make the refresh period as long as possible but to not refresh unused memory rows at all using \textbf{V} data.
+This is done by keeping track of all page frames allocated and deallocated by the linux kernel.
+The page frame data is translated to DRAM row data and kept in a bitmap with one full/empty bit per row group.
+A row group consist of rows from several DRAM chips.
+More rows per group gives a smaller storage overhead for the bitmap, but lowers the refresh granluarity.
+
+As these two techniques is orthogonal to each other they can be combined, meaning that PARIS removes unnecessary refreshes and RIO lowers the rate at which refreshes are performed.
+This is similar to what Z.
+Cui et al.
+does with DTail-RV.
+\todo[inline]{compare to DTAIL-RV}

### 2014-10-06 lab/MOESI_SMPCache.cpp
labbilabibalbiiblalallb

@@ -1,0 +1,267 @@
+#include "MESI_SMPCache.h"
+
+//extern void computeOpt(std::vector<uint64_t> &candidates, uint64_t &repl_addr);
+
+MESI_SMPCache::MESI_SMPCache(int cpuid, std::vector<SMPCache * > * cacheVector,
+                 int csize, int cassoc, int cbsize, int caddressable, const char * repPol, bool cskew) : 
+               SMPCache(cpuid,cacheVector){
+  fprintf(stderr,"Making a MESI cache with cpuid %d size %d assoc %d bsize %d\n",cpuid, csize, cassoc, cbsize);
+  CacheGeneric<MESI_SMPCacheState> *c = CacheGeneric<MESI_SMPCacheState>::create(csize, cassoc, cbsize, caddressable, repPol, cskew);
+  cache = (CacheGeneric<StateGeneric<> >*)c;
+}
+
+void MESI_SMPCache::fillLine(uint64_t addr, uint64_t mesi_state){
+  MESI_SMPCacheState *st = (MESI_SMPCacheState *)cache->findLine2Replace(addr); //this gets the contents of whateverline this would go into
+  if(st==0){
+    return;
+  }
+  //if(st->getTag() == llc_cache->calcTag(addr) && st->getTag() != 0)
+  //std::cout << "WTFL1" << std::endl;
+
+  if(st->getState() != 0x00000100 && optCache && (!hasLLC) && this->getCacheVector()->size() == 1) {
+    //Find and use repl candidate computed using OPT
+    std::vector<uint64_t> temp;
+    uint64_t repl_addr;
+    //Get other candidates from the set
+    cache->getCandidates(addr, temp);
+    //Compute the opt replacement candidate tag
+    //computeOpt(temp, repl_addr);
+    //Use the corresponding tags addres
+    st = (MESI_SMPCacheState *)cache->findLine(repl_addr);
+    if(!st || (st && !(st->isValid())) ){
+      std::cout << "This is not possible!" << std::endl;
+    }
+    temp.clear();
+  }
+    
+  //std::cout << "Evicting tag " << st->getTag() << " Inserting tag " << cache->calcTag(addr) << std::endl;  
+  st->setTag(cache->calcTag(addr));
+  st->changeStateTo((MESIState_t)mesi_state);
+  return;
+    
+}
+  
+void MESI_SMPCache::fillLLC(uint64_t addr){
[…]

### 2014-10-06 lab/kernel/parallel.c

@@ -1,0 +1,135 @@
+/* The sequential version of equation solver kernel used in Ocean */
+/* The Solve function is implemented with red-black ordering mechanism */
+/* You should only modify the Solve function to paralize the kernel */
+
+#include<stdio.h>
+#include<stdlib.h>
+#include<omp.h>
+
+
+int n;
+double **A, diff;
+int nthreads;
+
+void print(double **A);
+int input(char *filename);
+void Solve(double **A);
+
+int main(int argc, char **argv)
+{
+
+    if(argc != 3) {
+	printf("Usage: %s <input file> <nthreads>\n", argv[0]);
+	exit(1);
+    }
+            
+    nthreads = atoi(argv[2]);
+    if(input(argv[1]) < 0 && nthreads < 0 ) {
+	printf("Usage error occured while opening/reading file, make sure file exists and is in the correct format or nthreads > 0\n");
+	exit(1);
+    }
+
+    //Set number of threads in the parallel version
+    omp_set_num_threads(nthreads);
+
+    //kernel call
+    Solve(A);
+
+    //print(A);
+
+    return 0;
+}
+
+
+void Solve(double **A)
+{
+    int i, j, done = 0;
+    double temp;
+	double diff;
+
+    #pragma omp parallel default(shared) private(i, j, temp)    
+    while(!done){
+
+   	 	#pragma omp single
+		diff = 0;
+
+		/* sweep that updates red elements */
+		#pragma omp for  schedule(static) reduction(+:diff)
+		for(i=1; i<=n; ++i){
+		    for(j=i%2 ? 1:2; j<=n; j=j+2){
+				temp = A[i][j];
+				A[i][j] = 0.2 * (A[i][j] + A[i][j-1] + A[i-1][j] + A[i][j+1] + A[i+1][j]);
+				diff += abs(A[i][j] - temp);
+		    }
+		}
+
+		/* sweep that updates black elements */
+		#pragma omp for schedule(static) reduction(+:diff)
+		for(i=1; i<=n; ++i){
+		    for(j=i%2 ? 2:1; j<=n; j=j+2){
+				temp = A[i][j];
+				A[i][j] = 0.2 * (A[i][j] + A[i][j-1] + A[i-1][j] + A[i][j+1] + A[i+1][j]);
+				diff += abs(A[i][j] - temp);
+		    }
[…]

### 2014-10-06 lab/kernel/serial.c

@@ -1,0 +1,128 @@
+/* The sequential version of equation solver kernel used in Ocean */
+/* The Solve function is implemented with red-black ordering mechanism */
+/* You should only modify the Solve function to paralize the kernel */
+
+#include<stdio.h>
+#include<stdlib.h>
+#include<omp.h>
+
+
+int n;
+double **A, diff;
+int nthreads;
+
+void print(double **A);
+int input(char *filename);
+void Solve(double **A);
+
+int main(int argc, char **argv)
+{
+
+    if(argc != 3) {
+	printf("Usage: %s <input file> <nthreads>\n", argv[0]);
+	exit(1);
+    }
+            
+    nthreads = atoi(argv[2]);
+    if(input(argv[1]) < 0 && nthreads < 0 ) {
+	printf("Usage error occured while opening/reading file, make sure file exists and is in the correct format or nthreads > 0\n");
+	exit(1);
+    }
+
+    //Set number of threads in the parallel version
+    omp_set_num_threads(2);
+
+    //kernel call
+    Solve(A);
+
+    //print(A);
+
+    return 0;
+}
+
+
+void Solve(double **A)
+{
+    int i, j, done = 0;
+    double temp;
+	double diff;
+	
+    while(!done){
+	diff = 0;
+
+		/* sweep that updates red elements */
+		for(i=1; i<=n; ++i){
+		    for(j=i%2 ? 1:2; j<=n; j=j+2){
+			temp = A[i][j];
+			A[i][j] = 0.2 * (A[i][j] + A[i][j-1] + A[i-1][j] + A[i][j+1] + A[i+1][j]);
+			diff += abs(A[i][j] - temp);
+		    }
+		}
+
+		/* sweep that updates black elements */
+		for(i=1; i<=n; ++i){
+		    for(j=i%2 ? 2:1; j<=n; j=j+2){
+			temp = A[i][j];
+			A[i][j] = 0.2 * (A[i][j] + A[i][j-1] + A[i-1][j] + A[i][j+1] + A[i+1][j]);
+			diff += abs(A[i][j] - temp);
+		    }
+		}
+
+		{
+			if(diff / (n*n) < 0.01)
+			    done = 1;
+
+			//printf("%.4lf\n", diff);
+    	}
+    }
+}
+
+
+void print(double **A)
+{
+    int i, j;
+    for(i=1; i<=n; ++i){
[…]

### 2014-10-07 paper/include/background.tex
spellchecked

@@
 
-DRAM is hierarichally organized as in \reffig{fig:dram_orga}.
+DRAM is hierarchically organized as in \reffig{fig:dram_orga}.
 At the top of the hierarchy are the channels that are connected to one or more ranks, each composed of multiple storage arrays called banks.
-Each channel has individual address, data, and command buses that makes it possible for the channels to operate cuncurrently.
-All ranks on one channel can also operate in parallel constrained by the channel bandwith which is shared among the ranks.
-Further, the banks within the ranks are also able to operate in parallel, but are constrianed by both the shared channel bandwith and resources shared among the banks.
+Each channel has individual address, data, and command buses that makes it possible for the channels to operate concurrently.
+All ranks on one channel can also operate in parallel constrained by the channel bandwidth which is shared among the ranks.
+Further, the banks within the ranks are also able to operate in parallel, but are constrained by both the shared channel bandwidth and resources shared among the banks.
 
 The DRAM cell array in each bank is structured as in \reffig{fig:dram_orgb}.
-One cell consists of a transistor and a capacitor, where the transistor is controlled by the wordline wire and connects the capacitor to the bitline wire.
-The wordline connects a number of cells which forms a row.
-Each bitline is connected to a sens amplifier and the row of sense amplifier constitutes the row buffer.
+One cell consists of a transistor and a capacitor, where the transistor is controlled by the word line wire and connects the capacitor to the bit line wire.
+The word line connects a number of cells which forms a row.
+Each bit line is connected to a sens amplifier and the row of sense amplifier constitutes the row buffer.
 Data is stored as a charge on the cell capacitor.
@@
     \centering
-    \subfigure[DRAM hierarcy]{
+    \subfigure[DRAM hierarchy]{
         \centering
@@
 
-To read a row the wordline connected to the rows transistors must first be activated.
-When the wordline is activated, the transistors connects the capacitors to the bitlines.
-When the capacitors connects to the bitlines, charge will either flow from or to the capacitors depending on if the capacitors were charged or not, respectively.
-For this to work, the bitlines has to be precharged to \(V_{DD}\)/2.
+To read a row the word line connected to the rows transistors must first be activated.
+When the word line is activated, the transistors connects the capacitors to the bit lines.
+When the capacitors connects to the bit lines, charge will either flow from or to the capacitors depending on if the capacitors were charged or not, respectively.
+For this to work, the bit lines has to be precharged to \(V_{DD}\)/2.
 The sense amplifiers then detect the change of voltage, which makes it possible to interpret the stored data.
 The charge stored on the capacitors has, as mentioned, left the capacitor and a read operation is thus destructive.
-To maintain the data the capacitors are recarged by the bitlines, which are driven to either either \(V_{DD}\) or 0 V by the sense amplifiers depending on the former detected voltage change.
-Finally, if another row on the bank shall be accessed the open row has to be closed, which disconnects the capacitors from the bitlines.
+To maintain the data the capacitors are recharged by the bit lines, which are driven to either either \(V_{DD}\) or 0 V by the sense amplifiers depending on the former detected voltage change.
+Finally, if another row on the bank shall be accessed the open row has to be closed, which disconnects the capacitors from the bit lines.
 
-To write to a row the corresponding wordline is first activated.
-As during a read, the transistors then connects the capacitors to the bitlines and charge will either flow from or to the capacitor.
-The row buffer are then loaded with the data that are driven to the bitlines by the sense amplifiers, thus storing the information on the capacitors.
+To write to a row the corresponding word line is first activated.
+As during a read, the transistors then connects the capacitors to the bit lines and charge will either flow from or to the capacitor.
+The row buffer are then loaded with the data that are driven to the bit lines by the sense amplifiers, thus storing the information on the capacitors.
 The row will be closed when another row shall be accessed.
@@
 
-The memory controller (MC) normally issues the refresh operation periodically at an interval of 64 ms for each row, an interval that has been constant for the latests DRAM specifications [source].
-This interval is halved if the DRAM exeeds the specified working temprature of 85$^{\circ}$C.
+The memory controller (MC) normally issues the refresh operation periodically at an interval of 64 ms for each row, an interval that has been constant for the latest DRAM specifications [source].
+This interval is halved if the DRAM exceeds the specified working temperature of 85$^{\circ}$C.
 %The average time between refresh commands are issued by the MC is denoted \(t_{REFI}\) and the time it takes to perform the refresh is denoted \(t_{RFC}\), i.e the time where no normal access requests can be serviced.
@@
 To refresh one rank, the simplest approach is to refresh all banks in parallel row-by-row at the start of the 64ms interval.
-This approch is called Burst refresh and has some downsides.
-The approch blocks normal accesses for a longer period and form a peak of power consumption during the burst time.
+This approach is called Burst refresh and has some downsides.
+The approach blocks normal accesses for a longer period and form a peak of power consumption during the burst time.
 A better alternative is to distribute the refreshes across the interval, resulting in a lower average latency for the processor and a more stable power consumption.
@@
     }
-    \caption{Projections on DRAM throughput loss and power consumption in extended-temprature operation \cite{raidr}.}
+    \caption{Projections on DRAM throughput loss and power consumption in extended-temperature operation \cite{raidr}.}
     \label{fig:dram_data_proj}
@@
 In practice, this is realized to a high degree by adding additional rows in the banks.
-The rows could instead - or also - been made longer by adding DRAM cells, but this solution is not favourable as it increases the power consumption per row access.
+The rows could instead - or also - been made longer by adding DRAM cells, but this solution is not favorable as it increases the power consumption per row access.
 
-As refresh operations are made by row, the refresh operations will increase liniary with the device capacity.
+As refresh operations are made by row, the refresh operations will increase linear with the device capacity.
 In turn, the throughput loss increases exponentially as in \reffig{fig:dram_throughput}.

### 2014-10-07 paper/include/dtail.tex
more reorganization
stuff
u
wiriten cool stuff on all aproaches, or something.

@@
-Zehan Cui et al.
-\cite{dtail} propses DTail-R in which data for line refresh rate is stored in the DRAM itself instead of storing it in the memory controller which many other solutions do \cite{raidr}\cite{smartrefresh}\cite{refrint}.
+Z.
+Cui et al.
+\cite{dtail} proposes DTail-R in which data for line refresh rate is stored in the DRAM itself instead of storing it in the memory controller which many other solutions do \cite{raidr}\cite{smartrefresh}\cite{refrint}.
 To counter the added latency of a slower memory they utilizes spatial locality of the refresh data and performs prefetching into a FIFO buffer.
 The retention time data consists of 4 bits and is interpreted in such way that the refresh period becomes \(64ms \times 2^n, n = [0...7]\).
-With this structure the amount of memory required is weighing in on a negliable amount of 0.006\% DRAM capacity (i.e.
+With this structure the amount of memory required is weighing in on a negligible amount of 0.006\% DRAM capacity (i.e.
 1.92MB in a 32GB DRAM).
-One thing that DTail-R misses is refreshrate compensation for temperature variations.
-As the refresh rate is set per line DTail-R gets a higher resolution than techniques that set refreshrate on a higher level.
+One thing that DTail-R misses is refresh rate compensation for temperature variations.
+As the refresh rate is set per line DTail-R gets a higher resolution than techniques that set refresh rate on a higher level.
 Therefore DTail-R manages to decrease the amount of refreshes with as much as 87.9\% and thereby only have 16.7\% energy overhead compared to a refresh less 32GB DRAM.
-Zehan Cui et al.
-also proposes DTail-V \cite{dtail} which can be combined with DTail-R and is further discussed in \ref{sec:val}.
+
+Another technique proposed by Z. Cui et al. is DTail-V \cite{dtail} which can be combined with DTail-R. DTail-V uses \textbf{V} data to mitigate unnecessary refreshes.

### 2014-10-07 paper/include/reducing_refresh_power.tex
wiriten cool stuff on all aproaches, or something.

@@
 \subsubsection*{\textbf{Retention Time}}
-.
-
-\todo[inline]{Replace the copy-paste texts}
-Retention time refers to the period during which cells hold valid values as charge gradually leaks.
-Process variation [10][16] causes the retention time of DRAM cells to vary across the chip.
-Note that approaches that ex-ploit R information are insensitive to system workloads and global memory usage, which makes them attractive omponents for a refresh-optimized memory subsystem.
+The need for refresh is as mentioned caused by leakage in the memory cells.
+Retention time is a measurement of the time it takes before a cell loose its data if not refreshed (recharged).
+Variations in the manufacturing process causes fluctuation in the retention time of the DRAM cells on the same chip.
+To acomodate for the cells with the lowest retention time manufaturers produce chips with average retention times much higher than possible minimum.
+This gives room for optimications using variable refresh rate for different cells.
+The retention time is also highly sensitive to temperature variations and becomes drastically lower at higher temperatures.
+For techniques using this aproach the biggest advatage is that the impovments are independent of workload, except for the potential rise in temperature.
 
 \subsubsection*{\textbf{Data Tolerance}}
-.
+In the most aproaches all data loss is seen as intolerable.
+In this aproach data is categorised depending on how critical it is and e.g.
+aproximate computations, games, or media processing can tollerate some bit errors while still producing a acceptable result.
+This yields that cells containing data from such applications may be refreshed more seldom.
+Using this aproach the energy and performance savings achieved is highly depending on what kind of applications is most commonly run on the system.
 
-\todo[inline]{Replace the copy-paste texts}
-Applications like games, media processing, machine learning, and unstructured information analysis tolerate errors in portions of their data and still produce acceptably accurate results.
+% Applications like games, media processing, machine learning, and unstructured information analysis tolerate errors in portions of their data and still produce acceptably accurate results.
 Approximate computation [31, 8] exploits this approximate data to realize tradeoffs among performance, energy, and accuracy.
@@
 \subsubsection*{\textbf{Validity}}
-.
+In a system that does not uitilze all the availible memory space it is unnesessary to refresh parts not used.
+Therefore, cells containing data that are not valid does not need to be refreshed.
+This aproach is highly sensitive to memory uitilization of the system and its efects on energy consumption and performance improvments will be diminishing on systems with high workload.
 
-\todo[inline]{Replace the copy-paste texts}
-If the OS has not allocated the page frame, its data are meaningless, and refreshes to it are wasteful. Severa
-l software approaches thus attempt to trigger refreshes only for rows with valid data.
+%If the OS has not allocated the page frame, its data are meaningless, and refreshes to it are wasteful.
+Several software approaches thus attempt to trigger refreshes only for rows with valid data.
 The effectiveness of schemes using such V refresh information is sensitive to the total memory usage of the system.

### 2014-10-07 paper/include/background.tex
wiriten cool stuff on all aproaches, or something.

@@
 At the top of the hierarchy are the channels that are connected to one or more ranks, each composed of multiple storage arrays called banks.
-Each channel has individual address, data, and command buses that makes it possible for the channels to operate concurrently.
-All ranks on one channel can also operate in parallel constrained by the channel bandwidth which is shared among the ranks.
-Further, the banks within the ranks are also able to operate in parallel, but are constrained by both the shared channel bandwidth and resources shared among the banks.
+Each channel has individual address, data, and command buses that makes it possible for the channels to operate cuncurrently.
+All ranks on one channel can also operate in parallel constrained by the channel bandwith which is shared among the ranks.
+Further, the banks within the ranks are also able to operate in parallel, but are constrianed by both the shared channel bandwith and resources shared among the banks.
 

### 2014-10-07 paper/include/background.tex
workwork

--- 2014-10-07 paper/include/background.tex
@@
 
-To refresh one rank, the simplest approach is to refresh all banks in parallel row-by-row at the start of the 64ms interval.
+To refresh one rank, the simplest approach is to refresh all banks in parallel row-by-row at the start of the 64 ms interval.
 This approach is called Burst refresh and has some downsides.
@@
 The first is called RAS-only refresh (ROR) where refresh cycle refreshes one targeted row.
-The address for the targeted row has to be put on the bus and it is the MC that makes sure that a row are refreshed within the 64ms interval.
+The address for the targeted row has to be put on the bus and it is the MC that makes sure that a row are refreshed within the 64 ms interval.
 The second is called CAS before RAS refresh (CBR).
@@
 
-\todo[inline]{Drafty stuff below}
-
-In the next years it is projected that ...
-blablabla [source] so it is important and will be more important in the future to lower the total DRAM power consumption and an relatively easy way to do this is to optimizing the DRAM refresh.
-Next.
-we present a bunch of techniques that target DRAM refresh optimization.
+The conclusion drawn from \reffig{fig:dram_data_proj} is that it will be increasingly important to mitigate the throughput loss and DRAM power consumption of future systems.
+One way to decrease these negative trends of DRAM scaling is to optimize the DRAM refresh.

### 2014-10-07 paper/include/smart_refresh.tex
workwork

--- 2014-09-30 paper/include/smart_refresh.tex
@@
-The basic concept behind our scheme is that a memory line that has been recently read out or written to does not need to be refreshed again by the periodic refresh mechanism.
+M.
+Ghosh and H.-H.
+S.
+Lee proposes the technique Smart Refresh \cite{smartrefresh} that targets Access Recency.
+The basic idea is to maintain a time-out counter for each memory row stored in the MC and these counters is set if the corresponding row is accessed.
+When a counter time-out, the MC issues a ROR refresh to the corresponding row and also sets the counter.
 
-Uses a time-out counter for each memory line.
-The counter is set on a read/write/refresh and dicards refreshes until the timer hits zero.
-The counter is associated to a bank/line pair.
-The counters are stored in the memory controller.
-One counter consists of 2 SRAM bits.
-More bits give higher granularity and thus also better performance because the refresh is made less often.
+The counters are decremented in such a fashion so the refreshes are distributed evenly over time.
+A higher bit size on the counters increases time granularity, which results in that a counter can be set later in time and thus postpones the refresh further.
+
+Can be deactivated
+
+As mentioned, Smart Refresh modifies the MC. The memory footprint of the technique is 0.0024\% of total device capacity when 3-bits counters are used and assumptions , which is stored in the MC.
+
+Results
 
@@
 
-Implemented so only one line per each logical segments are deremented at the same time, which leads to that the refreshes are evenly distributed over time. (I.e. no burst refresh.)
+Implemented so only one line per each logical segments are decremented at the same time, which leads to that the refreshes are evenly distributed over time. (I.e. no burst refresh.)
 
@@
 
+move:
+Smart refresh:
 Employed its solution to 2 GB ram.

### 2014-10-07 paper/include/reducing_refresh_power.tex
added old retention time text as comment

@@
 \label{sec:red}
-\todo[inline]{Do an intro that includes a roadmap, mention the refresh scheduling approach that schedules real acces togheter with refresh ops. Make an intro to ARTV (somewhat done) }
+\todo[inline]{Do an intro that includes a road map, mention the refresh scheduling approach that schedules real access together with refresh ops. Make an intro to ARTV (somewhat done) }
 There are several methods that can be applied to decrease the refresh rate and these can be categorized depending on with sort of information the scheduling decisions are based upon.
 The four sort of information are: access recency, cell retention times, error tolerance of the data, and line validity.
-This categorization follows the taxonamy proposed by authors \cite{dtail}.
+This categorization follows the taxonomy proposed by authors \cite{dtail}.
 
@@
 Variations in the manufacturing process causes fluctuation in the retention time of the DRAM cells on the same chip.
-To acomodate for the cells with the lowest retention time manufaturers produce chips with average retention times much higher than possible minimum.
-This gives room for optimications using variable refresh rate for different cells.
+To accommodate for the cells with the lowest retention time manufacturers produce chips with average retention times much higher than possible minimum.
+This gives room for optimization using variable refresh rate for different cells.
 The retention time is also highly sensitive to temperature variations and becomes drastically lower at higher temperatures.
-For techniques using this aproach the biggest advatage is that the impovments are independent of workload, except for the potential rise in temperature.
+For techniques using this approach the biggest advantage is that the improvements are independent of workload, except for the potential rise in temperature.
+
+%Retention time refers to the period during which cells hold valid values as charge gradually leaks.
+Process variation [10][16] causes the retention time of DRAM cells to vary across the chip.
+Note that approaches that ex-ploit R information are insensitive to system workloads and global memory usage, which makes them attractive omponents for a refresh-optimized memory subsystem.
 
 \subsubsection*{\textbf{Data Tolerance}}
-In the most aproaches all data loss is seen as intolerable.
-In this aproach data is categorised depending on how critical it is and e.g.
-aproximate computations, games, or media processing can tollerate some bit errors while still producing a acceptable result.
+In the most approaches all data loss is seen as intolerable.
+In this approach data is categorized depending on how critical it is and e.g.
+approximate computations, games, or media processing can tolerate some bit errors while still producing a acceptable result.
 This yields that cells containing data from such applications may be refreshed more seldom.
-Using this aproach the energy and performance savings achieved is highly depending on what kind of applications is most commonly run on the system.
+Using this approach the energy and performance savings achieved is highly depending on what kind of applications is most commonly run on the system.
 
@@
 \subsubsection*{\textbf{Validity}}
-In a system that does not uitilze all the availible memory space it is unnesessary to refresh parts not used.
+In a system that does not utilize all the available memory space it is unnecessary to refresh parts not used.
 Therefore, cells containing data that are not valid does not need to be refreshed.
-This aproach is highly sensitive to memory uitilization of the system and its efects on energy consumption and performance improvments will be diminishing on systems with high workload.
+This approach is highly sensitive to memory utilization of the system and its effects on energy consumption and performance improvements will be diminishing on systems with high workload.
 
@@
 
-\subsection{Techniquies to reducing refresh power}
+\subsection{Techniques to reducing refresh power}
 

### 2014-10-07 paper/include/dtail.tex
Rewrite of DTail. Hopfully better!

--- 2014-09-30 paper/include/dtail.tex
@@
-Zehan Cui et al.
-\cite{dtail} propses DTail-R in which data for line refresh rate is stored in the DRAM itself instead of storing it in the memory controller which many other solutions do \cite{raidr}\cite{smartrefresh}\cite{refrint}.
+Z.
+Cui et al.
+\cite{dtail} proposes DTail.
+They evaluates two versions of DTail referenced as DTail-R and DTail-V utilizing \textbf{R} and \textbf{V} data respecively.
+Even so, their main idea is that more types of refresh information can be incorperated and stored together with the \textbf{R} and \textbf{V} information.
+DTail can then choose the most efficient refresh schema depending on the current situation, i.e.
+workload and type of application.
+Refresh data is kept per row and is stored in the DRAM itself instead of storing it in the memory controller which many other solutions do \cite{raidr}\cite{smartrefresh}\cite{refrint}.
 To counter the added latency of a slower memory they utilizes spatial locality of the refresh data and performs prefetching into a FIFO buffer.
-The retention time data consists of 4 bits and is interpreted in such way that the refresh period becomes \(64ms \times 2^n, n = [0...7]\).
-With this structure the amount of memory required is weighing in on a negliable amount of 0.006\% DRAM capacity (i.e.
-1.92MB in a 32GB DRAM).
-One thing that DTail-R misses is refreshrate compensation for temperature variations.
-As the refresh rate is set per line DTail-R gets a higher resolution than techniques that set refreshrate on a higher level.
-Therefore DTail-R manages to decrease the amount of refreshes with as much as 87.9\% and thereby only have 16.7\% energy overhead compared to a refresh less 32GB DRAM.
-Zehan Cui et al.
-also proposes DTail-V \cite{dtail} which can be combined with DTail-R and is further discussed in \ref{sec:val}.
+As the refreash data is kept on a row level DTail can save more refreshes than other techniques that have a lower granularity.
+
+For DTail-R, the retention time data consists of 3 bits and is interpreted in such way that the refresh period can become \(64ms \times 2^n, n = [0...7]\).
+With DTail-V one additional bit is added to the refresh data which is used to indicate if the row is valid or not.
+With this structure of four bits per row, the amount of memory required is weighing in on a negligible amount of 0.006\% DRAM capacity assuming a row size of 8 KB.
+
+When a row is to be scheduled for refresh the refresh data is checked.
+If the row is valid and the time since the last refresh is \(64ms \times 2^n, n = [0...7]\) a refresh is performed, otherwise not.
+The refreshes can be done either by RAS commands or by auto refresh depending on what is most efficient at the time.
+If a large amount of rows needs refresh then auto refresh is selected, otherwise RAS-only refresh are used.
+
+When only DTail-R is used, they manages to decrease the amount of refreshes with as much as 87.9\% and thereby only have 16.7\% energy overhead compared to a refresh less 32GB DRAM.
+With only DTail-V power consumption and performance overhead increase from 4\% and 3\% to 90\% and 23\% respectively when going from 10\% to 90\% memory capacity utilization compared to a refresh less 32 GB system.
+The power of DTail shows when DTail-R and DTail-V are used together and at 10\% memory utilization a performance overhead of only 1\% and close to 0\% increase in energy consumption is achived.
+When the utilization is increased to 90\% they increase to 2\% and 15\% respectively.
+%
+
+%One thing that DTail-R misses is refresh rate compensation for temperature variations.

### 2014-10-08 paper/include/background.tex
wrote stuff

@@
 To refresh one rank, the simplest approach is to refresh all banks in parallel row-by-row at the start of the 64 ms interval.
-This approach is called Burst refresh and has some downsides.
+This approach is called burst refresh and has some downsides.
 The approach blocks normal accesses for a longer period and form a peak of power consumption during the burst time.
@@
 One way to decrease these negative trends of DRAM scaling is to optimize the DRAM refresh.
+However, a reduction of the refreshes made has no linear relationship to the reduction of refresh energy consumption.
+The energy reduction depends on the state of the bank in which the targeted row resides; if the targeted row is closed more energy will be consumed than if the targeted row would have been open.

### 2014-10-08 paper/include/refrint.tex
more reorganization
made some placeholders for new subsections
wrote stuff

@@
+This technique is proposed by A.
+Agrawal et al.
+\cite{refrint} and is based on Access Recency.
+It is targeted towards caches and eDRAM and thus can not be directly compared with the other solutions, but the main concept is of good use and the technique could be implemented DRAM as well.
+The concept is based on two types of unnecessary refreshes which originates from so called cold and hot rows.
+Cold rows are those used far apart in time or not used at all.
+The technique identifies and refreshes only those rows which are expected to be accessed in the near future, the rows that are used less frequent is flushed and invalidated.
+Hot rows consists of the rows that gets accessed frequently and thus can the technique postpone the preceding refresh to the accessed rows.
+
+For the hot lines, an approach similar to Smart Refresh is employed. The differences are that Refrint does not maintain a counter for each row, but a snapshot of a global counter.
+
+\todo[inline]{Crunchable text below!}
+
 Tageted towards caches and eDRAM.
@@
 
-\subsection{Differences and similarities}
+\subsubsection*{Differences and similarities}
 

### 2014-10-08 paper/include/smart_refresh.tex
wrote stuff

--- 2014-09-30 paper/include/smart_refresh.tex
@@
-The basic concept behind our scheme is that a memory line that has been recently read out or written to does not need to be refreshed again by the periodic refresh mechanism.
+M.
+Ghosh and H.-H.
+S.
+Lee proposes the technique Smart Refresh \cite{smartrefresh} that targets Access Recency.
+The basic idea is to maintain a time-out counter for each memory row and when a counter time-out, the MC issues a ROR refresh to the corresponding row.
+The counters are set if the row is accessed or refreshed.
 
-Uses a time-out counter for each memory line.
-The counter is set on a read/write/refresh and dicards refreshes until the timer hits zero.
-The counter is associated to a bank/line pair.
-The counters are stored in the memory controller.
-One counter consists of 2 SRAM bits.
-More bits give higher granularity and thus also better performance because the refresh is made less often.
+The counters are decremented in such a fashion so the refreshes are distributed evenly over time and thus avoiding burst refresh.
+To achive this, the counters are initially staggered and divided into n logical blocks where the counters are accessed in a sequential order.
+Only n counters will be accessed, set or decremented at the same point in time, which result in that maximum n rows are refreshed simultainously.
 
-This scheme will be most effective if all lines are accessed before a refresh is needed.
-If no accesses are made = the entire working set fits into cache, Smart Refresh becomes CBR (CAS before RAS).
-Only refresh one line at a time = inefficient, so they have made it possible to turn the technique off.
+The counters bit size corresponds to the time granularity in where an access can be captured.
+A higher bit size results in that an access can be captured later in time, which effectively postpones the refresh further.
+The least optimal case for postponing the refresh is if the row access occurs a small time intervall before the counter is accessed.
+If a 2-bit counter is used the least optimal row access result in that the row is refreshed after 3/4 of the maximum refresh period.
+Analogously, a 3-bit counter result in that a row is refreshed after minimum 7/8 of the maximum refresh period.
 
-Implemented so only one line per each logical segments are deremented at the same time, which leads to that the refreshes are evenly distributed over time. (I.e. no burst refresh.)
+If the row accesses drops below 1\% during a maximum refresh period, Smart Refresh is disabled.
+As stated, an access recency technique is dependent on row accesses to be of gain.
+When disabled, the MC uses it's default CBR refresh scheme.
+Smart refresh is then activated if the row accesses increases above 2\%.
 
-Overhead: 48KB for 2GB and 768 KB for 32GB = 0,0024\% overhead.
+Smart Refresh modifies the MC but does not change the DRAM module nor the interface between them.
+The modifications of the MC constitutes of counter update logic, the counter array, and a pending refresh request queue.
+The memory footprint of the modifications is 0.0024\% of total device capacity when 3-bits counters are used.
 
-Average reduction of refreshes by 59.3\% for 2GB and 52.3\% energy saving regarding the DRAM refresh. No performance degradation.
-
-Employed its solution to 2 GB ram.
-What about todays sizes?
-How effective is the technique now?
-- Only works effective if large parts the memory is acccessed, i.e.
-large part of the memory has to be valid to begin with.
-Is it so today?
+The technique result in an average refresh reduction of 59\% and saves on average 52\% of refresh power consumption without any performance degradation.
+The simulation was performed on 2 GB of main memory.

### 2014-10-08 paper/include/flikker.tex
wrote, added TODOs and sources

%flikker
Tolerance!\cite{flikker}
\todo[inline]{Add description.}

### 2014-10-08 paper/include/reducing_refresh_power.tex
Rewrite of DTail. Hopfully better!
wrote stuff
wrote, added TODOs and sources

@@
 
-\subsubsection*{\textbf{DTail-RV}}
+\subsubsection*{\textbf{DTail}}
 \label{par:dtail}
@@
 \subsection{Differences and similarities of the approaches}
+\todo[inline]{Drafty text below!}
 
@@
 How much knowledge does the techniques have about the running software?
+
+Smart Refresh:
+Employed its solution to 2 GB ram.
+What about todays sizes?
+How effective is the technique now?
+- Only works effective if large parts the memory is acccessed, i.e.
+large part of the memory has to be valid to begin with.
+Is it so today?
+
+Smart Refresh \& Refrint:
+Refrint is somewhat a extension of Smart Refresh.
+Smart Refresh's features can be laid somewhat equal to Refrint's time-based polices, wheras the data-based polices have no eqvivalence in Smart Refresh.
+
+They both modify the memory controller, and almost in the same way.
+But Refrint has no timers that is getting decremented the whole time, and instead copies a global "timer" to the line associated bits upon access.
+Of cource Refrint's data-based polices also adds logic to the memory controller.
+
+Discuss energy savings. Hard to do eDRAM vs DRAM...

### 2014-10-08 paper/include/refrint.tex
wrote, added TODOs and sources

@@
 Cold rows are those used far apart in time or not used at all.
-The technique identifies and refreshes only those rows which are expected to be accessed in the near future, the rows that are used less frequent is flushed and invalidated.
+The technique identifies and refreshes only those rows which are expected to be accessed in the near future, the rows that are used less frequent is invalidated.
 Hot rows consists of the rows that gets accessed frequently and thus can the technique postpone the preceding refresh to the accessed rows.
 
-For the hot lines, an approach similar to Smart Refresh is employed. The differences are that Refrint does not maintain a counter for each row, but a snapshot of a global counter.
+For the hot rows, an approach similar to Smart Refresh is employed.
+The differences are that Refrint does not maintain a counter for each row, but a snapshot of a global counter as well as the row's valid bit.
+From experiments, they found out that a good size for the global counter was 1 or 2 bits.
+When a row is accessed, the corresponding snapshot is updated with the current value of the global counter.
+Whenever the global counter steps to a new value, all normal accesses are stalled and the logic checks whether any of the valid row's snapshot  matches the global counter.
+If there is a match, a refresh is scheduled or the row invalidated depending on the polices used for the cold rows.
 
-\todo[inline]{Crunchable text below!}
+Four different policies can be applied for the cold rows, where the policies decides what to refresh; \textit{All}, \textit{Valid}, \textit{Dirty}, or \textit{WB(n,m)}.
+\textit{All} refreshes every row, irregardless of whether it is valid or not.
+\textit{Valid} and \textit{Dirty} refreshes valid and dirty rows, respectively, and otherwise invalidate the row.
+\textit{WB(n,m)} refreshes a dirty row \textit{n} times before flushing it and changing the row state to valid clean, then the row is refreshed \textit{m} times after the last access before it is invalidated.
+\textit{N} and \textit{m} are time-out counters that are decremented upon refresh.
 
-Tageted towards caches and eDRAM.
+\todo[inline]{Drafty text below!}
 
-Our goal is to refresh only the data that will be used in the near future, and only refresh if it is really needed. The other data is invalidated and/ or written back to main memory.
-
-The technique target to minimize the refresh for the so called hot and cold lines.
-Hot lines are those being accessed and thus unnesssesary refreshed by the refresh mechanism, whereas the cold lines are those that hasn't been used in a while but still gets refreshed.
-
-To minimize the refresh for hot lines time-based policies are used.
-Normally a counter - Polyphase - is used, which practicly works in the same way as the time-out counters for Smart Refresh.
-It consists of two bits.
-They also has a basic periodic refresh scheme (why?).
-
-For the cold lines four different (data-based) policies can be applied: \textit{all}, \textit{valid}, \textit{dirty} or WB(n,m).
-\textit{All} refreshes all lines.
-\textit{Valid} refreshes the valid lines.
-\textit{Dirty} refreshes the dirty lines.
-The WB policy is associated with a tuple (n, m).
-WB refreshes a Dirty line that is not being accessed for n times before writing it back and changing its state to Valid Clean; moreover, it refreshes a Valid Clean line that is not being accessed for m times before invalidating it.
-WB retains a Dirty line in the cache Ion ger because evicting it has the additional cost of writing the line back to lower-Ievel memory.
-To implement WB, we maintain a per-Iine Count.
-W hen the line is read or written, Count is set to n (if Dirty) or m (if Valid Clean).
-In addition, when the line is refreshed, Count is decremented.
-W hen Count reaches zero, the line is either written back or invalidated.
-Note that the Dirty policy is equivalent to WB(oo,O), while Valid is equivalent to WB(oo,oo).
-Finally, every policy refreshes cache lines in transient states as well.
-
-They categorize applications based on how they are affected by the data-based policies.
-Fig.
-\ref{fig:app_cat} presents an application categorization based on two axes: application footprint and visibility.
-It is likely........
-TODO
+Depending on the application's memory footprint and cache visibility, the program can be categorized to maximize the gain of the technique.
+Visibility means that the In \reffig{fig:app_cat} three categories are displayed.
+Explain the categories.
 
@@
 	\includegraphics[width=\textwidth/2]{app_cat}
-	\caption{Application categorization according to the data-based policy \cite{refrint}.}
+	\caption{Application categorization according to the cold row's policies \cite{refrint}.}
 	\label{fig:app_cat}
@@
 
-Implementation: Hot lines.
-They use a global counter with M bits (M = 1 or 2).
-When a line is accessed, the global counter is copied to N bits associated to the line and those N bits are stored with the valid bit in the memory controller.
-If N = 2, the overhead becomes 0.6\%.
-When the global counter steps to a new phase, all normal accesses are stalled and the logic controlls the valid lines and its local phase bits to see if it matches the global phase.
-If there is a match, a refresh is scheduled.
-Otherwise the memory goes back to normal operation after the scan.
+The technique modifies the MC. Which modifications and how large overhead? 5 bit snapshot for each row + state and n+m counters. some logic as well.
 
-Implementation: Cold lines. For each line there is a 5 bit counter and the line's state bits. Depending on the data policy  for each line, but they are somehow neglectible.  ...
-
-Remember that this technique is targeted towards caches and eDRAM and only is tested in that enviroment.
-This means that we can not really use the results shown in their paper, but the techniques can be applied to DRAM.
-
-\subsubsection*{Differences and similarities}
-
-Refrint is somewhat a extension of Smart Refresh.
-Smart Refresh's features can be laid somewhat equal to Refrint's time-based polices, wheras the data-based polices have no eqvivalence in Smart Refresh.
-
-They both modify the memory controller, and almost in the same way.
-But Refrint has no timers that is getting decremented the whole time, and instead copies a global "timer" to the line associated bits upon access.
-Of cource Refrint's data-based polices also adds logic to the memory controller.
-
-Discuss energy savings. Hard to do eDRAM vs DRAM...
+Presents the results. Add that we can not really use the result as they are from eDRAM.

### 2014-10-09 paper/include/refrint.tex
work

@@
 
-Depending on the application's memory footprint and cache visibility, the program can be categorized to maximize the gain of the technique.
-Visibility means that the In \reffig{fig:app_cat} three categories are displayed.
-Explain the categories.
-
-\begin{figure}[t!]
-	\includegraphics[width=\textwidth/2]{app_cat}
-	\caption{Application categorization according to the cold row's policies \cite{refrint}.}
-	\label{fig:app_cat}
-\end{figure}
+Depending on the application's memory footprint and cache visibility the program is categorized to maximize the gain of the cold row's policies.
+Visibility correspond to the amount of row accesses made by the application.
+If the application has high visibility \textit{WB} is used, and if the visibility is low \textit{Valid} is used.
+%In turn, if the visibility is high and the footprint small  with small \textit{n} and \textit{m}.
 

### 2014-10-09 paper/include/dtail.tex
write write

@@
 \cite{dtail} proposes DTail.
-They evaluates two versions of DTail referenced as DTail-R and DTail-V utilizing \textbf{R} and \textbf{V} data respecively.
-Even so, their main idea is that more types of refresh information can be incorperated and stored together with the \textbf{R} and \textbf{V} information.
+They evaluates two versions of DTail referenced as DTail-R and DTail-V utilizing \textbf{R} and \textbf{V} data respectively.
+Even so, their main idea is that more types of refresh information can be incorporated and stored together with the \textbf{R} and \textbf{V} information.
 DTail can then choose the most efficient refresh schema depending on the current situation, i.e.
@@
 To counter the added latency of a slower memory they utilizes spatial locality of the refresh data and performs prefetching into a FIFO buffer.
-As the refreash data is kept on a row level DTail can save more refreshes than other techniques that have a lower granularity.
+As the refresh data is kept on a row level DTail can save more refreshes than other techniques that have a lower granularity.
 
@@
 With only DTail-V power consumption and performance overhead increase from 4\% and 3\% to 90\% and 23\% respectively when going from 10\% to 90\% memory capacity utilization compared to a refresh less 32 GB system.
-The power of DTail shows when DTail-R and DTail-V are used together and at 10\% memory utilization a performance overhead of only 1\% and close to 0\% increase in energy consumption is achived.
+The power of DTail shows when DTail-R and DTail-V are used together and at 10\% memory utilization a performance overhead of only 1\% and close to 0\% increase in energy consumption is achieved.
 When the utilization is increased to 90\% they increase to 2\% and 15\% respectively.
-%
-
-%One thing that DTail-R misses is refresh rate compensation for temperature variations.

### 2014-10-09 paper/include/introduction.tex
write write

@@
 \label{sec:int}
-Computer systems today are more powerfull and energy effecient than ever before.
-Battery powered mobile systems are getting more and more common and power consupmtion correspons to a large part of the cost in driving large data centers.
+Computer systems today are more powerful and energy efficient than ever before.
+Battery powered mobile systems are getting more and more common and power consumption corresponds to a large part of the cost in driving large data centers.
 This makes energy consumption an increasingly big system design constraint.
@@
 Several different approaches on how this can be done exists and some of them are surveyed in this paper.
-The other way to decrease power consupmtion, scheduling of all needed DRAM accesses smarter, will not be examined in any deeper detail in this paper.
+The other way to decrease power consumption, scheduling of all needed DRAM accesses smarter, will not be examined in any deeper detail in this paper.

### 2014-10-09 paper/include/raidr.tex
stuff
write write

@@
-RAIDR by Jinghui Liu et al.
+RAIDR by J.
+Liu et al.
 \cite{raidr} divides all memory lines into different groups, called bins, based on the cell with the shortest retention time in the line.
-In the basic configuration there are two bins, one with a refreshrate of \(64ms - 128ms\) and one with \(128ms - 256ms\).
+In the basic configuration there are two bins, one with a refresh rate of \(64ms - 128ms\) and one with \(128ms - 256ms\).
 Lines in the first bin gets refreshed every \(64ms\), lines in the second bin every \(128ms\), and lines not in any bin gets refreshed at the default refresh rate which is set to \(256ms\).
@@
 The storage requirement can however vary depending the amount of bins used.
-RAIDR also acomodate for temperature variations by scaling all bins refresh rates using a timer.
-With RAIDR a performance improvment of 4.1\% and a decreas in powerconsumption by 8.3\% is achieved in a system with 32GB DRAM compared to auto-refresh.
+RAIDR also accommodate for temperature variations by scaling all bins refresh rates using a timer.
+With RAIDR a performance improvement of 4.1\% and a decrease in power consumption by 8.3\% is achieved in a system with 32GB DRAM compared to auto-refresh.

### 2014-10-09 paper/include/reducing_refresh_power.tex
write write

@@
 How effective is the technique now?
-- Only works effective if large parts the memory is acccessed, i.e.
+- Only works effective if large parts the memory is accessed, i.e.
 large part of the memory has to be valid to begin with.
@@
 Refrint is somewhat a extension of Smart Refresh.
-Smart Refresh's features can be laid somewhat equal to Refrint's time-based polices, wheras the data-based polices have no eqvivalence in Smart Refresh.
+Smart Refresh's features can be laid somewhat equal to Refrint's time-based polices, whereas the data-based polices have no equivalence in Smart Refresh.
 
@@
 But Refrint has no timers that is getting decremented the whole time, and instead copies a global "timer" to the line associated bits upon access.
-Of cource Refrint's data-based polices also adds logic to the memory controller.
+Of course Refrint's data-based polices also adds logic to the memory controller.
 

### 2014-10-09 paper/include/rioparis.tex
write write

--- 2014-09-30 paper/include/rioparis.tex
@@
-RIO is a technique proposed by Seungjae Baek et al.
+RIO is a \textbf{R} based technique proposed by S.
+Baek et al.
 \cite{rioparis} in which they identify cells with low retention time (bad cells) and `deletes' them.
 This is done by changing the memory allocating modules in the OS to not allocate pages that have bad cells in them.
-They implement RIO in a Linux kernel which is run and tested on a real hardware system and is able to achieve 87.5\% reduction in refresh count with a performance increas of 4.5\% on average.
-As the implementaion only relies on changes in the kernel and there is no need to modify the hardware this technique would be easy to adopt in systems today.
+They implement RIO in a Linux kernel which is run and tested on a real hardware system and is able to achieve 87.5\% reduction in refresh count with a performance increase of 4.5\% on average.
+It is shown that going for a refresh period longer than \(256ms\) only yields a small performance and power improvement and therefore they does not push it higher than that.
+Their implementation almost only relies on changes in the kernel and there is no need to modify the hardware so this technique would be easy to adopt in systems today.
 To keep fragmentation of the memory low a maximum of 0.1\% of total memory pages gets `deleted'.
 Even though deletion of a larger amount of pages could decrease the amount of refreshes further it would cause trouble for the kernel which relies on being able to allocate large contiguous address spaces.
-BLASDLBASDAHSUDGASJHDGJHASGDJHASGDJHAGSDJHGASJHDGAJSHDand it is temperature aware.
-Seungjae Baek et al.
-also propose PARIS discussed in \ref{sec:val} which can be combined with RIO for further improvments.
+As retention time varies hugely with temperature RIO continuously measures the system temperature and compensate the refresh period for it.
+
+S.
+Baek et al.
+also propose PARIS which can be combined with RIO for further improvements.
+In PARIS the goal is not to make the refresh period as long as possible but to not refresh unused memory rows at all using \textbf{V} data.
+This is done by keeping track of all page frames allocated and deallocated by the Linux kernel.
+The page frame data is translated to DRAM row data and kept in a bitmap with one full/empty bit per row group.
+A row group consist of rows from several DRAM chips and more rows per group gives a smaller storage overhead for the bitmap, but lowers the refresh granularity.
+The bitmap is stored in the memory controller and as PARIS utilizes RAS-only refresh this needs to be supported.
+
+As mentioned earlier, RIO and PARIS can be combined which results in PARIS removing unnecessary refreshes and RIO lowering the rate at which the performed refreshes are done.
+This is similar to what Z.
+Cui et al.
+does with DTail-RV.
+\todo[inline]{compare to DTAIL-RV}

### 2014-10-10 lab/report/main.tex
modified the make script

\documentclass[a4paper]{report}

\include{packages}

%MACROS---
\newcommand{\reffig}[1]{Fig. \ref{#1}}

\begin{document}
\section{test}
heeeej!!! :)
\end{document}

### 2014-10-10 paper/include/background.tex
changed make and corrected text

@@
 DRAM is hierarchically organized as in \reffig{fig:dram_orga}.
-At the top of the hierarchy are the channels that are connected to one or more ranks, each composed of multiple storage arrays called banks.
+At the top of the hierarchy are the \textit{channels} that are connected to one or more \textit{ranks}, each composed of multiple storage arrays called \textit{banks}.
 Each channel has individual address, data, and command buses that makes it possible for the channels to operate concurrently.
-All ranks on one channel can also operate in parallel constrained by the channel bandwidth which is shared among the ranks.
-Further, the banks within the ranks are also able to operate in parallel, but are constrained by both the shared channel bandwidth and resources shared among the banks.
+All ranks on one channel can operate in parallel, but are constrained by the channel bandwidth, which is shared among the ranks.
+Further, the banks within the ranks operate in parallel, but are constrained by both the shared channel bandwidth and resources shared among the banks.
 
 The DRAM cell array in each bank is structured as in \reffig{fig:dram_orgb}.
-One cell consists of a transistor and a capacitor, where the transistor is controlled by the word line wire and connects the capacitor to the bit line wire.
+One cell consists of a transistor and a capacitor, where the transistor is controlled by the \textit{word line} wire and connects the capacitor to the \textit{bit line} wire.
 The word line connects a number of cells which forms a row.
-Each bit line is connected to a sens amplifier and the row of sense amplifier constitutes the row buffer.
+Each bit line is connected to a sense amplifier, and the row of sense amplifiers constitutes the \textit{row buffer}.
 Data is stored as a charge on the cell capacitor.
@@
 
-To read a row the word line connected to the rows transistors must first be activated.
-When the word line is activated, the transistors connects the capacitors to the bit lines.
-When the capacitors connects to the bit lines, charge will either flow from or to the capacitors depending on if the capacitors were charged or not, respectively.
-For this to work, the bit lines has to be precharged to \(V_{DD}\)/2.
+To read a row, the word line connected to the row's transistors must first be activated.
+When the word line is activated, the transistors connect the capacitors to the bit lines.
+When the capacitors connect to the bit lines, charge will either flow from or to the capacitors depending on whether they were charged or not, respectively.
+For this to work, the bit lines have to be precharged to $V_{DD}/2$.
 The sense amplifiers then detect the change of voltage, which makes it possible to interpret the stored data.
 The charge stored on the capacitors has, as mentioned, left the capacitor and a read operation is thus destructive.
-To maintain the data the capacitors are recharged by the bit lines, which are driven to either either \(V_{DD}\) or 0 V by the sense amplifiers depending on the former detected voltage change.
-Finally, if another row on the bank shall be accessed the open row has to be closed, which disconnects the capacitors from the bit lines.
+To maintain the data the capacitors are recharged by the bit lines, which are driven to either either $V_{DD}/2$ or $0\:V$ by the sense amplifiers depending on the former voltage detected.
+Finally, when another row on the bank is accessed, the open row has to be closed, which disconnects the capacitors from the bit lines.
 
-To write to a row the corresponding word line is first activated.
-As during a read, the transistors then connects the capacitors to the bit lines and charge will either flow from or to the capacitor.
-The row buffer are then loaded with the data that are driven to the bit lines by the sense amplifiers, thus storing the information on the capacitors.
-The row will be closed when another row shall be accessed.
+To write to a row, the corresponding word line is first activated.
+As during a read, the transistors then connect the capacitors to the bit lines, and charge will either flow from or to the capacitor.
+The row buffer is then loaded with the data that are driven to the bit lines by the sense amplifiers, thus storing the information on the capacitors.
+The row will be closed when another row is accessed.
 
-As mentioned, the capacitors leaks charge over time and has to be recharged in order to preserve the data.
-The refresh is performed by opening a row to let the sense amplifiers amplify the detected voltage change and thus recharge the cells to either \(V_{DD}\) or 0 V.
-This follows the same procedure as a read operation and thus is the row refreshed when a read is performed.
+As mentioned, the capacitors leak charge over time and have to be recharged in order to preserve the data.
+The refresh is performed by opening a row to let the sense amplifiers magnify the detected voltage change and thus recharge the cells to either $V_{DD}/2$ or $0\:V$.
+This follows the same procedure as a read operation and thus is the row always refreshed when a read is performed.
+% Is it clear that the row is refreshed during a write?
 
-The memory controller (MC) normally issues the refresh operation periodically at an interval of 64 ms for each row, an interval that has been constant for the latest DRAM specifications [source].
-This interval is halved if the DRAM exceeds the specified working temperature of 85$^{\circ}$C.
-%The average time between refresh commands are issued by the MC is denoted \(t_{REFI}\) and the time it takes to perform the refresh is denoted \(t_{RFC}\), i.e the time where no normal access requests can be serviced.
-% Maybe not needed, we shall see.
+The memory controller (MC) normally issues refresh operations periodically at an interval of $64\:ms$ for each row, an interval that has been constant for the latest DRAM specifications [source].
+This interval is conservatively halved if the DRAM exceeds the specified working temperature of $85^{\circ}C$.
 
-To refresh one rank, the simplest approach is to refresh all banks in parallel row-by-row at the start of the 64 ms interval.
-This approach is called burst refresh and has some downsides.
-The approach blocks normal accesses for a longer period and form a peak of power consumption during the burst time.
-A better alternative is to distribute the refreshes across the interval, resulting in a lower average latency for the processor and a more stable power consumption.
+To refresh one rank, the simplest approach is to refresh all banks in parallel row-by-row at the start of the $64\:ms$ interval.
+This approach is called \textit{burst refresh}.
+The approach blocks normal accesses for a long period, and power consumption peaks during the burst time.
+A better alternative is to use \textit{distributed refresh}, which scatter the refreshes across the interval, resulting in a lower average latency for the processor and more stable power consumption.
 
@@
 The first is called RAS-only refresh (ROR) where refresh cycle refreshes one targeted row.
-The address for the targeted row has to be put on the bus and it is the MC that makes sure that a row are refreshed within the 64 ms interval.
+The address for the targeted row has to be put on the bus and it is the MC that makes sure that a row are refreshed within the $64\:ms$ interval.
 The second is called CAS before RAS refresh (CBR).

### 2014-10-10 paper/include/introduction.tex
changed make and corrected text

@@
 \label{sec:int}
-Computer systems today are more powerful and energy efficient than ever before.
-Battery powered mobile systems are getting more and more common and power consumption corresponds to a large part of the cost in driving large data centers.
-This makes energy consumption an increasingly big system design constraint.
+
+Energy consumption in modern computer system is an increasingly big system design constraint.
+For example, battery powered mobile systems are getting more common and they needs to be energy efficient to have a long battery life.
+Moreover, in large data centers the power consumption corresponds to a large part of the upkeep cost.
 
@@
 A large part of the DRAM power consumption comes from refreshing all memory-cells in the DRAM.
-This refreshing has to be done as DRAM store data as a charge in a capacitor and charges leak over time.
-To increase DRAM performance and lower access latencies 3D stacked DRAM has been proposed in which the working temperature will be higher than in regular DRAM.
-This increases the leakage and the DRAM will have to be refreshed more often.
+This refreshing has to be done as DRAM's store data as a charge in a capacitor, and these charges leak over time.
+To increase DRAM performance and lower access latencies, 3D stacked DRAM has been proposed, but stacked design have trouble dissipating heat.
+The higher temperature increase leakage, and the DRAM will have to be refreshed more often.
 
-There are two main ways to decrease power consumption in DRAM.
-Minimizing unnecessary operations and scheduling the needed ones smarter.
-As such, one straightforward way to mitigate DRAM power consumption is to reduce the amount of unnecessary refreshes operations.
-Several different approaches on how this can be done exists and some of them are surveyed in this paper.
-The other way to decrease power consumption, scheduling of all needed DRAM accesses smarter, will not be examined in any deeper detail in this paper.
+There are two main ways to decrease power consumption in DRAM; minimizing unnecessary operations and scheduling the needed ones smarter.
+As such, one straightforward way to mitigate DRAM power consumption is to reduce the number of unnecessary refresh operations.
+Here we survey some proposed approaches to avoid or delaying refresh.
+Smarter scheduling, while promising interesting, is beyond the scope of this paper.

### 2014-10-10 paper/include/background.tex
corrected

@@
 
-The memory controller (MC) normally issues refresh operations periodically at an interval of $64\:ms$ for each row, an interval that has been constant for the latest DRAM specifications [source].
+The memory controller (MC) normally issues refresh operations periodically at an interval of $64\:ms$ for each row, an interval that has been constant for the latest DRAM specifications [source] and called \textit{maximum refresh period}.
 This interval is conservatively halved if the DRAM exceeds the specified working temperature of $85^{\circ}C$.
@@
 The implementation of a DRAM refresh cycle can be made in two different ways, to either affect one row or many.
-The first is called RAS-only refresh (ROR) where refresh cycle refreshes one targeted row.
-The address for the targeted row has to be put on the bus and it is the MC that makes sure that a row are refreshed within the $64\:ms$ interval.
-The second is called CAS before RAS refresh (CBR).
-In this scheme the memory module has an internal address counter that displays which row that shall be refreshed.
+The first is called \textit{RAS-only refresh} (ROR) where a refresh cycle refreshes one targeted row.
+To target one row, the row's address is put on the bus, and the MC have the responsibility that a row is refreshed within the maximum refresh period.
+The second is called \textit{CAS before RAS refresh} (CBR).
+In this scheme, the memory module have an internal address counter that displays which row that will be refreshed.
 On a refresh cycle, the row stated by the internal address counter is refreshed and the counter is then incremented.
-The advantage with this method is that the address is never put on the bus, which gives a lower power consumption and less bus usage than ROR.
+Contrary to ROR, CBR do not need provided addresses on the bus, which results in a lower power consumption and less bus usage.
+% During CBR, a row in each bank is updated, right?
+if so.
+write that in the text.
 
 When using CBR, all banks in a rank are blocked for normal accesses.
-All banks are also blocked when ROR is used due to DRAM chip design, however some DRAM chips support per-bank refresh commands that makes it possible to only block the specific bank which is getting refreshed.
+All banks are also blocked when ROR is used due to DRAM chip design, however some DRAM chips support per-bank refresh commands that makes it possible to only block the specific bank being refreshed.
 
@@
 
-The trends of DRAM chips is that they become larger and larger.
+The trends in DRAM chips is to increase capacity with each generation.
 In practice, this is realized to a high degree by adding additional rows in the banks.
@@
 
-As refresh operations are made by row, the refresh operations will increase linear with the device capacity.
-In turn, the throughput loss increases exponentially as in \reffig{fig:dram_throughput}.
-Additional refresh operations gives larger power consumption and the refreshes becomes a growing portion of the total DRAM energy consumption as seen in \reffig{fig:dram_power}.
-When the device capacity reaches 64 GB the refresh power nearly constitute half of the total consumption, an increase from 15\% if the device capacity is 8 GB.
+As refresh operations are made by row, the refresh operations will increase linearly with the device capacity.
+In turn, the throughput loss increases exponentially, as shown in \reffig{fig:dram_throughput}.
+Additional refresh operations give larger power consumption and the refreshes become a growing portion of the total DRAM energy consumption, as shown in \reffig{fig:dram_power}.
+When the device capacity reaches $64\:GB$, the refresh power nearly constitutes half of the total consumption, an increase from 15\% if the device capacity is 8 GB.
 
-The conclusion drawn from \reffig{fig:dram_data_proj} is that it will be increasingly important to mitigate the throughput loss and DRAM power consumption of future systems.
+The conclusion to draw from \reffig{fig:dram_data_proj} is that it will be increasingly important to mitigate the throughput loss and DRAM power consumption of future systems.
 One way to decrease these negative trends of DRAM scaling is to optimize the DRAM refresh.
-However, a reduction of the refreshes made has no linear relationship to the reduction of refresh energy consumption.
+However, a reduction in the number of refreshes made has no linear relationship to the reduction in refresh energy consumption.
 The energy reduction depends on the state of the bank in which the targeted row resides; if the targeted row is closed more energy will be consumed than if the targeted row would have been open.

### 2014-10-11 lab/report/include/lab1.tex
added todo

\section{Lab 1}
\label{sec:lab1}
\todo[inline]{Introduction, one paragraph}

\subsection{Task 1}
\todo[inline]{Explain the benchmarks}

\subsection{Task 2}
\todo[inline]{
Compare MSI and MESI.
When is having the E state beneficial (over baseline MSI)?
How does the behavior change when the dataset size changes?}

\subsection{Task 3}
\todo[inline]{
Compare MESI and MESI+migratory.
Why is one better than the other?
How does the behavior change when the dataset size changes?}

### 2014-10-11 lab/report/include/lab2.tex
added todo

\section{Lab 2}
\label{sec:lab2}
\todo[inline]{Introduction, one paragraph}

\subsection{Task 1}
\todo[inline]{Analyze the source code of the benchmark and provide a detailed description of how it works. Comment on the prominent communication patterns in the benchmark.}

\subsection{Task 2}
\todo[inline]{Explain our parallel version}

\subsection{Task 3}
\todo[inline]{
Evaluate MSI, MESI, and MESI+migratory in the context of this benchmark. Perform a detailed characterization of the benchmark using different protocols. Discuss how the behavior changes when the working set size is modified. Show how execution time relates to the cache access statistics.}

\subsection{Task 4}
\todo[inline]{
Outline a strategy to modify the MESI protocol to support the O state.
Analyze the MESI\_SMPCache.cpp source to understand how the MESI protocol is currently modeled.
Discuss changes that need to be made to the readLine, writeLine, readRemoteAction, and writeRemoteAction functions. Show the necessary changes in pseudocode. Should the code that tracks cache access statistics be modified? If so, how?}

### 2014-10-11 lab/report/include/introduction.tex
added intro

\section{Introduction}
\label{sec:int}
\todo[inline]{Introduction, one paragraph}
These laboratory assignments aim to increase our knowledge of different cache coherence protocols. The goal is to simulate cache coherence protocols on a modeled cache coherent multi-core system using diverse benchmarks. The system model have a shared Last Level Cache (LLC), and each processor have a private L1 cache. Snooping based write-invalidate cache coherence protocols are used by the private L1 caches to ensure coherence, and a simple bus stand for the communication between LLC and private caches. The system is modeled by the Pin tool MultiCacheSim and Pin.

### 2014-10-11 lab/report/include/lab2.tex
listing

@@
 \label{sec:lab2}
+OpenMP (Open \textbf Multi-\textbf Processing)
 \todo[inline]{Introduction, one paragraph}
@@
 
+The program reads a matrix on the form $n \times n, n \in [4,8,25,100,200]$ containing data points from a file.
+To make the program simpler and avoid special cases a padding is added to the matrix changing its dimensions to $(n+2) \times (n+2)$.
+Then every other data point, here by called red dots, is looped through.
+For every red dot an average of itself and its neighbors to the west, south, north, and east is calculated and written as the dots new value.
+For all the other data points, called black dots, the same thing is performed.
+The difference of the dots new and old value is added together and the process is repeated until this accumulated difference of averages is smaller than 0.01.
+After the program finishes its execution the matrix have converged to an average of all red and black dots.
+
 \subsection{Task 2}
 \todo[inline]{Explain our parallel version}
+In listing~\ref{lst:solve} a parallelized version of the function \texttt{void Solve(double **A)} are shown.
+The parallelization have been done using directives from the OpenMP API.
+Only this function is shown as the other parts of the program for reading the input file and printing to the console is not included in the performance analysis.
+The approach taken for making the function execute in parallel is to distribute all the lines of the matrix evenly between all threads.
+
+\lstinputlisting[caption={\textit{Parallel version of original} \texttt{void Solve(double **A)}}, label={lst:solve}]{include/solve.c}
 
@@
 Discuss changes that need to be made to the readLine, writeLine, readRemoteAction, and writeRemoteAction functions.
-Show the necessary changes in pseudocode.
+Show the necessary changes in pseducode.
 Should the code that tracks cache access statistics be modified?

### 2014-10-11 lab/report/main.tex
hrj
added intro
listing

@@
-\documentclass[a4paper]{report}
+\documentclass[a4paper]{article}
 
 \include{packages}
-
+\usepackage{color}
 %MACROS---
@@
 
+\title{Report on Lab 1 \& Lab 2 in EDA282, Parallel Computer Organization and Design}
+\date{October 2014}
+\author{Dan Larsson, Jonas Hemlin
+\\Master Program of Embedded Electronic Systems Design\\
+Chalmers University of Technology, Gothenburg, Sweden\\
+Email: \{larsdan, jhemlin\}@student.chalmers.se}
+
+\definecolor{mygreen}{rgb}{0,0.6,0}
+\definecolor{mygray}{rgb}{0.5,0.5,0.5}
+\definecolor{mymauve}{rgb}{0.58,0,0.82}
+
+\lstset{ 
+	language=C,
+	backgroundcolor=\color{white},
+	basicstyle=\tiny,        % the size of the fonts that are used for the code
+	breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
+	breaklines=true,                 % sets automatic line breaking
+	captionpos=t,                    % sets the caption-position to bottom
+	commentstyle=\color{mygreen},    % comment style
+	deletekeywords={},
+	morekeywords={pragma,schedule,reduction,omp,single,private,shared, parallel},            
+	extendedchars=true,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
+	frame=single,                    % adds a frame around the code
+	keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
+	keywordstyle=\color{blue},       % keyword style
+	numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
+	numbersep=5pt,                   % how far the line-numbers are from the code
+	numberstyle=\tiny\color{mygray}, % the style that is used for the line-numbers
+	rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
+	showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
+	showstringspaces=false,          % underline spaces within strings only
+	showtabs=false,                  % show tabs within strings adding particular underscores
+	stepnumber=2,                    % the step between two line-numbers. If it's 1, each line will be numbered
+	stringstyle=\color{mymauve},     % string literal style
+	tabsize=2                       % sets default tabsize to 2 spaces                   % show the filename of files included with \lstinputlisting; also try caption instead of title
+}
+
 \begin{document}
-\section{test}
-heeeej!!! :)
+\maketitle
+
+\input{include/lab1}
+\input{include/lab2}
+\input{include/discussion}
 \end{document}

### 2014-10-11 lab/kernel/parallel.c

@@ -41,45 +41,32 @@ int main(int argc, char **argv)
 }
 
 
-void Solve(double **A)
-{
+void Solve(double **A) {
     int i, j, done = 0;
-    double temp;
-	double diff;
-
+    double temp, diff;
     #pragma omp parallel default(shared) private(i, j, temp)    
     while(!done){
-
    	 	#pragma omp single
 		diff = 0;
-
-		/* sweep that updates red elements */
 		#pragma omp for  schedule(static) reduction(+:diff)
-		for(i=1; i<=n; ++i){
-		    for(j=i%2 ? 1:2; j<=n; j=j+2){
+		for(i=1; i<=n; ++i) {
+		    for(j=i%2 ? 1:2; j<=n; j=j+2) {
 				temp = A[i][j];
 				A[i][j] = 0.2 * (A[i][j] + A[i][j-1] + A[i-1][j] + A[i][j+1] + A[i+1][j]);
 				diff += abs(A[i][j] - temp);
 		    }
 		}
-
-		/* sweep that updates black elements */
 		#pragma omp for schedule(static) reduction(+:diff)
-		for(i=1; i<=n; ++i){
-		    for(j=i%2 ? 2:1; j<=n; j=j+2){
+		for(i=1; i<=n; ++i) {
+		    for(j=i%2 ? 2:1; j<=n; j=j+2) {
 				temp = A[i][j];
 				A[i][j] = 0.2 * (A[i][j] + A[i][j-1] + A[i-1][j] + A[i][j+1] + A[i+1][j]);
 				diff += abs(A[i][j] - temp);
 		    }
 		}
-
-		#pragma omp single
-		{
+		#pragma omp single {
 			if(diff / (n*n) < 0.01)
 			    done = 1;
-
-			//printf("%.4lf\n", diff);
-			//printf("%lu\n", sizeof(double *));
     	}
     }
 }

### 2014-10-11 lab/report/include/solve.c

@@ -1,0 +1,29 @@
+void Solve(double **A) {
+    int i, j, done = 0;
+    double temp, diff;
+    #pragma omp parallel default(shared) private(i, j, temp)    
+    while(!done){
+   	 	#pragma omp single
+			diff = 0;
+			#pragma omp for schedule(static) reduction(+:diff)
+			for(i=1; i<=n; ++i) {
+			    for(j=i%2 ? 1:2; j<=n; j=j+2) {
+					temp = A[i][j];
+					A[i][j] = 0.2 * (A[i][j] + A[i][j-1] + A[i-1][j] + A[i][j+1] + A[i+1][j]);
+					diff += abs(A[i][j] - temp);
+			    }
+			}
+			#pragma omp for schedule(static) reduction(+:diff)
+			for(i=1; i<=n; ++i) {
+			    for(j=i%2 ? 2:1; j<=n; j=j+2) {
+					temp = A[i][j];
+					A[i][j] = 0.2 * (A[i][j] + A[i][j-1] + A[i-1][j] + A[i][j+1] + A[i+1][j]);
+					diff += abs(A[i][j] - temp);
+			    }
+			}
+			#pragma omp single {
+				if(diff / (n*n) < 0.01)
+				    done = 1;
+	    	}
+    }
+}

### 2014-10-11 lab/report/include/introduction.tex

@@
 \label{sec:int}
-\todo[inline]{Introduction, one paragraph}
 These laboratory assignments aim to increase our knowledge of different cache coherence protocols.
 The goal is to simulate cache coherence protocols on a modeled cache coherent multi-core system using diverse benchmarks.
+Three cache coherence protocols are targeted; MSI, MESI, and MESI-MG, where the last protocol is an modified version of MESI optimized for migratory sharing.
 The system model have a shared Last Level Cache (LLC), and each processor have a private L1 cache.

### 2014-10-11 lab/report/include/lab2.tex
WriteWrite

@@
 
+\todo[inline]{Comment on the prominent communication patterns in the benchmark.}
+
 \subsection{Task 2}
@@
 Only this function is shown as the other parts of the program for reading the input file and printing to the console is not included in the performance analysis.
-The approach taken for making the function execute in parallel is to distribute all the lines of the matrix evenly between all threads.
+The approach taken for making the function execute in parallel is to distribute the lines of the matrix evenly between all threads.
+This is done twice, once for the red dots and once for the black.
+Only one thread resets \texttt{diff} to zero on line 5 and test whether the program is finished or not on line 25.
+At each for loop that goes through the red and black dots the directive \texttt{reduction(+:diff)} is used which causes each thread to have its own private version of \texttt{diff} that all gets added together after the for loop.
 
-\lstinputlisting[caption={\textit{Parallel version of original} \texttt{void Solve(double **A)}}, label={lst:solve}]{include/solve.c}
+\lstinputlisting[caption={\textit{Parallel version of original} \texttt{void Solve(double **A).}}, label={lst:solve}]{include/solve.c}
 

### 2014-10-11 lab/report/include/introduction.tex

@@
 These laboratory assignments aim to increase our knowledge of different cache coherence protocols.
-The goal is to simulate cache coherence protocols on a modeled cache coherent multi-core system using diverse benchmarks.
+The goal is to simulate cache coherence protocols (CCP) on a modeled cache coherent multi-core system using diverse benchmarks.
 Three cache coherence protocols are targeted; MSI, MESI, and MESI-MG, where the last protocol is an modified version of MESI optimized for migratory sharing.
@@
 The system is modeled by the Pin tool MultiCacheSim and Pin.
+In listing~\ref{lst:setup} is the command used for simulating a CCP on the modeled system with a chosen benchmark.
+
+\begin{lstlisting}[label=lst:setup, basicstyle=\scriptsize, caption={Command used for simulation.}]
+#> ./pin -mt -t ./source/tools/MultiCacheSim/obj-intel64/mcs.so -csize 65536 -bsize 64 -assoc 8 -llc -llcsize 4194304 -llcbsize 64 -llcassoc 16 -numcaches 8 -protos ./source/tools/MultiCacheSim/obj-intel64/mcs_<CCP>.so -- <benchmark> [#threads,..]
+\end{lstlisting}

### 2014-10-11 lab/report/include/lab1.tex
added intro

@@
 \label{sec:lab1}
-\todo[inline]{Introduction, one paragraph}
+\todo[inline]{little drafty}
+In this laboratory assignment, we evaluate overheads in cache coherence protocols using microbenchmarks.
+Three diverse benchmarks, described in \ref{sec:lab11}, are used to compare the coherence protocols.
+In \ref{sec:lab12}, MSI and MESI is compared and in \ref{sec:lab13}, MESI and MESI-MG is compared.
+
+\todo[inline]{Add working size set}
 
 \subsection{Task 1}
-\todo[inline]{Explain the benchmarks}
+\label{sec:lab11}
+Three different microbenchmarks are used with various prominent communication patterns. Here follows a description of each benchmark.
+\subsubsection*{Benchmark 1 - Producer consumer}
+In this benchmark, one thread increments the values in a memory array, whereas the other threads read these values and accumulate the read values in a private variable.
+However, it is not ensured that the incrementing thread increments the values before the other threads reads, and thus can this happen in any order.
+The memory accesses is done in mutual exclusion using a lock, and before a new lap of increments \& reads there is a barrier.
+
+\subsubsection*{Benchmark 2 - No sharing}
+The memory array is divided up equally among the threads and a thread only increments the values in it's assigned array elements, thus is the memory array not shared.
+A barrier is placed before a new incrementation lap of the assigned array elements begins.
+
+\subsubsection*{Benchmark 3 - All shared}
+All threads increments every element in the memory array.
 
 \subsection{Task 2}
+\label{sec:lab12}
 \todo[inline]{
@@
 \subsection{Task 3}
+\label{sec:lab13}
 \todo[inline]{

### 2014-10-12 lab/report/main.tex
changed margins

@@
 
+\usepackage[margin=2cm]{geometry}
 \include{packages}
@@
 \newcommand{\reffig}[1]{Fig. \ref{#1}}
+\newcommand\xput[2][0.5]{%
+    \rule{#1\linewidth}{0pt}\makebox[0pt][c]{#2}\hfill}
 
@@
 	backgroundcolor=\color{white},
-	basicstyle=\tiny,        % the size of the fonts that are used for the code
+	basicstyle=\scriptsize,        % the size of the fonts that are used for the code
 	breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
@@
 
+\input{include/introduction}
 \input{include/lab1}

### 2014-10-12 lab/report/include/lab2.tex
changed margins
write write

@@
 \label{sec:lab2}
-OpenMP (Open \textbf Multi-\textbf Processing)
-\todo[inline]{Introduction, one paragraph}
+For the second part of the lab the task is to parallelize a serial implementation of the Red-Back Gauss-Seidal algorithm.
+This is done using OpenMP (Open \textbf Multi-\textbf Processing) which is an API for forming multi-threaded programs in C, C++ and, Fortran.
+This parallel version is evaluated using the same simulator and CCP:s as in \nameref{sec:lab1}.
 
 \subsection{Task 1}
-\todo[inline]{Analyze the source code of the benchmark and provide a detailed description of how it works. Comment on the prominent communication patterns in the benchmark.}
-
 The program reads a matrix on the form $n \times n, n \in [4,8,25,100,200]$ containing data points from a file.
@@
 For all the other data points, called black dots, the same thing is performed.
-The difference of the dots new and old value is added together and the process is repeated until this accumulated difference of averages is smaller than 0.01.
+The difference of the dots new and old value is added together and the process is repeated until this accumulated difference of averages, divided by $n^2$, is smaller than $0.01$.
 After the program finishes its execution the matrix have converged to an average of all red and black dots.
 
-\todo[inline]{Comment on the prominent communication patterns in the benchmark.}
+The distribution of the red and black dots forms a chess pattern.
+This means that when we calculate the average for a dot and its neighbors, all the neighbors are of the other color.
+As such, when going through for example the red dots there is no need to invalidate any other dots because the black dots are only read and you are the only one accessing your red dot.
 
@@
 
+\centerline{
+\begin{minipage}{1.0\textwidth}
 \lstinputlisting[caption={\textit{Parallel version of original} \texttt{void Solve(double **A).}}, label={lst:solve}]{include/solve.c}
-
+\end{minipage}
+}
 \subsection{Task 3}

### 2014-10-12 lab/report/include/lab1.tex
wrote

--- 2014-10-11 lab/report/include/lab1.tex
@@
 \label{sec:lab1}
-\todo[inline]{Introduction, one paragraph}
+In this laboratory assignment, we evaluate overheads in cache coherence protocols using microbenchmarks.
+Three diverse benchmarks, described in section \ref{sec:lab11}, are used to compare the coherence protocols.
+In section \ref{sec:lab12}, MSI and MESI is compared and in section \ref{sec:lab13}, MESI and MESI-MG is compared.
+In each comparison, we use three working set: the first is smaller than an private L1 cache (512 counters), the second is larger than an private L1 cache and smaller than the LLC (1536 counters), and the last is larger than the LLC (98304 counters).
+The counters are stored in an array, the counter array, which is accessed by the benchmark.
 
 \subsection{Task 1}
-\todo[inline]{Explain the benchmarks}
+\label{sec:lab11}
+Three different microbenchmarks with disgustingly various communication patterns are used in the simulations and here follow a description of each benchmark.
+\subsubsection*{Benchmark 1 - Producer consumer}
+In this benchmark, one thread increments the values in the counter array, whereas the other threads read these values and accumulate the read values in a private variable.
+However, it is not ensured that the incrementing thread increments the values before the other threads reads, and thus can this happen in any order.
+The counter accesses is done in mutual exclusion using a lock, and before a new lap of increments \& reads there is a barrier.
+
+\subsubsection*{Benchmark 2 - No sharing}
+This benchmark have no sharing of the counter array.
+The counter array is divided up equally among the threads and a thread only increments the values in it's assigned array elements.
+A barrier is placed before a new incrementation lap of the assigned array elements begins.
+
+\subsubsection*{Benchmark 3 - All shared}
+In this benchmark, the counter array is shared among all threads.
+Each thread increments every element in the counter array while in mutual exclusion.
+A lock is used to provide mutual exclusion and a barrier is placed after all threads have incremented.
 
 \subsection{Task 2}
-\todo[inline]{
-Compare MSI and MESI.
-When is having the E state beneficial (over baseline MSI)?
-How does the behavior change when the dataset size changes?}
+\label{sec:lab12}
+Here the CCPs MSI and MESI are compared using benchmarks 1 and 2.
+The difference between the CCPs is that MESI includes an exclusive (E) state, which is beneficial if data is read exclusively by one processor.
+If the data is then read by another processor, a move to the shared (S) state is made without cost.
+I.e.
+MESI can only produce equal or better results than MSI.
+
+In the producer consumer benchmark, the reading threads transition the data element to the E state when accessing the array for the first time if the writing thread have not accessed it already.
+However, the whole counter array is shared and thus will not the E state be of any gain.
+Therefore, MSI and MESI performs equally regardless of working set size, which is shown in \reffig{fig:resultslab1}.
+
+\begin{figure}[t]
+	\center
+	\includegraphics[width=0.7\textwidth]{bars}
+	\caption{Simulation results using various cache coherent protocols combined with different working sets and benchmarks.}
+	\label{fig:resultslab1}
+\end{figure}
 
 \subsection{Task 3}
+\label{sec:lab13}
 \todo[inline]{

### 2014-10-12 lab/report/include/lab2.tex
formulA

@@
 \subsection{Task 2}
-\todo[inline]{Explain our parallel version}
 In listing~\ref{lst:solve} a parallelized version of the function \texttt{void Solve(double **A)} are shown.
@@
 
+Samma mönster som i Lab 1, MSI får mer writemiss osv än MESI vilket gör MESI lite bättre.
+\begin{figure}[t]
+	\center
+	\includegraphics[width=0.8\textwidth]{lab2bars}
+	\caption{Results from simulation of the parallelized benchmark for all cache coherence protocols combined with different working sets.}
+	\label{fig:resultslab1}
+\end{figure}
+
 \subsection{Task 4}

### 2014-10-12 lab/report/include/lab2.tex
Appendices added, or something.

@@
 
-\centerline{
-\begin{minipage}{1.0\textwidth}
-\lstinputlisting[caption={\textit{Parallel version of original} \texttt{void Solve(double **A).}}, label={lst:solve}]{include/solve.c}
-\end{minipage}
-}
 \subsection{Task 3}
@@
 
-Samma mönster som i Lab 1, MSI får mer writemiss osv än MESI vilket gör MESI lite bättre.
+In \reffig{fig:lab2} the results from running the parallelized version of the program in the simulator with the different CCP:s and three different matrix dimensions.
+
+DRAFTY:
+MESI gains on MSI with big working sets as the working set does not fit in the cache and in MSI every readMiss will fetch the line into Shared state and then move to modified causing a invalidate to be sent.
+In MESI this is skipped as the readMiss will fetch the line into Exclusive state and then Modified not causing any invalidates to be sent.
+
 \begin{figure}[t]
@@
 	\caption{Results from simulation of the parallelized benchmark for all cache coherence protocols combined with different working sets.}
-	\label{fig:resultslab1}
+	\label{fig:lab2}
 \end{figure}

### 2014-10-12 lab/report/include/lab2.tex
wirte stuff

@@
 
-In \reffig{fig:lab2} the results from running the parallelized version of the program in the simulator with the different CCP:s and three different matrix dimensions.
+In \reffig{fig:lab2} results from running the parallelized version of the program in the simulator with different CCP:s and different matrix dimensions are shown.
+It can be seen that using MSI or MESI does not cause any major change in performance but there is a small speed improvement with MESI over MSI at the bigger matrix sizes.
+This can be explained by the fact that it is cheaper to go to M state from E state than from S state as this would cause invalidations to be sent on the buss.
+
+Changing to MESI+MG yields a \~32\% performance boost at the smallest matrix size and \~14\% at the two bigger sizes.
+These improvements is due to MESI+MG not using the S state at all which the other two CCP:s does.
+With the smallest matrix size, which fits entirely in the L1 cache,  \todo{WRITE SMART STUFF HERE!!}
 

### 2014-10-12 lab/report/main.tex
formulA
Appendices added, or something.
wirte stuff

@@
 \include{packages}
+\usepackage[toc,page]{appendix}
 \usepackage{color}
@@
 \definecolor{mygreen}{rgb}{0,0.6,0}
-\definecolor{mygray}{rgb}{0.5,0.5,0.5}
+\definecolor{mygray}{rgb}{0.1,0.1,0.1}
 \definecolor{mymauve}{rgb}{0.58,0,0.82}
@@
 	backgroundcolor=\color{white},
-	basicstyle=\scriptsize,        % the size of the fonts that are used for the code
+	basicstyle=\small,        % the size of the fonts that are used for the code
 	breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
@@
 	numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
-	numbersep=5pt,                   % how far the line-numbers are from the code
-	numberstyle=\tiny\color{mygray}, % the style that is used for the line-numbers
+	numbersep=8pt,                   % how far the line-numbers are from the code
+	numberstyle=\footnotesize\color{mygray}, % the style that is used for the line-numbers
 	rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
@@
 \maketitle
-
+$$ Cost = 1 * (WriteHit+ReadHit) + 30 * (readMissesServicedByShared + invalidationsSent)$$ $$+ 20 * LLCHit + 50 * LLCMiss $$
 \input{include/introduction}
@@
 \input{include/discussion}
+\newpage
+\begin{appendices}
+\section{Parallelized program for Lab 2}
+\centerline{
+\begin{minipage}{1.0\textwidth}
+\lstinputlisting[caption={\textit{Parallel version of original} \texttt{void Solve(double **A).}}, label={lst:solve}]{include/solve.c}
+\end{minipage}
+}
+\end{appendices}
 \end{document}

### 2014-10-12 lab/report/include/lab1.tex
written some text

@@
 This benchmark have no sharing of the counter array.
-The counter array is divided up equally among the threads and a thread only increments the values in it's assigned array elements.
+The counter array is divided up equally among the threads, giving each thread access to every nth counter element, where n is the number of threads.
+Each thread then increments it's assigned counters sequentially.
 A barrier is placed before a new incrementation lap of the assigned array elements begins.
@@
 In this benchmark, the counter array is shared among all threads.
-Each thread increments every element in the counter array while in mutual exclusion.
+Each thread increments every element in the counter array sequentially while in mutual exclusion.
 A lock is used to provide mutual exclusion and a barrier is placed after all threads have incremented.
@@
 In the producer consumer benchmark, the reading threads transition the data element to the E state when accessing the array for the first time if the writing thread have not accessed it already.
-However, the whole counter array is shared and thus will not the E state be of any gain.
+However, the whole counter array is shared and thus will not state E be of any gain.
 Therefore, MSI and MESI performs equally regardless of working set size, which is shown in \reffig{fig:resultslab1}.
+
+In the no sharing benchmark the first read access to every element end in state S for MSI and state E for MESI, respectively.
+The following write transitions the element to state M for both the CCPs, but MSI have to issue a BusUpdate whereas MESI's transition is costless.
+The BusUpdate broadcasts an invalidate signal which gives MSI a slight reduction in performance compared to MESI.
+In the small working set the elements will stay in state M to the end of execution, but in the larger working sets the element is fetched from LLC upon each read access due to cache trashing.
+This makes MSI issue an invalidate broadcast for each incrementation and the accumulated overhead is easily spotted in \reffig{fig:resultslab1}, it is also seen that the overhead become linearly to the size of the working set.
 
@@
 	\includegraphics[width=0.7\textwidth]{bars}
-	\caption{Simulation results using various cache coherent protocols combined with different working sets and benchmarks.}
+	\caption{Simulation results using various cache coherence protocols combined with different working sets and benchmarks.}
 	\label{fig:resultslab1}

### 2014-10-14 lab/report/include/lab1.tex
wrote

@@
 How does the behavior change when the dataset size changes?}
+MESI-MG is a modified version of MESI where the S state is not utilized.
+If a cache reads a data element it will be forwarded from a remote private cache if it holds the data element in the E or M state, and then the remote cache invalidates it's copy.

### 2014-10-14 excersice 1/main.tex
bläbläblääblä

\documentclass[a4paper]{article}

\renewcommand{\labelenumi}{(\alph{enumi})}

\usepackage{listings}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{fancyhdr}
\pagestyle{fancy}

\lhead{Chalmers - EDA282}
\chead{Excercise 1}
\rhead{Dan Larsson}
\lfoot{November 2014}

\begin{document}

\section*{Q1}
\begin{enumerate}
\item % a
The goal of the algorithm in Listing 1 is to group n points into m clusters. This is done by invoking the function \texttt{kernel} until its return value \texttt{delta} is equal to 0. In \texttt{kernel} all points are itterated over to find which one of the m clusters supplied in \texttt{clusters} is closest. To find the closest cluster all clusters are itterated over in \texttt{find\_nearest\_point} and the Euclidean distance is calulated in \texttt{euclid\_dist\_2}. When all clusters have been itterated over the index of the cloest one is returned to \texttt{kernel}. This index is used to check if the point already belongs to this cluster, if not delta is increased to indicate that a point have changed cluster. To calulate the center of each cluster an average of the cooridnates for all points belonging to the cluster are calulated.

\item % b
\label{itm:b}
One way to parallelize the \texttt{kernel} would be to simply divide all points into p parts, where p is the amount of processors. For this to work the variables \texttt{new\_centers} and \texttt{new\_centers\_len} would need to be syncronized. This can be done using the \texttt{parallel for} and \texttt{critical} directive of \texttt{OpenMP}.

\item % c
\label{itm:c}
As mentioned in (b), there would be a need for synchronization of the variables \texttt{new\_centers} and \texttt{new\_centers\_len}. For the other global variables, \texttt{clusters} and \texttt{feature} are only read and \texttt{membership} is only accessed based on the index of each point and as the different processors have no overlapping points there will accesses to the same addresses. In listing~\ref{lst:kernelcoarse} a modified kernel with \texttt{OpenMP} directives for a coarse grained aproach is shown.

\item % d
A more fine grained solution than the one presented in (c) is shown in listing~\ref{lst:kernelfine} where the \texttt{atomic} directive is used on each of the critical varaibles. As explained in (c) only \texttt{new\_centers} and \texttt{new\_centers\_len} need this protection so two locks is enough.

\end{enumerate}
\lstset{language=C,frame=single}
\lstinputlisting[lastline=31, caption={Modified kernel coarse grained},label={lst:kernelcoarse}]{upgC.c}
\lstinputlisting[firstline=32, caption={Modified kernel fine grained},label={lst:kernelfine}]{upgC.c}

\section*{Q2}
The request write/read sequence will be as follows: \\

\texttt{W0~W1~R0~R1~W2~W3~R2~R3~W4~W5~R4~R5~W6~W7~R6~R7} \\ \\
and the values that the 8 processors will read will be the following:\\

\begin{tabular}{ l | c c c c c c c c }

\textbf{Processor} & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 \\
\textbf{Read}     & 1 & 1 & 3 & 3 & 5 & 5 & 7 & 7 \\

\end{tabular}

\section*{Q3}
\begin{enumerate}
\item % a 
\label{itm:3a}
The amount of cycles will be the following for MSI:
\texttt{
  \\Proc1: LvL2Rd + (LvL1RdX + BusUP + Wr) + Wr
  \\Proc2: LvL2Rd + (LvL1RdX + BusUP + Wr) + Wr
  \\Proc3: LvL2Rd + (LvL1RdX + BusUP + Wr) + Wr
  \\Proc4: LvL2Rd + (LvL1RdX + BusUP + Wr) + Wr 
}
\[4\times(40+(20+10+1)+1) = 288\]
and for MESI:
\texttt{
  \\Proc1: LvL2Rd + Wr + Wr
  \\Proc2: LvL2Rd + (LvL1RdX + BusUP + Wr) + Wr
  \\Proc3: LvL2Rd + (LvL1RdX + BusUP + Wr) + Wr
  \\Proc4: LvL2Rd + (LvL1RdX + BusUP + Wr) + Wr 
}
\[(40+1+1)+3\times(40+(20+10+1)+1) = 258\]
where the difference is due to the first processor going to the E state instead of the S state and thereby saves 30 cycles when it then wants to write. There are no Snoops done as we only care about the cycles wasted in the current cache.

\item % b 
The amount of trafic that gets generated for MSI based on the sequences in (a)):
\[4\times(38B+38B+10B) = 344B\]
and for MESI:
\[38B+3\times(38B+38B+10B) = 296B\]
\end{enumerate}

\section*{Q4} % 7.2
\begin{enumerate}
\item % a
There is no need to protect the read as if the read returns the wrong value, it will be a lower value than it should and the read can simply be done again.
\item % b 
See listing~\ref{lst:q3b}.

\item % c
See listing~\ref{lst:q3c}.
\end{enumerate}

\lstset{language={[x86masm]Assembler},frame=single}
\lstinputlisting[firstline=20, lastline=33, caption={Sync. using T\&S},label={lst:q3b}]{upgASM.asm}
\lstinputlisting[firstline=37, lastline=43, caption={Sync. using F\&A},label={lst:q3c}]{upgASM.asm}

\section*{Q5} % 7.3
\begin{enumerate}
\item % a
See listing~\ref{lst:q4a}.
\setcounter{enumi}{3}
\item % d
See listing~\ref{lst:q4e}.
\item % e
See listing~\ref{lst:q4d}.
\end{enumerate}

\lstset{language={[x86masm]Assembler},frame=single}
\lstinputlisting[firstline=0, lastline=4, caption={F\&ADD X, Rx, a},label={lst:q4a}]{upgASM.asm}

\lstinputlisting[firstline=6, lastline=10, caption={SWAP(Rx, X)},label={lst:q4e}]{upgASM.asm}

\lstinputlisting[firstline=12, lastline=18, caption={CAS Rx, Ry, X},label={lst:q4d}]{upgASM.asm}

\end{document}

### 2014-10-14 excersice 1/upgC.c

@@ -1,0 +1,63 @@
+
+int kernel(int npoints, int nclusters) {
+
+  int i, j, delta = 0, index;
+  
+  #pragma omp parallel for default(shared) \
+                            private(i,j,index) \
+                            schedule(static) \
+                            reduction(+:delta)
+
+  for(i = 0; i < npoints; i++) {
+    
+    index = find_nearest_point(feature[i], 2, clusters, nclusters);
+    
+    if (membership[i] != index) {
+      delta += 1;
+    }
+    
+    membership[i] = index;
+    
+    #pragma omp critical
+    { 
+      new_centers_len[index]++;
+      for(j = 0; j < 2; j++) {
+        new_centers[index][j] += feature[i][j];
+      }
+    }
+
+  }
+  return delta;
+}
+
+int kernel(int npoints, int nclusters) {
+  
+  int i, j, delta = 0, index;
+  
+  #pragma omp parallel for default(shared) \
+                            private(i,j,index) \
+                            schedule(static) \
+                            reduction(+:delta)
+  
+  for(i = 0; i < npoints; i++) {
+    
+    index = find_nearest_point(feature[i], 2, clusters, nclusters);
+    
+    if (membership[i] != index) {
+      delta += 1;
+    }
+    
+    membership[i] = index;
+    
+    #pragma omp atomic
+    new_centers_len[index]++;
+
+    for(j = 0; j < 2; j++) {
+      #pragma omp atomic
+      new_centers[index][j] += feature[i][j];
+    }
+
+  }
+
+  return delta;
+}

### 2014-10-14 lab/report/include/lab2.tex
write

@@
 In \reffig{fig:lab2} results from running the parallelized version of the program in the simulator with different CCP:s and different matrix dimensions are shown.
-It can be seen that using MSI or MESI does not cause any major change in performance but there is a small speed improvement with MESI over MSI at the bigger matrix sizes.
-This can be explained by the fact that it is cheaper to go to M state from E state than from S state as this would cause invalidations to be sent on the buss.
+It can be seen that using MESI over MSI does not cause any noticeable boost in performance.
+This is because the gain of the extra E state becomes of no importance as the cache lines is in the M state for most of the time.
+Some benefit of the E state can though be seen in table~\ref{tbl:lab2} as the amount of write misses and invalidations sent are a bit lower for MESI than MSI.
 

### 2014-10-14 lab/report/include/lab2.tex
task 3 isch finished.

--- 2014-10-12 lab/report/include/lab2.tex
@@
 For the second part of the lab the task is to parallelize a serial implementation of the Red-Back Gauss-Seidal algorithm.
-This is done using OpenMP (Open \textbf Multi-\textbf Processing) which is an API for forming multi-threaded programs in C, C++ and, Fortran.
+This is done using OpenMP (Open \textbf Multi-\textbf Processing) which is an API for forming multi-threaded programs in C, C++, and Fortran.
 This parallel version is evaluated using the same simulator and CCP:s as in \nameref{sec:lab1}.
@@
 \subsection{Task 1}
-The program reads a matrix on the form $n \times n, n \in [4,8,25,100,200]$ containing data points from a file.
+The program reads a matrix on the form $n \times n$ containing data points from a file.
 To make the program simpler and avoid special cases a padding is added to the matrix changing its dimensions to $(n+2) \times (n+2)$.
@@
 \subsection{Task 2}
+\label{subsec:lab2:task2}
 In listing~\ref{lst:solve} a parallelized version of the function \texttt{void Solve(double **A)} are shown.
@@
 \subsection{Task 3}
-\todo[inline]{
-Evaluate MSI, MESI, and MESI+migratory in the context of this benchmark.
-Perform a detailed characterization of the benchmark using different protocols.
-Discuss how the behavior changes when the working set size is modified.
-Show how execution time relates to the cache access statistics.}
+To be able to evaluate the program described in \nameref{subsec:lab2:task2} we decide to mimic the working set sizes in \nameref{sec:lab1}.
+However, the provided data files does not include a matrix big enough to fill neither each threads L1 cache or the LLC cache.
 
-In \reffig{fig:lab2} the results from running the parallelized version of the program in the simulator with the different CCP:s and three different matrix dimensions.
+In \reffig{fig:lab2} results from running the parallelized version of the program in the simulator with different CCP:s and different matrix dimensions are shown.
+It can be seen that using MESI over MSI does not cause any noticeable boost in performance.
+This is because the gain of the extra E state becomes of no importance as the cache lines is in the M state most of the time.
+Some benefit of the E state can though be seen in Table~\ref{tbl:lab2} as the amount of \textit{WriteMisses} and \textit{InvalidationsSent} are a bit lower for MESI than MSI and the gap becomes bigger as the matrix size increases.
+This does not show in \reffig{fig:lab2} as the cycle count are not dominated by \textit{WriteMisses} and \textit{InvalidationsSent}.
 
-DRAFTY:
-MESI gains on MSI with big working sets as the working set does not fit in the cache and in MSI every readMiss will fetch the line into Shared state and then move to modified causing a invalidate to be sent.
-In MESI this is skipped as the readMiss will fetch the line into Exclusive state and then Modified not causing any invalidates to be sent.
+Changing to MESI-MG yields a huge performance degradation of 420\%, 394\%, and 430\% for matrix size 25, 100, and 200 respectively.
+This comes from the sharing performed of the brim lines between threads.
+For each read of a dot in any of these lines, the dot gets migrated over to the thread performing the read.
+This frequent migration of dots generates a large amount of \textit{ReadMissesSerivcedByModfied} as can be seen in Table~\ref{tbl:lab2}.
+The overhead stays fairly constant compared to the other CCP:s when the matrix size increases.
+This is duo to the proportion of sharing between the threads stays the same.
 

### 2014-10-14 lab/kernel/parallel.c

@@ -64,7 +64,8 @@ void Solve(double **A) {
 				diff += abs(A[i][j] - temp);
 		    }
 		}
-		#pragma omp single {
+		#pragma omp single 
+        {
 			if(diff / (n*n) < 0.01)
 			    done = 1;
     	}

### 2014-10-14 lab/report/include/lab1.tex
new figures
wrote and added appendix

--- 2014-10-12 lab/report/include/lab1.tex
@@
 Each thread increments every element in the counter array sequentially while in mutual exclusion.
-A lock is used to provide mutual exclusion and a barrier is placed after all threads have incremented.
+A lock is used to provide mutual exclusion and a barrier is placed before a new incrementation lap.
 
@@
 However, the whole counter array is shared and thus will not state E be of any gain.
-Therefore, MSI and MESI performs equally regardless of working set size, which is shown in \reffig{fig:resultslab1}.
+Therefore, MSI and MESI performs equally regardless of working set size, which is shown in \reffig{fig:lab1bars}.
 
@@
 The BusUpdate broadcasts an invalidate signal which gives MSI a slight reduction in performance compared to MESI.
-In the small working set the elements will stay in state M to the end of execution, but in the larger working sets the element is fetched from LLC upon each read access due to cache trashing.
-This makes MSI issue an invalidate broadcast for each incrementation and the accumulated overhead is easily spotted in \reffig{fig:resultslab1}, it is also seen that the overhead become linearly to the size of the working set.
+In the small working set the elements will stay in state M to the end of execution, but in working sets larger than a private L1 cache the element has to be fetched from higher levels.
+This makes MSI issue an invalidate broadcast for each increment and the accumulated overhead is easily spotted in \reffig{fig:lab1bars}.
 
@@
 	\center
-	\includegraphics[width=0.7\textwidth]{bars}
+	\includegraphics[width=0.7\textwidth]{lab1bars}
 	\caption{Simulation results using various cache coherence protocols combined with different working sets and benchmarks.}
-	\label{fig:resultslab1}
+	\label{fig:lab1bars}
 \end{figure}
@@
 \label{sec:lab13}
-\todo[inline]{
-Compare MESI and MESI+migratory.
-Why is one better than the other?
-How does the behavior change when the dataset size changes?}
+MESI is here compared with MESI-MG by executing benchmark 1 and 3.
+MESI-MG is a modified version of MESI where the S state is not utilized.
+If a cache issues a read request of a data element which exists in state E or M in a remote cache, the data is forwarded from the remote cache and then the remote cache invalidates it's copy.
+By this scheme, the S state is never used.
+
+The producer consumer benchmark benefit from the S state, and therefor produces MESI better results than MESI-MG.
+In the benchmark, one thread update an element whereas the other threads read it and by migrating the element from one private cache to the other is less optimal than letting the reading threads share the element and read it in parallel.
+This overhead will be a smaller part of the total cycles as the working set increases, as the misses increase in the private caches and also in the LLC, which is costly in terms of cycles.
+This trend is shown in \reffig{fig:lab1bars}.
+
+In the all shared benchmark, every thread increments every data element during mutual exclusion which favors MESI-MG.
+Each data element is read and modified exclusively by one processor at a time, thus is the S state not beneficial.
+As MESI-MG do not utilize the S state, as described, it is more suitable than MESI in this benchmark.
+In Appendix [???] it is shown that the CCPs results are a like, except that MESI have to issue a significant number of invalidates, which is needed for the transition from state S to M.
+As shown in \reffig{fig:lab1bars}, MESI-MG's benefit is not visible for the working set size increasing the size of the LLC.
+That is because the cycle count is dominated by the LLC misses, as in the producer consumer benchmark.

### 2014-10-14 lab/report/include/lab2.tex
write write

@@
 \subsection{Task 4}
-\todo[inline]{
-Outline a strategy to modify the MESI protocol to support the O state.
+%\todo[inline]{Outline a strategy to modify the MESI protocol to support the O state.
 Analyze the MESI\_SMPCache.cpp source to understand how the MESI protocol is currently modeled.
@@
 If so, how?}
+
+To add the Owner (O) state to the existing MESI CCP only some small changes are needed.
+In \textit{readLine()} and \textit{remoteWriteAction()} no changes are needed.
+Turning to \textit{readRemoteAction()}, transitions from the E, M, and O states to the O state are added, and all of these transitions provide data to the calling cache.
+The action taken when in the shared state is removed.
+For the \textit{writeLine()} the only change needed is to take the same actions when in the O state as in the S state.
+For the cache access statistics there is no need to change anything

### 2014-10-14 lab/report/include/lab2.tex
task4

@@
 This is done using OpenMP (Open \textbf Multi-\textbf Processing) which is an API for forming multi-threaded programs in C, C++, and Fortran.
-This parallel version is evaluated using the same simulator and CCP:s as in \nameref{sec:lab1}.
+This parallel version is evaluated using the same simulator and CCPs as in \nameref{sec:lab1}.
 
@@
 
-In \reffig{fig:lab2} results from running the parallelized version of the program in the simulator with different CCP:s and different matrix dimensions are shown.
+In \reffig{fig:lab2} results from running the parallelized version of the program in the simulator with different CCPs and different matrix dimensions are shown.
 It can be seen that using MESI over MSI does not cause any noticeable boost in performance.
@@
 This frequent migration of dots generates a large amount of \textit{ReadMissesSerivcedByModfied} as can be seen in Table~\ref{tbl:lab2}.
-The overhead stays fairly constant compared to the other CCP:s when the matrix size increases.
+The overhead stays fairly constant compared to the other CCPs when the matrix size increases.
 This is duo to the proportion of sharing between the threads stays the same.
@@
 For the cache access statistics there is no need to change anything
+
+\par\noindent
+\begin{minipage}[t]{.5\textwidth}
+\begin{lstlisting}[language=C,frame=lrtb]
+In readLine()
+	if(MODIFIED) {...}
+	if(EXCLUSIVE) {...}
+
+	if(SHARED)
+		return data
+
+In writeRemoteAction()
+	if(SHARED) {...}
+\end{lstlisting}%
+\end{minipage}%
+\hfill
+\begin{minipage}[t]{.5\textwidth}
+\begin{lstlisting}[frame=lrtb]
+In readLine()
+	if(MODIFIED || EXCLUSIVE || OWNER) 
+		state = OWNER
+		return data
+
+	if(SHARED)
+	{}
+
+In writeRemoteAction()
+	if(SHARED || OWNER) {...}
+\end{lstlisting}%
+\end{minipage}%

### 2014-10-14 lab/report/include/introduction.tex
written some text
new figures
wrote and added appendix
correcting

@@
 These laboratory assignments aim to increase our knowledge of different cache coherence protocols.
-The goal is to simulate cache coherence protocols (CCP) on a modeled cache coherent multi-core system using diverse benchmarks.
+To reach this goal, we simulate cache coherence protocols (CCP) on a modeled cache coherent multi-core system using diverse benchmarks.
 Three cache coherence protocols are targeted; MSI, MESI, and MESI-MG, where the last protocol is an modified version of MESI optimized for migratory sharing.
@@
 In listing~\ref{lst:setup} is the command used for simulating a CCP on the modeled system with a chosen benchmark.
+To relate the simulation results to excecution time, Eq.
+\ref{eq:cycles} is used.
 
@@
 \end{lstlisting}
+\begin{equation}
+	$$\label{eq:cycles}$$
+	Cycles = 1 * (WriteHit+ReadHit) + 30 * (readMissesServicedByModified + invalidationsSent)$$ $$+ 20 * LLCHit + 50 * LLCMiss
+\end{equation}

### 2014-10-14 lab/report/include/lab2.tex
correcting

--- 2014-10-14 lab/report/include/lab2.tex
@@
 This is done using OpenMP (Open \textbf Multi-\textbf Processing) which is an API for forming multi-threaded programs in C, C++, and Fortran.
-This parallel version is evaluated using the same simulator and CCP:s as in \nameref{sec:lab1}.
+This parallel version is evaluated using the same simulator and CCPs as in \nameref{sec:lab1}.
 
 \subsection{Task 1}
-The program reads a matrix on the form $n \times n$ containing data points from a file.
-To make the program simpler and avoid special cases a padding is added to the matrix changing its dimensions to $(n+2) \times (n+2)$.
+The program first reads a matrix on the form $n \times n$ containing data points from a file.
+To make the program simpler and avoid special cases padding is added changing it's dimensions to $(n+2) \times (n+2)$.
 Then every other data point, here by called red dots, is looped through.
-For every red dot an average of itself and its neighbors to the west, south, north, and east is calculated and written as the dots new value.
-For all the other data points, called black dots, the same thing is performed.
-The difference of the dots new and old value is added together and the process is repeated until this accumulated difference of averages, divided by $n^2$, is smaller than $0.01$.
-After the program finishes its execution the matrix have converged to an average of all red and black dots.
+For every red dot, an average of itself and it's neighbors to the west, south, north, and east is calculated and written as the dot's new value.
+The same procedure is performed on all the other data points, called black dots.
+The difference of the dots new and old values is added together and the process of calculating new averages is repeated until this accumulated difference of averages, divided by $n^2$, is smaller than $0.01$.
+When the program finishes the matrix have converged to an average of all red and black dots.
 
 The distribution of the red and black dots forms a chess pattern.
-This means that when we calculate the average for a dot and its neighbors, all the neighbors are of the other color.
-As such, when going through for example the red dots there is no need to invalidate any other dots because the black dots are only read and you are the only one accessing your red dot.
+This means that when the average for a dot and it's neighbors is calculated, all the neighbors are of the other color.
+As such, when calculating the average for example the red dots there is no need to invalidate any other dots (than red dots???) because the black dots are only read and you are the only one accessing your red dot.
 
@@
 \label{subsec:lab2:task2}
-In listing~\ref{lst:solve} a parallelized version of the function \texttt{void Solve(double **A)} are shown.
-The parallelization have been done using directives from the OpenMP API.
-Only this function is shown as the other parts of the program for reading the input file and printing to the console is not included in the performance analysis.
-The approach taken for making the function execute in parallel is to distribute the lines of the matrix evenly between all threads.
+In listing~\ref{lst:solve} a parallelized version of the function \texttt{void Solve(double **A)} is shown.
+The approach taken to parallilize the function is to distribute the rows of the matrix evenly between all threads.
 This is done twice, once for the red dots and once for the black.

### 2014-10-14 lab/report/main.tex
written some text
write
added stupid table. crapy shiet
task4
correcting

@@
 \usepackage{color}
+\usepackage{lscape}
+\usepackage{chngpage}
+\usepackage{float}
 %MACROS---
@@
 
-\title{Report on Lab 1 \& Lab 2 in EDA282, Parallel Computer Organization and Design}
+\title{Report on laboratory assignments 1 and 2 for EDA282, \\ Parallel Computer Organization and Design}
 \date{October 2014}
@@
 \maketitle
-$$ Cost = 1 * (WriteHit+ReadHit) + 30 * (readMissesServicedByShared + invalidationsSent)$$ $$+ 20 * LLCHit + 50 * LLCMiss $$
 \input{include/introduction}
@@
 \begin{appendices}
-\section{Parallelized program for Lab 2}
-\centerline{
-\begin{minipage}{1.0\textwidth}
-\lstinputlisting[caption={\textit{Parallel version of original} \texttt{void Solve(double **A).}}, label={lst:solve}]{include/solve.c}
-\end{minipage}
-}
+\input{include/lab2appendice}
 \end{appendices}

### 2014-10-14 lab/report/include/lab2.tex
corrections

@@
 The same procedure is performed on all the other data points, called black dots.
+The distribution of the red and black dots forms a chess pattern.
+This means that when the average for a dot and it's neighbors is calculated, all the neighbors are of the other color.
 The difference of the dots new and old values is added together and the process of calculating new averages is repeated until this accumulated difference of averages, divided by $n^2$, is smaller than $0.01$.
 When the program finishes the matrix have converged to an average of all red and black dots.
-
-The distribution of the red and black dots forms a chess pattern.
-This means that when the average for a dot and it's neighbors is calculated, all the neighbors are of the other color.
-As such, when calculating the average for example the red dots there is no need to invalidate any other dots (than red dots???) because the black dots are only read and you are the only one accessing your red dot.
 
@@
 In listing~\ref{lst:solve} a parallelized version of the function \texttt{void Solve(double **A)} is shown.
-The approach taken to parallilize the function is to distribute the rows of the matrix evenly between all threads.
-This is done twice, once for the red dots and once for the black.
+The approach taken to parallelize the function is to distribute the rows of the matrix evenly between all threads.
+This is done twice, once for the red dots and once for the black dots.
 Only one thread resets \texttt{diff} to zero on line 5 and test whether the program is finished or not on line 25.
-At each for loop that goes through the red and black dots the directive \texttt{reduction(+:diff)} is used which causes each thread to have its own private version of \texttt{diff} that all gets added together after the for loop.
+At each for loop that goes through the red and black dots the directive \texttt{reduction(+:diff)} is used which causes each thread to have its own private version of \texttt{diff} that gets added together after the for loop.
 
 \subsection{Task 3}
-To be able to evaluate the program described in \nameref{subsec:lab2:task2} we decide to mimic the working set sizes in \nameref{sec:lab1}.
-However, the provided data files does not include a matrix big enough to fill neither each threads L1 cache or the LLC cache.
+To be able to evaluate the program described in \nameref{subsec:lab2:task2}, we decide to mimic the working set sizes in \nameref{sec:lab1}.
+However, the provided data files does not include a matrix big enough to fill neither each threads L1 cache or the LLC cache and thus is it expected that the results will not differentiate to a high degree by using the various working sets.
 
-In \reffig{fig:lab2} results from running the parallelized version of the program in the simulator with different CCPs and different matrix dimensions are shown.
-It can be seen that using MESI over MSI does not cause any noticeable boost in performance.
-This is because the gain of the extra E state becomes of no importance as the cache lines is in the M state most of the time.
-Some benefit of the E state can though be seen in Table~\ref{tbl:lab2} as the amount of \textit{WriteMisses} and \textit{InvalidationsSent} are a bit lower for MESI than MSI and the gap becomes bigger as the matrix size increases.
-This does not show in \reffig{fig:lab2} as the cycle count are not dominated by \textit{WriteMisses} and \textit{InvalidationsSent}.
+In \reffig{fig:lab2}, results from running the parallelized version of the program in the simulator with different CCPs and different matrix dimensions are shown.
+It can be seen that using MESI over MSI does not cause any noticeable increase in performance.
+The data elements are mostly in the M state, which makes that the benefit of MESI's E state is low.
+The benefit is shown in Table~\ref{tbl:lab2}, where the number of \textit{WriteMisses} and \textit{InvalidationsSent} are a bit lower for MESI than MSI and the gap becomes bigger as the matrix size increases.
+This does not show in \reffig{fig:lab2}, as the cycle count is not dominated by \textit{WriteMisses} and \textit{InvalidationsSent}.
 
-Changing to MESI-MG yields a huge performance degradation of 420\%, 394\%, and 430\% for matrix size 25, 100, and 200 respectively.
-This comes from the sharing performed of the brim lines between threads.
-For each read of a dot in any of these lines, the dot gets migrated over to the thread performing the read.
-This frequent migration of dots generates a large amount of \textit{ReadMissesSerivcedByModfied} as can be seen in Table~\ref{tbl:lab2}.
+To use MESI-MG instead of MESI or MSI yields a huge performance degradation of 420\%, 394\%, and 430\% for matrix size 25, 100, and 200, respectively.
+This is because the sharing performed of the brim rows between threads.
+For each read of a dot in any of these rows, the dot is migrated to the reading thread.
+The frequent migration of dots generates a large number of \textit{ReadMissesSerivcedByModfied}, as can be seen in Table~\ref{tbl:lab2}.
 The overhead stays fairly constant compared to the other CCPs when the matrix size increases.
-This is duo to the proportion of sharing between the threads stays the same.
+This is due to the proportion of sharing between the threads stays the same.
 
@@
 \subsection{Task 4}
-%\todo[inline]{Outline a strategy to modify the MESI protocol to support the O state.
-Analyze the MESI\_SMPCache.cpp source to understand how the MESI protocol is currently modeled.
-Discuss changes that need to be made to the readLine, writeLine, readRemoteAction, and writeRemoteAction functions.
-Show the necessary changes in pseducode.
-Should the code that tracks cache access statistics be modified?
-If so, how?}
-
-To add the Owner (O) state to the existing MESI CCP only some small changes are needed.
+To add the Owner (O) state to the existing MESI CCP, only small changes are needed.
 In \textit{readLine()} and \textit{remoteWriteAction()} no changes are needed.
@@
 The action taken when in the shared state is removed.
-For the \textit{writeLine()} the only change needed is to take the same actions when in the O state as in the S state.
-For the cache access statistics there is no need to change anything
+For \textit{writeLine()}, the only change needed is to take the same actions as when in the O state as in the S state.
+For the cache access statistics there is no need for change.
 

### 2014-10-14 lab/report/include/discussion.tex
DIscussion

\section{Discussion}
\label{sec:dis}
From the two laboratory assignments performed it has become clear that from a performance point of view MESI is always preferable over MSI. This as the transition to S state comes without any cost so the performance of MESI will be greater or equal to that of MSI. When it comes to MESI-MG the performance impact highly depends on the application. In an application like benchmark 3 where the threads sharing data all performs writes MESI-MG will have a positive impact on performance. On the other hand, in an application where the most sharing consists of reads, like benchmark 1, the impact will be negative.

When analyzing the simulation results it has been noticed that between runs some variations occurred. One solution to this could have been to run the same simulation several times and then use the average. This has however not been done as we are only interested in the general proportions and not the exact values because we want to analyze shown trends.

In \nameref{sec:lab2} the provided working sets were insufficient to mimic the sizes used in \nameref{sec:lab1}. Due to this it was impossible to see if the observed tendencies concerning working set sizes for the three CCPs also would apply on the parallelized program.

### 2014-10-14 lab/report/include/lab1.tex

@@
 \label{sec:lab11}
-Three different microbenchmarks with disgustingly various communication patterns are used in the simulations and here follow a description of each benchmark.
+Three different microbenchmarks with disgustingly various communication patterns are used in the simulations and here follow a description of each.
 \subsubsection*{Benchmark 1 - Producer consumer}
@@
 However, it is not ensured that the incrementing thread increments the values before the other threads reads, and thus can this happen in any order.
-The counter accesses is done in mutual exclusion using a lock, and before a new lap of increments \& reads there is a barrier.
+A counter access is done in mutual exclusion using a lock, and before a new lap of increments \& reads there is a barrier.
 
@@
 Each thread then increments it's assigned counters sequentially.
-A barrier is placed before a new incrementation lap of the assigned array elements begins.
+A barrier is placed before a new incrementation lap of the assigned array elements begin.
 
@@
 \label{sec:lab12}
-Here the CCPs MSI and MESI are compared using benchmarks 1 and 2.
+Here the CCPs MSI and MESI are compared using benchmark 1 and 2.
 The difference between the CCPs is that MESI includes an exclusive (E) state, which is beneficial if data is read exclusively by one processor.
@@
 
-In the producer consumer benchmark, the reading threads transition the data element to the E state when accessing the array for the first time if the writing thread have not accessed it already.
+\todo[inline]{change}
+In the producer consumer benchmark, the reading threads transition the data element to the E state if the writing thread have not accessed it already when accessing the array for the first time or when fetching it from higher level caches.
 However, the whole counter array is shared and thus will not state E be of any gain.
@@
 MESI is here compared with MESI-MG by executing benchmark 1 and 3.
-MESI-MG is a modified version of MESI where the S state is not utilized.
-If a cache issues a read request of a data element which exists in state E or M in a remote cache, the data is forwarded from the remote cache and then the remote cache invalidates it's copy.
+MESI-MG is a modified version of MESI where the S state is not utilized; if a cache issues a read request of a data element which exists in state E or M in a remote cache, the data is forwarded from the remote cache and then the remote cache invalidates it's copy.
 By this scheme, the S state is never used.
@@
 
-In the all shared benchmark, every thread increments every data element during mutual exclusion which favors MESI-MG.
+In the all shared benchmark, every thread increments every data element in mutual exclusion which favors MESI-MG.
 Each data element is read and modified exclusively by one processor at a time, thus is the S state not beneficial.
 As MESI-MG do not utilize the S state, as described, it is more suitable than MESI in this benchmark.
-In Appendix [???] it is shown that the CCPs results are a like, except that MESI have to issue a significant number of invalidates, which is needed for the transition from state S to M.
+In Table \ref{tbl:lab1:task3} it is shown that the CCPs results are alike, except that MESI have to issue a significant number of invalidates, which is needed for the transition from state S to M.
 As shown in \reffig{fig:lab1bars}, MESI-MG's benefit is not visible for the working set size increasing the size of the LLC.

### 2014-10-14 lab/report/include/lab1.tex
corrected

@@
 
-\todo[inline]{change}
-In the producer consumer benchmark, the reading threads transition the data element to the E state if the writing thread have not accessed it already when accessing the array for the first time or when fetching it from higher level caches.
-However, the whole counter array is shared and thus will not state E be of any gain.
-Therefore, MSI and MESI performs equally regardless of working set size, which is shown in \reffig{fig:lab1bars}.
+In the producer consumer benchmark, the E state is beneficial if the writing thread accesses the data element for the first time or when it has been fetched from higher level caches.
+Then, the thread will transition the data element from I to M through E, when performing an increment.
+This avoids the S state, which give an increased performance as no bus command has to be issued.
+However, the probability that the writing thread accesses the data first is low, $\frac{1}{nthreads}$, which is shown in \reffig{fig:lab1bars} where MSI and MESI seems to perform equally.
+As the working set size increases, the benefit of the E state increases due to more frequent reads from globally invalid cache lines.
+This can however not be seen in the simulation results, due to fluctuations.
 

### 2014-10-15 lab/report/include/lab1.tex
corrections

@@
-\section{Lab 1}
+\section{Laboratory assignment 1}
 \label{sec:lab1}
@@
 Three diverse benchmarks, described in section \ref{sec:lab11}, are used to compare the coherence protocols.
-In section \ref{sec:lab12}, MSI and MESI is compared and in section \ref{sec:lab13}, MESI and MESI-MG is compared.
+In section \ref{sec:lab12}, MSI and MESI are compared and in section \ref{sec:lab13}, MESI and MESI-MG are compared.
 In each comparison, we use three working set: the first is smaller than an private L1 cache (512 counters), the second is larger than an private L1 cache and smaller than the LLC (1536 counters), and the last is larger than the LLC (98304 counters).
@@
 \label{sec:lab11}
-Three different microbenchmarks with disgustingly various communication patterns are used in the simulations and here follow a description of each.
+Three different microbenchmarks with disgustingly various communication patterns are used in the simulations, and here follow a description of each.
 \subsubsection*{Benchmark 1 - Producer consumer}
@@
 \subsubsection*{Benchmark 2 - No sharing}
-This benchmark have no sharing of the counter array.
+This benchmark has no sharing of the counter array.
 The counter array is divided up equally among the threads, giving each thread access to every nth counter element, where n is the number of threads.
@@
 In the no sharing benchmark the first read access to every element end in state S for MSI and state E for MESI, respectively.
-The following write transitions the element to state M for both the CCPs, but MSI have to issue a BusUpdate whereas MESI's transition is costless.
+The following write transitions the element to state M for both the CCPs, but MSI has to issue a BusUpdate whereas MESI's transition is costless.
 The BusUpdate broadcasts an invalidate signal which gives MSI a slight reduction in performance compared to MESI.
@@
 As MESI-MG do not utilize the S state, as described, it is more suitable than MESI in this benchmark.
-In Table \ref{tbl:lab1:task3} it is shown that the CCPs results are alike, except that MESI have to issue a significant number of invalidates, which is needed for the transition from state S to M.
+In Table \ref{tbl:lab1:task3} it is shown that the CCPs results are alike, except that MESI has to issue a significant number of invalidates, which is needed for the transition from state S to M.
 As shown in \reffig{fig:lab1bars}, MESI-MG's benefit is not visible for the working set size increasing the size of the LLC.

### 2014-10-15 lab/report/include/discussion.tex
corrections

@@
 \label{sec:dis}
-From the two laboratory assignments performed it has become clear that from a performance point of view MESI is always preferable over MSI.
-This as the transition to S state comes without any cost so the performance of MESI will be greater or equal to that of MSI.
-When it comes to MESI-MG the performance impact highly depends on the application.
+From the two laboratory assignments performed, it has become clear that from a performance point of view MESI is always preferable over MSI.
+This as the transition to S state comes without any cost; the performance of MESI will be greater or equal to that of MSI.
+When it comes to MESI-MG, the performance impact highly depends on the application.
 In an application like benchmark 3 where the threads sharing data all performs writes MESI-MG will have a positive impact on performance.
-On the other hand, in an application where the most sharing consists of reads, like benchmark 1, the impact will be negative.
+On the other hand, in an application where most sharing consists of reads, like benchmark 1, the impact will be negative.
 
-When analyzing the simulation results it has been noticed that between runs some variations occurred.
-One solution to this could have been to run the same simulation several times and then use the average.
-This has however not been done as we are only interested in the general proportions and not the exact values because we want to analyze shown trends.
+When analyzing the simulation results it has been noticed that variations occurred between runs One solution to counter this can be to run the same simulation several times and then use an average.
+This has however not been done, as we are only interested in the general proportions and not the exact values, because we want to analyze shown trends.
 
-In \nameref{sec:lab2} the provided working sets were insufficient to mimic the sizes used in \nameref{sec:lab1}.
-Due to this it was impossible to see if the observed tendencies concerning working set sizes for the three CCPs also would apply on the parallelized program.
+In \nameref{sec:lab2}, the provided working sets were insufficient to mimic the sizes used in \nameref{sec:lab1}.
+Due to this it was impossible to conclude if the observed tendencies concerning working set sizes for the three CCPs from \nameref{sec:lab1} also would apply on the parallelized program.

### 2014-10-16 excersice 1/main.tex
fixed asembler

@@
 \usepackage{fancyhdr}
+\usepackage[margin=1in]{geometry}
 \pagestyle{fancy}
@@
 \item % b 
-See listing~\ref{lst:q3b}.
+See listing~\ref{lst:q4b}.
 
 \item % c
-See listing~\ref{lst:q3c}.
+See listing~\ref{lst:q4c}.
 \end{enumerate}
@@
 \lstset{language={[x86masm]Assembler},frame=single}
-\lstinputlisting[firstline=20, lastline=33, caption={Sync. using T\&S},label={lst:q3b}]{upgASM.asm}
-\lstinputlisting[firstline=37, lastline=43, caption={Sync. using F\&A},label={lst:q3c}]{upgASM.asm}
+\lstinputlisting[firstline=24, lastline=36, caption={Sync. using T\&S},label={lst:q4b}]{upgASM.asm}
+\lstinputlisting[firstline=39, lastline=44, caption={Sync. using F\&A},label={lst:q4c}]{upgASM.asm}
 
@@
 \item % a
-See listing~\ref{lst:q4a}.
+See listing~\ref{lst:q5a}.
 \setcounter{enumi}{3}
 \item % d
-See listing~\ref{lst:q4e}.
+See listing~\ref{lst:q5d}.
 \item % e
-See listing~\ref{lst:q4d}.
+See listing~\ref{lst:q5e}.
 \end{enumerate}
@@
 \lstset{language={[x86masm]Assembler},frame=single}
-\lstinputlisting[firstline=0, lastline=4, caption={F\&ADD X, Rx, a},label={lst:q4a}]{upgASM.asm}
+\lstinputlisting[firstline=2, lastline=5, caption={F\&ADD X, Rx, a},label={lst:q5a}]{upgASM.asm}
 
-\lstinputlisting[firstline=6, lastline=10, caption={SWAP(Rx, X)},label={lst:q4e}]{upgASM.asm}
+\lstinputlisting[firstline=8, lastline=12, caption={SWAP(Rx, X)},label={lst:q5d}]{upgASM.asm}
 
-\lstinputlisting[firstline=12, lastline=18, caption={CAS Rx, Ry, X},label={lst:q4d}]{upgASM.asm}
+\lstinputlisting[firstline=15, lastline=21, caption={CAS Rx, Ry, X},label={lst:q5e}]{upgASM.asm}
 

### 2014-10-16 excersice 1/main.tex
redid

@@
 \usepackage[margin=1in]{geometry}
+\usepackage{color}
 \pagestyle{fancy}
@@
 \begin{document}
-
+\textcolor{red}{\huge Changed stuff marked in red.}
 \section*{Q1}
@@
 \item % d
-A more fine grained solution than the one presented in (c) is shown in listing~\ref{lst:kernelfine} where the \texttt{atomic} directive is used on each of the critical varaibles.
-As explained in (c) only \texttt{new\_centers} and \texttt{new\_centers\_len} need this protection so two locks is enough.
+\textcolor{red}{
+A more fine grained solution than the one presented in (c) is shown in listing~\ref{lst:kernelfine} where \texttt{OpenMP}s runtime locks have been used.
+As explained in (c) \texttt{new\_centers} and \texttt{new\_centers\_len} need protection an now also \texttt{delta} as we no longer use the \texttt{reduction} directive.
+As we protext every singel address new no need $ 1 + 3 \times nclusters$ locks in total.
+}
 
@@
 \lstinputlisting[lastline=31, caption={Modified kernel coarse grained},label={lst:kernelcoarse}]{upgC.c}
+\textcolor{red}{
 \lstinputlisting[firstline=32, caption={Modified kernel fine grained},label={lst:kernelfine}]{upgC.c}
-
+}
 \section*{Q2}
@@
 \lstinputlisting[firstline=24, lastline=36, caption={Sync. using T\&S},label={lst:q4b}]{upgASM.asm}
+\textcolor{red}{
 \lstinputlisting[firstline=39, lastline=44, caption={Sync. using F\&A},label={lst:q4c}]{upgASM.asm}
-
+}
 \section*{Q5} % 7.3
@@
 \lstset{language={[x86masm]Assembler},frame=single}
+\textcolor{red}{
 \lstinputlisting[firstline=2, lastline=5, caption={F\&ADD X, Rx, a},label={lst:q5a}]{upgASM.asm}
-
+}
+\textcolor{red}{
 \lstinputlisting[firstline=8, lastline=12, caption={SWAP(Rx, X)},label={lst:q5d}]{upgASM.asm}
-
+}
+\textcolor{red}{
 \lstinputlisting[firstline=15, lastline=21, caption={CAS Rx, Ry, X},label={lst:q5e}]{upgASM.asm}
+}
 

### 2014-10-16 paper/include/reducing_refresh_power.tex
corrections

@@
 \label{sec:red}
-\todo[inline]{Do an intro that includes a road map, mention the refresh scheduling approach that schedules real access together with refresh ops. Make an intro to ARTV (somewhat done) }
-There are several methods that can be applied to decrease the refresh rate and these can be categorized depending on with sort of information the scheduling decisions are based upon.
-The four sort of information are: access recency, cell retention times, error tolerance of the data, and line validity.
-This categorization follows the taxonomy proposed by authors \cite{dtail}.
+There are several approaches to decrease the refresh power by reducing unnecessary refreshes and these can be categorized depending on with sort of information the scheduling decisions are based upon.
+The four types of information are: access recency, cell retention time, error tolerance of the data, and line validity.
+This categorization follows the taxonomy proposed by Cui et al.
+\cite{dtail}.
+These four approaches are explained further in \ref{sec:red:app}, whereas techniques based on the approaches are summarized in \ref{sec:red:tech}.
 
 \subsection{Approaches to reduce refresh power}
+\label{sec:red:app}
 
@@
 \subsection{Techniques to reducing refresh power}
+\label{sec:red:tech}
 

### 2014-10-16 paper/include/refrint.tex
corrections

--- 2014-10-08 paper/include/refrint.tex
@@
-This technique is proposed by A.
-Agrawal et al.
+This technique is proposed by Agrawal et al.
 \cite{refrint} and is based on Access Recency.
-It is targeted towards caches and eDRAM and thus can not be directly compared with the other solutions, but the main concept is of good use and the technique could be implemented DRAM as well.
-The concept is based on two types of unnecessary refreshes which originates from so called cold and hot rows.
+It is targeted towards caches and eDRAM, and thus can not be directly compared with the other solutions, but the main concept is of good use and the technique could be implemented DRAM as well.
+The concept is based on two types of unnecessary refreshes which originate from so called cold and hot rows.
 Cold rows are those used far apart in time or not used at all.
 The technique identifies and refreshes only those rows which are expected to be accessed in the near future, the rows that are used less frequent is invalidated.
-Hot rows consists of the rows that gets accessed frequently and thus can the technique postpone the preceding refresh to the accessed rows.
+Hot rows consist of the rows that gets accessed frequently and thus can the technique postpone the preceding refresh to the accessed rows.
 
@@
 The differences are that Refrint does not maintain a counter for each row, but a snapshot of a global counter as well as the row's valid bit.
-From experiments, they found out that a good size for the global counter was 1 or 2 bits.
 When a row is accessed, the corresponding snapshot is updated with the current value of the global counter.
 Whenever the global counter steps to a new value, all normal accesses are stalled and the logic checks whether any of the valid row's snapshot  matches the global counter.
-If there is a match, a refresh is scheduled or the row invalidated depending on the polices used for the cold rows.
+If there is a match, a refresh is scheduled or the row is invalidated depending on the polices used for the cold rows.
 
-Four different policies can be applied for the cold rows, where the policies decides what to refresh; \textit{All}, \textit{Valid}, \textit{Dirty}, or \textit{WB(n,m)}.
-\textit{All} refreshes every row, irregardless of whether it is valid or not.
+Four different policies can be applied for the cold rows, where the policies decide what to refresh; \textit{All}, \textit{Valid}, \textit{Dirty}, or \textit{WB(n,m)}.
+\textit{All} refreshes every row, regardless of whether it is valid or not.
 \textit{Valid} and \textit{Dirty} refreshes valid and dirty rows, respectively, and otherwise invalidate the row.
-\textit{WB(n,m)} refreshes a dirty row \textit{n} times before flushing it and changing the row state to valid clean, then the row is refreshed \textit{m} times after the last access before it is invalidated.
-\textit{N} and \textit{m} are time-out counters that are decremented upon refresh.
+\textit{WB(n,m)} refreshes a dirty row $n$ times before flushing it and changing the row state to valid clean, then the row is refreshed $m$ times after the last access before it is invalidated.
+$n$ and $m$ are time-out counters that are decremented upon refresh.
 
@@
 
-Depending on the application's memory footprint and cache visibility, the program can be categorized to maximize the gain of the technique.
-Visibility means that the In \reffig{fig:app_cat} three categories are displayed.
-Explain the categories.
-
-\begin{figure}[t!]
-	\includegraphics[width=\textwidth/2]{app_cat}
-	\caption{Application categorization according to the cold row's policies \cite{refrint}.}
-	\label{fig:app_cat}
-\end{figure}
+Depending on the application's memory footprint and cache visibility the program is categorized to maximize the gain of the policies for cold rows.
+Visibility correspond to the amount of cache row accesses made by the application.
+If the application has high visibility, \textit{WB} is used and if the visibility is low, \textit{Valid} is used.
+A high visibility and a small footprint works best with small $n$ and $m$, whereas a large footprint is likely more usefull with small \textit{n} and \textit{m}.
 

### 2014-10-16 paper/include/smart_refresh.tex
corrections

@@
-M.
-Ghosh and H.-H.
-S.
-Lee proposes the technique Smart Refresh \cite{smartrefresh} that targets Access Recency.
+% Smart Refresh
+Ghosh and Lee propose the technique Smart Refresh \cite{smartrefresh} that targets Access Recency.
 The basic idea is to maintain a time-out counter for each memory row and when a counter time-out, the MC issues a ROR refresh to the corresponding row.
@@
 To achive this, the counters are initially staggered and divided into n logical blocks where the counters are accessed in a sequential order.
-Only n counters will be accessed, set or decremented at the same point in time, which result in that maximum n rows are refreshed simultainously.
+Only $n$ counters will be accessed, set or decremented at the same point in time, which result in that maximum $n$ rows are refreshed simultainously.
 
@@
 The least optimal case for postponing the refresh is if the row access occurs a small time intervall before the counter is accessed.
-If a 2-bit counter is used the least optimal row access result in that the row is refreshed after 3/4 of the maximum refresh period.
-Analogously, a 3-bit counter result in that a row is refreshed after minimum 7/8 of the maximum refresh period.
+If a two-bit counter is used the least optimal row access result in that the row is refreshed after $3/4$ of the maximum refresh period.
+Analogously, a three-bit counter result in that a row is refreshed after minimum $7/8$ of the maximum refresh period.
 
-If the row accesses drops below 1\% during a maximum refresh period, Smart Refresh is disabled.
+If the row accesses drops below $1\%$ during a maximum refresh period, Smart Refresh is disabled.
 As stated, an access recency technique is dependent on row accesses to be of gain.
-When disabled, the MC uses it's default CBR refresh scheme.
-Smart refresh is then activated if the row accesses increases above 2\%.
+When disabled, the MC uses its default CBR refresh scheme.
+Smart refresh is then activated if the row accesses increases above $2\%$.
 
 Smart Refresh modifies the MC but does not change the DRAM module nor the interface between them.
-The modifications of the MC constitutes of counter update logic, the counter array, and a pending refresh request queue.
-The memory footprint of the modifications is 0.0024\% of total device capacity when 3-bits counters are used.
+The modifications of the MC constitute of counter update logic, the counter array, and a pending refresh request queue.
+The memory footprint of the modifications is $0.0024\%$ of total device capacity when three-bits counters are used assuming a cache row size of $16\:KB$.
 
-The technique result in an average refresh reduction of 59\% and saves on average 52\% of refresh power consumption without any performance degradation.
-The simulation was performed on 2 GB of main memory.
+The technique was simulated on $2$ GB of main memory.
+Result shows that the refresh reduction was in average $59\%$ and refresh power consumption savings was on average $52\%$, all without any performance degradation.

### 2014-10-16 excersice 1/main.tex

@@
 \label{itm:3a}
+\textcolor{red}{
 The amount of cycles will be the following for MSI:
 \texttt{
-  \\Proc1: LvL2Rd + (LvL1RdX + BusUP + Wr) + Wr
-  \\Proc2: LvL2Rd + (LvL1RdX + BusUP + Wr) + Wr
-  \\Proc3: LvL2Rd + (LvL1RdX + BusUP + Wr) + Wr
-  \\Proc4: LvL2Rd + (LvL1RdX + BusUP + Wr) + Wr 
+  \\Proc1: (Snoop + LvL2Rd + Rd) + (BusUP + Snoop + Wr) + Wr
+  \\Proc2: (Snoop + LvL1Rd + Rd) + (BusUP + Snoop + Wr) + Wr
+  \\Proc3: (Snoop + LvL1Rd + Rd) + (BusUP + Snoop + Wr) + Wr
+  \\Proc4: (Snoop + LvL1Rd + Rd) + (BusUP + Snoop + Wr) + Wr 
 }
-\[4\times(40+(20+10+1)+1) = 288\]
+}
+\textcolor{red}{
+\[(5+40+1)+(10+5+1)+1+3\times((5+20+1)+(10+5+1)+1) = 192\]
 and for MESI:
 \texttt{
-  \\Proc1: LvL2Rd + Wr + Wr
-  \\Proc2: LvL2Rd + (LvL1RdX + BusUP + Wr) + Wr
-  \\Proc3: LvL2Rd + (LvL1RdX + BusUP + Wr) + Wr
-  \\Proc4: LvL2Rd + (LvL1RdX + BusUP + Wr) + Wr 
+  \\Proc1: (Snoop + LvL2Rd + Rd) + Wr + Wr
+  \\Proc2: (Snoop + LvL1Rd + Rd) + (BusUP + Snoop + Wr) + Wr
+  \\Proc3: (Snoop + LvL1Rd + Rd) + (BusUP + Snoop + Wr) + Wr
+  \\Proc4: (Snoop + LvL1Rd + Rd) + (BusUP + Snoop + Wr) + Wr 
 }
-\[(40+1+1)+3\times(40+(20+10+1)+1) = 258\]
-where the difference is due to the first processor going to the E state instead of the S state and thereby saves 30 cycles when it then wants to write.
-There are no Snoops done as we only care about the cycles wasted in the current cache.
+}
+\textcolor{red}{
+\[(5+40+1)+1+1+3\times(5+20+1+(10+5+1)+1) = 177\]
+where the difference is due to the first processor going to the E state instead of the S state and thereby saves 15 cycles when it then wants to write by not needing to emit a busupgrade and or snoop the bus.
+I assume that the bus upgrades does not include any snoop.}
 
 \item % b 
+\textcolor{red}{
 The amount of trafic that gets generated for MSI based on the sequences in (a)):
-\[4\times(38B+38B+10B) = 344B\]
+\[4\times(38B+10B) = 192B\]
 and for MESI:
-\[38B+3\times(38B+38B+10B) = 296B\]
+\[38B+3\times(38B+10B) = 182B\]
+}
 \end{enumerate}

### 2014-10-16 paper/include/dtail.tex
durrHurr

@@
-Z.
 Cui et al.
-\cite{dtail} proposes DTail.
-They evaluates two versions of DTail referenced as DTail-R and DTail-V utilizing \textbf{R} and \textbf{V} data respectively.
-Even so, their main idea is that more types of refresh information can be incorporated and stored together with the \textbf{R} and \textbf{V} information.
+\cite{dtail} propose DTail.
+They evaluate two versions of DTail referenced as DTail-R and DTail-V utilizing \textit{retention time} and \textit{validity} information respectively.
+Even so, their main idea is that more types of refresh information can be incorporated and stored together with the \textit{retention time} and \textit{validity} information.
 DTail can then choose the most efficient refresh schema depending on the current situation, i.e.

### 2014-10-16 paper/include/raidr.tex
durrHurr

@@
-RAIDR by J.
-Liu et al.
-\cite{raidr} divides all memory lines into different groups, called bins, based on the cell with the shortest retention time in the line.
-In the basic configuration there are two bins, one with a refresh rate of \(64ms - 128ms\) and one with \(128ms - 256ms\).
-Lines in the first bin gets refreshed every \(64ms\), lines in the second bin every \(128ms\), and lines not in any bin gets refreshed at the default refresh rate which is set to \(256ms\).
-The implementation is done inside the memory controller and to minimize the storage needed for keeping track of bin members they use Bloom filters.
-Thanks to this only 1.25 KB storage is needed for a 32 GB DRAM with the default configuration.
-The storage requirement can however vary depending the amount of bins used.
-RAIDR also accommodate for temperature variations by scaling all bins refresh rates using a timer.
-With RAIDR a performance improvement of 4.1\% and a decrease in power consumption by 8.3\% is achieved in a system with 32GB DRAM compared to auto-refresh.
+In this approach, RAIDR by Liu et al.
+\cite{raidr}, all memory rows gets divided into different groups, called bins, based on the cell with the shortest retention time in the row.
+In the basic configuration there are two bins, one with rows that need a refresh rate of $64ms \to 128ms$ and one with rows that need to be refreshed every $128ms \to 256ms$.
+Rows in the first bin gets refreshed every $64ms$, rows in the second bin every $128ms$, and all remaining rows gets refreshed at the default refresh rate which is set to $256ms$.
+This is done using two counters, one row counter that rolls over every $64ms$, and one counter that keeps track on how many times the row counter have rolled.
+With this every row will get selected as refresh candidate every $64ms$, the refresh will be performed depending on which bin the row belongs to and on the second counter's value.
+The actual refreshes are performed by activating the row in question essentially performing a ROR.
+
+The implementation is done inside the memory controller meaning RAIDR can not be adopted in currently existing systems.
+To minimize the storage needed for keeping track of bin members they use Bloom filters.
+Thanks to this, only 1.25 KB storage is needed for a 32 GB DRAM with the default configuration.
+The storage requirement can however vary depending the amount of bins used, but on the other hand the granularity of the approach becomes higher with more bins.
+
+RAIDR conform to temperature variations by utilizing a third counter that is used to scale all bins refresh rate with the temperature.
+Comparing RAIDR with the somewhat similar DTail-R Lie et al.
+does not achieve quite as big decrease in refresh operations as Cui et al.
+does.
+This is due to DTail-R having a higher granularity as the refresh rate is kept per row and not per row group as in RAIDR.
+Even so, a performance improvement of 4.1\% and a decrease in power consumption by 8.3\% is achieved in a system with 32GB DRAM compared to auto-refresh.

### 2014-10-16 paper/include/background.tex

@@
 The word line connects a number of cells which forms a row.
-Each bit line is connected to a sense amplifier, and the row of sense amplifiers constitutes the \textit{row buffer}.
+Each bit line is connected to a sense amplifier, and the row of sense amplifiers constitute the \textit{row buffer}.
 Data is stored as a charge on the cell capacitor.
@@
 The charge stored on the capacitors has, as mentioned, left the capacitor and a read operation is thus destructive.
-To maintain the data the capacitors are recharged by the bit lines, which are driven to either either $V_{DD}/2$ or $0\:V$ by the sense amplifiers depending on the former voltage detected.
+To maintain the data the capacitors are recharged by the bit lines, which are driven to either $V_{DD}/2$ or $0\:V$ by the sense amplifiers depending on the former voltage detected.
 Finally, when another row on the bank is accessed, the open row has to be closed, which disconnects the capacitors from the bit lines.
@@
 The row buffer is then loaded with the data that are driven to the bit lines by the sense amplifiers, thus storing the information on the capacitors.
-The row will be closed when another row is accessed.
+The row is then closed when another row is accessed.
 
@@
 To refresh one rank, the simplest approach is to refresh all banks in parallel row-by-row at the start of the $64\:ms$ interval.
-This approach is called \textit{burst refresh}.
-The approach blocks normal accesses for a long period, and power consumption peaks during the burst time.
+This approach is called \textit{burst refresh} and blocks normal accesses during the burst time, where power consumption peaks.
 A better alternative is to use \textit{distributed refresh}, which scatter the refreshes across the interval, resulting in a lower average latency for the processor and more stable power consumption.
@@
 The first is called \textit{RAS-only refresh} (ROR) where a refresh cycle refreshes one targeted row.
-To target one row, the row's address is put on the bus, and the MC have the responsibility that a row is refreshed within the maximum refresh period.
+To target one row, the row's address is put on the bus, and the MC has the responsibility that a row is refreshed within the maximum refresh period.
 The second is called \textit{CAS before RAS refresh} (CBR).
-In this scheme, the memory module have an internal address counter that displays which row that will be refreshed.
+In this scheme, the memory module has an internal address counter that hold which row that will be refreshed.
 On a refresh cycle, the row stated by the internal address counter is refreshed and the counter is then incremented.

### 2014-10-16 paper/include/introduction.tex

@@
 Energy consumption in modern computer system is an increasingly big system design constraint.
-For example, battery powered mobile systems are getting more common and they needs to be energy efficient to have a long battery life.
+For example, battery powered mobile systems are getting more common and they need to be energy efficient to have a long battery life.
 Moreover, in large data centers the power consumption corresponds to a large part of the upkeep cost.
 
-DRAM is the most commonly used main memory technology in modern computer systems and represents a growing portion of system power consumption \cite{exascale}.
+DRAM is the most commonly used main memory technology in modern computer systems and represent a growing portion of system power consumption \cite{exascale}.
 This portion will continue to increase as we strive for higher and higher memory capacity.
-A large part of the DRAM power consumption comes from refreshing all memory-cells in the DRAM.
+A large part of the DRAM power consumption originates from refreshing all DRAM memory-cells.
 This refreshing has to be done as DRAM's store data as a charge in a capacitor, and these charges leak over time.
 To increase DRAM performance and lower access latencies, 3D stacked DRAM has been proposed, but stacked design have trouble dissipating heat.
-The higher temperature increase leakage, and the DRAM will have to be refreshed more often.
+The higher temperature increases leakage, and the DRAM will have to be refreshed more often.
 

### 2014-10-17 paper/include/background.tex
flikker started

@@
 
-The memory controller (MC) normally issues refresh operations periodically at an interval of $64\:ms$ for each row, an interval that has been constant for the latest DRAM specifications [source] and called \textit{maximum refresh period}.
+\todo[inline]{change the next paragraph to introduce retention time.}
+The memory controller (MC) normally issues refresh operations periodically at an interval of \textit{64~ms} for each row, an interval that has been constant for the latest DRAM specifications [source] and called \textit{maximum refresh period}.
 This interval is conservatively halved if the DRAM exceeds the specified working temperature of $85^{\circ}C$.
 
-To refresh one rank, the simplest approach is to refresh all banks in parallel row-by-row at the start of the $64\:ms$ interval.
+To refresh one rank, the simplest approach is to refresh all banks in parallel row-by-row at the start of the \textit{64~ms} interval.
 This approach is called \textit{burst refresh} and blocks normal accesses during the burst time, where power consumption peaks.
@@
 Additional refresh operations give larger power consumption and the refreshes become a growing portion of the total DRAM energy consumption, as shown in \reffig{fig:dram_power}.
-When the device capacity reaches $64\:GB$, the refresh power nearly constitutes half of the total consumption, an increase from 15\% if the device capacity is 8 GB.
+When the device capacity reaches \textit{64~GB}, the refresh power nearly constitutes half of the total consumption, an increase from 15\% if the device capacity is 8 GB.
 

### 2014-10-17 paper/include/refrint.tex
flikker started

@@
 This technique is proposed by Agrawal et al.
-\cite{refrint} and is based on Access Recency.
-It is targeted towards caches and eDRAM, and thus can not be directly compared with the other solutions, but the main concept is of good use and the technique could be implemented DRAM as well.
+\cite{refrint} and is based on Access Recency and Validity.
+It is targeted towards caches and eDRAM, and thus can not be directly compared with the other solutions, but the main concept is of good use and the technique could be implemented in DRAM as well.
 The concept is based on two types of unnecessary refreshes which originate from so called cold and hot rows.
@@
 \textit{WB(n,m)} refreshes a dirty row $n$ times before flushing it and changing the row state to valid clean, then the row is refreshed $m$ times after the last access before it is invalidated.
-$n$ and $m$ are time-out counters that are decremented upon refresh.
+Moreover, \textit{WB} is implemented using so called \textit{Count} bits, which is set either to $n$ or $m$ and is decremented upon refresh.
+% Mention the State bits as well?
 
-\todo[inline]{Drafty text below!}
+Depending on the application's memory footprint and cache visibility the program is categorized to maximize the gain of the policies for cold rows, where visibility correspond to the number of cache row accesses made by the application.
+If the application has high visibility, \textit{WB} is used and if the visibility is low, \textit{Valid} is used.
+A high visibility and a small footprint works best with large $n$ and $m$, whereas a high visibility and a large footprint likely benefits more from small \textit{n} and \textit{m}.
 
-Depending on the application's memory footprint and cache visibility the program is categorized to maximize the gain of the policies for cold rows.
-Visibility correspond to the amount of cache row accesses made by the application.
-If the application has high visibility, \textit{WB} is used and if the visibility is low, \textit{Valid} is used.
-A high visibility and a small footprint works best with small $n$ and $m$, whereas a large footprint is likely more usefull with small \textit{n} and \textit{m}.
+\todo[inline]{Drafty below. Ask Sally, confusing paper.}
 
-The technique modifies the MC. Which modifications and how large overhead? 5 bit snapshot for each row + state and n+m counters. some logic as well.
+Refrint modifies the MC and also stores data on chip(?).
+In an unmodified MC, there already exist a counter which can be used as the global counter and the authors have chosen to store the $N$ snapshot bits in the MC as well as the rows' valid bits in the MC.
+From experiments, the authors found out that one or two are good values for $N$.
+The logic to process the policies for the cold rows consists of decision logic, as well as Count bits and State bits for each cache row.
+This logic is located on chip(?)/in DRAM(?)(???) and has apparently a negligible overhead.
+In simulations, $N$ was two bits and Count was 5 bits, giving a total overhead of $0.6\%$, when cache row size is 64 bytes.
 

### 2014-10-17 paper/include/smart_refresh.tex
flikker started

@@
 The modifications of the MC constitute of counter update logic, the counter array, and a pending refresh request queue.
-The memory footprint of the modifications is $0.0024\%$ of total device capacity when three-bits counters are used assuming a cache row size of $16\:KB$.
+The memory footprint of the modifications is $0.0024\%$ of total device capacity when three-bits counters are used assuming a cache row size of \textit{16~KB}.
 
-The technique was simulated on $2$ GB of main memory.
+The technique was simulated on \textit{2~GB} of main memory.
 Result shows that the refresh reduction was in average $59\%$ and refresh power consumption savings was on average $52\%$, all without any performance degradation.

### 2014-10-19 paper/include/flikker.tex
flikker started
mostly wrote about Flikker

@@
 %flikker
-Tolerance!\cite{flikker}
-\todo[inline]{Add description.}
+The technique targets Data Tolerance and is proposed by Liu et al.
+\cite{flikker} and  distinguish data as critical or non-critical.
+The critical data is refreshed at the normal maximal refresh period, whereas the non-critical data is refreshed at a much lower rate.
+Refreshing at a lower rate saves refreshing power but increases data corruption.
+This trade-off was found to be balanced with a refresh period of one second, where the refresh power savings was expected to be $23\%$ and the error rate $4.0 \times 10^{-8}$, if the critical part is $25\%$.
+
+Flikker refreshes the data by extending the \textit{Partial Array Self Refresh} (PASR) mode which was introduced 2002 in mobile DRAMs.
+PASR mode refreshes a portion of the DRAM array of one bank when the mobile system is in sleep mode, and the technique modifies PASR's internal address counter to support two refresh rates.
+
+To support the proposed partition scheme, modifications in the operating system, runtime system, and each application are required.
+In the application, the developer distinguish the program variables as either critical or non-critical.
+Then, the runtime system uses a custom allocator that places the two types of variables on separate pages.
+The OS extends the page-table entries one bit, which are utilized by the runtime system to mark the pages depending on its criticality.
+Pages are then mapped by the OS based on the extended bit, to either the normal rate refreshed part or the low rate refreshed part of a bank.
+% To detailed.
+
+% Add the different configurations that were tested.
+
+Flikker was simulated on a mobile device.
+The results of some of the tested applications were degraded, whereas most applications were not affected by the injected faults at all.
+The partitioning scheme can potentially lower performance as locality of data elements changes, but the found performance degradation was $1\%$ or less.
+No loss of reliability was experienced, and the technique saved between $20$-$25\%$ of total DRAM power consumption depending on application.

### 2014-10-20 paper/include/sparkk.tex
wrote, added TODOs and sources
mostly wrote about Flikker
started writing on Sparkk

--- 2014-09-19 paper/include/retention.tex
@@
-\section{Retention}
-\label{sec:ret}
-\cite{ACM:dtail}
+%Sparkk
+Lucas et al.
+\cite{sparkk} proses Sparkk, a technique based on Data Tolerance and Retention Time.
+Sparkk gives fine grained control of refresh rates on a banks' rows.
+Bit mapping is used to distinguish bits with different criticality, where the bits of the same criticality is placed near each other of the global rank row.
+This scheme gives the advantage to refresh the highly critical bits at one refresh rate, whereas the refresh rate decreases along with the criticality.
+ADD that it need to have per bank refresh........
+
+The refresh scheme relies on an ordered list of one entry per memory area, which is traversed through by the internal address counter.
+Each memory area is formed of any number of consecutive rows and can have different refresh periods for each bank.
+An entry consists of the chosen refresh rates, time-out bits, and the address to the end of the memory area.
+When the counter address and the address to the end of the memory area matches, the MC advances to the next entry.
+There, the MC decides if to issue a refresh operation and to which subranks.
+
+\todo[inline]{Drafty below}
+Downside: SW modifications. Mapping has to be created for all data types. Somewhat complex -> draws more power = mitigates advantages.

### 2014-10-20 paper/include/background.tex
wrote wrote

@@
 The word line connects a number of cells which forms a row.
+In turn, one row from each bank form a global row.
 Each bit line is connected to a sense amplifier, and the row of sense amplifiers constitute the \textit{row buffer}.
@@
 The second is called \textit{CAS before RAS refresh} (CBR).
-In this scheme, the memory module has an internal address counter that hold which row that will be refreshed.
-On a refresh cycle, the row stated by the internal address counter is refreshed and the counter is then incremented.
+In this scheme, the MC has an internal address counter that hold which global row that will be refreshed.
+On a refresh cycle, the global row stated by the internal address counter is refreshed and the counter is then incremented.
 Contrary to ROR, CBR do not need provided addresses on the bus, which results in a lower power consumption and less bus usage.
@@
 As refresh operations are made by row, the refresh operations will increase linearly with the device capacity.
-In turn, the throughput loss increases exponentially, as shown in \reffig{fig:dram_throughput}.
-Additional refresh operations give larger power consumption and the refreshes become a growing portion of the total DRAM energy consumption, as shown in \reffig{fig:dram_power}.
+In turn, the throughput loss increases exponentially, as in \reffig{fig:dram_throughput}.
+Additional refresh operations give larger power consumption and the refreshes become a growing portion of the total DRAM energy consumption, as in \reffig{fig:dram_power}.
 When the device capacity reaches \textit{64~GB}, the refresh power nearly constitutes half of the total consumption, an increase from 15\% if the device capacity is 8 GB.

### 2014-10-20 paper/include/flikker.tex
wrote wrote

@@
 %flikker
-The technique targets Data Tolerance and is proposed by Liu et al.
-\cite{flikker} and  distinguish data as critical or non-critical.
+The technique is proposed by Liu et al.
+\cite{flikker} and targets Data Tolerance and Validity.
+Flikker distinguish data as critical or non-critical.
 The critical data is refreshed at the normal maximal refresh period, whereas the non-critical data is refreshed at a much lower rate.
@@
 In the application, the developer distinguish the program variables as either critical or non-critical.
-Then, the runtime system uses a custom allocator that places the two types of variables on separate pages.
-The OS extends the page-table entries one bit, which are utilized by the runtime system to mark the pages depending on its criticality.
-Pages are then mapped by the OS based on the extended bit, to either the normal rate refreshed part or the low rate refreshed part of a bank.
-% To detailed.
+Then the runtime system and OS keep the data of different criticality apart by allocating them on different pages and mapping these pages to the corresponding part of the bank.
 

### 2014-10-20 paper/include/sparkk.tex
wrote wrote

@@
 Lucas et al.
-\cite{sparkk} proses Sparkk, a technique based on Data Tolerance and Retention Time.
+\cite{sparkk} proses Sparkk, a technique based on Data Tolerance and Validity.
 Sparkk gives fine grained control of refresh rates on a banks' rows.
-Bit mapping is used to distinguish bits with different criticality, where the bits of the same criticality is placed near each other of the global rank row.
+Bit mapping is used to distinguish bits with different criticality, where the bits of the same criticality is placed near each other in the global row.
 This scheme gives the advantage to refresh the highly critical bits at one refresh rate, whereas the refresh rate decreases along with the criticality.
-ADD that it need to have per bank refresh........
 
@@
 Each memory area is formed of any number of consecutive rows and can have different refresh periods for each bank.
-An entry consists of the chosen refresh rates, time-out bits, and the address to the end of the memory area.
+An entry consists of the chosen refresh period for each bank, time-out bits for each bank, and the address to the end of the memory area.
 When the counter address and the address to the end of the memory area matches, the MC advances to the next entry.
-There, the MC decides if to issue a refresh operation and to which subranks.
+There, the MC decides if to issue a refresh operation and if so, which banks to send the command to.
+The DRAM device has to provide the ability of per bank refresh commands for this to work.
 
 \todo[inline]{Drafty below}
-Downside: SW modifications. Mapping has to be created for all data types. Somewhat complex -> draws more power = mitigates advantages.
+
+Similar to Flikker, Sparkk needs software support to partition the data.
+As Sparkk is a more fine grained technique then Flikker, a more extensive support is needed.
+Lucas et al.
+proposes a complex .....
+
+Results:

### 2014-10-21 paper/include/dtail.tex
corrections

@@
 \cite{dtail} propose DTail.
-They evaluate two versions of DTail referenced as DTail-R and DTail-V utilizing \textit{retention time} and \textit{validity} information respectively.
-Even so, their main idea is that more types of refresh information can be incorporated and stored together with the \textit{retention time} and \textit{validity} information.
+They evaluate two versions of DTail referenced as DTail-R and DTail-V utilizing Retention Time and Validity information, respectively.
+Even so, their main idea is that more types of refresh information can be incorporated and stored together with the Retention time and Validity information.
 DTail can then choose the most efficient refresh schema depending on the current situation, i.e.
 workload and type of application.
-Refresh data is kept per row and is stored in the DRAM itself instead of storing it in the memory controller which many other solutions do \cite{raidr}\cite{smartrefresh}\cite{refrint}.
-To counter the added latency of a slower memory they utilizes spatial locality of the refresh data and performs prefetching into a FIFO buffer.
+Refresh data is kept per row and is stored in the DRAM itself instead of storing it in the MC which many other solutions do \cite{raidr}\cite{smartrefresh}\cite{refrint}.
+To counter the added latency of a slower memory, Cui et al.
+utilizes spatial locality of the refresh data and performs prefetching into a FIFO buffer.
 As the refresh data is kept on a row level DTail can save more refreshes than other techniques that have a lower granularity.
 
-For DTail-R, the retention time data consists of 3 bits and is interpreted in such way that the refresh period can become \(64ms \times 2^n, n = [0...7]\).
-With DTail-V one additional bit is added to the refresh data which is used to indicate if the row is valid or not.
-With this structure of four bits per row, the amount of memory required is weighing in on a negligible amount of 0.006\% DRAM capacity assuming a row size of 8 KB.
+For DTail-R, the Retention Time data $n$ consists of three bits and is interpreted in such way that the refresh period can become \(64ms \times 2^n, n = [0..7]\).
+With DTail-V, one additional bit is added to the refresh data which is used to indicate if the row is valid or not.
+With this structure of four bits per row, the amount of memory required is $0.006\%$ of DRAM capacity assuming a row size of $8\:KB$.
 
 When a row is to be scheduled for refresh the refresh data is checked.
-If the row is valid and the time since the last refresh is \(64ms \times 2^n, n = [0...7]\) a refresh is performed, otherwise not.
-The refreshes can be done either by RAS commands or by auto refresh depending on what is most efficient at the time.
-If a large amount of rows needs refresh then auto refresh is selected, otherwise RAS-only refresh are used.
+If the row is valid and the time since the last refresh is \(64ms \times 2^n, n = [0..7]\) a refresh is performed, otherwise not.
+The refreshes can be done either by ROR or by CBR depending on what is most efficient at the time.
+If a large number of rows need refresh then CBR is selected, otherwise ROR is used.
 
-When only DTail-R is used, they manages to decrease the amount of refreshes with as much as 87.9\% and thereby only have 16.7\% energy overhead compared to a refresh less 32GB DRAM.
-With only DTail-V power consumption and performance overhead increase from 4\% and 3\% to 90\% and 23\% respectively when going from 10\% to 90\% memory capacity utilization compared to a refresh less 32 GB system.
-The power of DTail shows when DTail-R and DTail-V are used together and at 10\% memory utilization a performance overhead of only 1\% and close to 0\% increase in energy consumption is achieved.
-When the utilization is increased to 90\% they increase to 2\% and 15\% respectively.
+When only DTail-R is used, Cui et al.
+manages to decrease the number of refreshes with as much as $87.9\%$ and thereby only have $16.7\%$ energy overhead compared to a refresh less 32GB DRAM.
+With only DTail-V power consumption and performance overhead increase from $4\%$ and $3\%$ to $90\%$ and $23\%$, respectively, when going from $10\%$ to $90\%$ memory utilization compared to a refresh less 32 GB system.
+The power of DTail shows when DTail-R and DTail-V are used together and at $10\%$ memory utilization, a performance overhead of only $1\%$ and close to $0\%$ increase in energy consumption is achieved.
+When the utilization is increased to $90\%$, they increase to $2\%$ and $15\%$, respectively.

### 2014-10-21 paper/include/rioparis.tex
durrHurr
write

@@
-RIO is a \textbf{R} based technique proposed by S.
-Baek et al.
+RIO is a Retention Time based technique proposed by Baek et al.
 \cite{rioparis} in which they identify cells with low retention time (bad cells) and `deletes' them.
@@
 
-S.
 Baek et al.
-also propose PARIS which can be combined with RIO for further improvements.
-In PARIS the goal is not to make the refresh period as long as possible but to not refresh unused memory rows at all using \textbf{V} data.
+also propose PARIS \cite{rioparis} which can be combined with RIO for further improvements.
+In PARIS the goal is not to make the refresh period as long as possible but to not refresh unused memory rows at all using Validity information.
 This is done by keeping track of all page frames allocated and deallocated by the Linux kernel.

### 2014-10-21 paper/include/dtail.tex
reordering of sections

@@
 
-For DTail-R, the Retention Time data $n$ consists of three bits and is interpreted in such way that the refresh period can become \(64ms \times 2^n, n = [0..7]\).
+For DTail-R, the Retention Time data $n$ consists of three bits and is interpreted in such way that the refresh period can become \(64\:ms \times 2^n, n = [0..7]\).
 With DTail-V, one additional bit is added to the refresh data which is used to indicate if the row is valid or not.
-With this structure of four bits per row, the amount of memory required is $0.006\%$ of DRAM capacity assuming a row size of $8\:KB$.
+With this structure of four bits per row, the amount of memory required is $0.006\%$ of DRAM capacity assuming a row size of \textit{8 KB}.
 
 When a row is to be scheduled for refresh the refresh data is checked.
-If the row is valid and the time since the last refresh is \(64ms \times 2^n, n = [0..7]\) a refresh is performed, otherwise not.
+If the row is valid and the time since the last refresh is \(64\:ms \times 2^n, n = [0..7]\) a refresh is performed, otherwise not.
 The refreshes can be done either by ROR or by CBR depending on what is most efficient at the time.

### 2014-10-21 paper/include/dtail.tex
write raidr

@@
 When the utilization is increased to $90\%$, they increase to $2\%$ and $15\%$, respectively.
+
+Comparing RAIDR with the somewhat similar DTail-R Lie et al.
+does not achieve quite as big decrease in refresh operations as Cui et al.
+does.
+This is due to DTail-R having a higher granularity as the refresh rate is kept per row and not per row group as in RAIDR.
+Even so, a performance improvement of 4.1\% and a decrease in power consumption by 8.3\% is achieved in a system with 32GB DRAM compared to CBR.

### 2014-10-21 paper/include/raidr.tex
write
write raidr

--- 2014-10-09 paper/include/raidr.tex
@@
-RAIDR by J.
-Liu et al.
-\cite{raidr} divides all memory lines into different groups, called bins, based on the cell with the shortest retention time in the line.
-In the basic configuration there are two bins, one with a refresh rate of \(64ms - 128ms\) and one with \(128ms - 256ms\).
-Lines in the first bin gets refreshed every \(64ms\), lines in the second bin every \(128ms\), and lines not in any bin gets refreshed at the default refresh rate which is set to \(256ms\).
-The implementation is done inside the memory controller and to minimize the storage needed for keeping track of bin members they use Bloom filters.
-Thanks to this only 1.25 KB storage is needed for a 32 GB DRAM with the default configuration.
-The storage requirement can however vary depending the amount of bins used.
-RAIDR also accommodate for temperature variations by scaling all bins refresh rates using a timer.
-With RAIDR a performance improvement of 4.1\% and a decrease in power consumption by 8.3\% is achieved in a system with 32GB DRAM compared to auto-refresh.
+In this approach, RAIDR by Liu et al.
+\cite{raidr}, all memory rows gets divided into different groups, called bins, based on the cell with the shortest Retention Time in the row.
+In the basic configuration there are two bins, one with rows that need a refresh rate of $64\:ms \to 128\:ms$ and one with rows that need to be refreshed every $128\:ms \to 256\:ms$.
+Rows in the first bin gets refreshed every $64\:ms$, rows in the second bin every $128\:ms$, and all remaining rows gets refreshed at the default refresh rate which is set to $256\:ms$.
+This is done using two counters, one row counter that rolls over every $64\:ms$, and one counter that keeps track on how many times the row counter have rolled over.
+With this every row will get selected as refresh candidate every $64\:ms$, the refresh will be performed depending on which bin the row belongs to and on the second counter's value.
+The actual refreshes are performed by activating the row in question, essentially performing a ROR.
+
+The implementation is done inside the memory controller meaning RAIDR can not be adopted in currently existing systems.
+To minimize the storage needed for keeping track of bin members Bloom filters is used.
+Thanks to this, only \textit{1.25 KB} storage is needed for a \textit{32 GB} DRAM with the default configuration.
+The storage requirement increases together with the number of bins used, but the granularity, and thereby the decrease in refresh operations, of the approach becomes higher with more bins.
+Other than the amount of bins, the size chosen for the Bloom filters also affects the storage overhead.
+Lie et al.
+evaluates twelve different configurations ranging from one bin with small Bloom filters to three bins with large Bloom filters.
+The later gives the highest decrease in refresh operations but needs \textit{65.25 KB} memory while the first only needs \textit{64 B} but have much smaller impact on the number of refreshes performed.
+
+RAIDR conform to temperature variations by utilizing a third counter that is used to scale all bins refresh rate with the temperature.

### 2014-10-21 paper/include/raidr.tex
hej

@@
 The actual refreshes are performed by activating the row in question, essentially performing a ROR.
+RAIDR conform to temperature variations by utilizing a third counter that is used to scale all bins refresh rate with the temperature.
 
@@
 Thanks to this, only \textit{1.25 KB} storage is needed for a \textit{32 GB} DRAM with the default configuration.
-The storage requirement increases together with the number of bins used, but the granularity, and thereby the decrease in refresh operations, of the approach becomes higher with more bins.
-Other than the amount of bins, the size chosen for the Bloom filters also affects the storage overhead.
+The storage requirement increases together with the number of bins used while the granularity, and thereby the decrease in refresh operations, of the approach becomes higher with more bins.
+As well as the number of bins, the size chosen for the Bloom filters also affects the storage overhead.
 Lie et al.
@@
 The later gives the highest decrease in refresh operations but needs \textit{65.25 KB} memory while the first only needs \textit{64 B} but have much smaller impact on the number of refreshes performed.
-
-RAIDR conform to temperature variations by utilizing a third counter that is used to scale all bins refresh rate with the temperature.
+The default configuration is a trade of between these two extremes.

### 2014-10-21 paper/include/flikker.tex
corrections

@@
 The technique is proposed by Liu et al.
-\cite{flikker} and targets Data Tolerance and Validity.
+\cite{flikker} and targets Data Tolerance.
 Flikker distinguish data as critical or non-critical.
@@
 Flikker refreshes the data by extending the \textit{Partial Array Self Refresh} (PASR) mode which was introduced 2002 in mobile DRAMs.
-PASR mode refreshes a portion of the DRAM array of one bank when the mobile system is in sleep mode, and the technique modifies PASR's internal address counter to support two refresh rates.
+PASR mode refreshes a portion of the DRAM array of one bank when the mobile system is in sleep mode, and the technique modifies PASR's address counter to support two refresh rates.
 

### 2014-10-21 paper/include/raidr.tex
corrections

@@
 In this approach, RAIDR by Liu et al.
-\cite{raidr}, all memory rows gets divided into different groups, called bins, based on the cell with the shortest Retention Time in the row.
+\cite{raidr}, all memory rows get divided into different groups, called bins, based on the cell with the shortest Retention Time in the row.
 In the basic configuration there are two bins, one with rows that need a refresh rate of $64\:ms \to 128\:ms$ and one with rows that need to be refreshed every $128\:ms \to 256\:ms$.
 Rows in the first bin gets refreshed every $64\:ms$, rows in the second bin every $128\:ms$, and all remaining rows gets refreshed at the default refresh rate which is set to $256\:ms$.
-This is done using two counters, one row counter that rolls over every $64\:ms$, and one counter that keeps track on how many times the row counter have rolled over.
-With this every row will get selected as refresh candidate every $64\:ms$, the refresh will be performed depending on which bin the row belongs to and on the second counter's value.
+This is done using two counters, one address counter that rolls over every $64\:ms$, and one counter that keeps track on how many times the address counter have rolled over.
+With this, every row will get selected as refresh candidate every $64\:ms$ and the refresh will be performed depending on which bin the row belongs to and on the second counter's value.
 The actual refreshes are performed by activating the row in question, essentially performing a ROR.
@@
 
-The implementation is done inside the memory controller meaning RAIDR can not be adopted in currently existing systems.
-To minimize the storage needed for keeping track of bin members Bloom filters is used.
-Thanks to this, only \textit{1.25 KB} storage is needed for a \textit{32 GB} DRAM with the default configuration.
-The storage requirement increases together with the number of bins used while the granularity, and thereby the decrease in refresh operations, of the approach becomes higher with more bins.
-As well as the number of bins, the size chosen for the Bloom filters also affects the storage overhead.
-Lie et al.
+The implementation is realized by modifying the MC.
+To minimize the storage needed for keeping track of bin members, Bloom filters are used.
+Thanks to this, only \textit{1.25 KB} storage is needed for a \textit{32 GB} DRAM with the default configuration, which becomes 0.037 ppm.
+The storage requirement increases together with the number of bins used, while the granularity, thereby the decrease in refresh operations, of the technique becomes higher with more bins.
+Similarly to the number of bins, the size chosen for the Bloom filters also affects the storage overhead.
+
+Liu et al.
 evaluates twelve different configurations ranging from one bin with small Bloom filters to three bins with large Bloom filters.
-The later gives the highest decrease in refresh operations but needs \textit{65.25 KB} memory while the first only needs \textit{64 B} but have much smaller impact on the number of refreshes performed.
+The latter gives the highest decrease in refresh operations but needs \textit{65.25 KB} of memory, while the first only needs \textit{64 B} but have much smaller impact on the number of refreshes performed.
 The default configuration is a trade of between these two extremes.

### 2014-10-21 paper/include/sparkk.tex
corrections

@@
 
-The refresh scheme relies on an ordered list of one entry per memory area, which is traversed through by the internal address counter.
+The refresh scheme relies on an ordered list of one entry per memory area, which is traversed through by a address counter.
 Each memory area is formed of any number of consecutive rows and can have different refresh periods for each bank.
@@
 The DRAM device has to provide the ability of per bank refresh commands for this to work.
-
-\todo[inline]{Drafty below}
+By this scheme, Sparkk only refreshes valid data and thus utilize Validity information.
 
@@
 Lucas et al.
-proposes a complex .....
+present a vision where the technique is part of a tailored system for approximate data with specialized languages, instruction set, and hardware.
 
-Results:
+Through stochastic modeling, Sparkk has over $10~dB$ higher peak SNR than Flikker at the same average refresh rate.
+Moreover, Sparkk produces higher peak SNR at half the average refresh rate used for Flikker.
+Experiments regarding energy savings has unfortunately not been performed.

### 2014-10-21 paper/include/dtail.tex
dtail
git write write pull

--- 2014-10-21 paper/include/dtail.tex
@@
 
-When a row is to be scheduled for refresh the refresh data is checked.
-If the row is valid and the time since the last refresh is \(64\:ms \times 2^n, n = [0..7]\) a refresh is performed, otherwise not.
-The refreshes can be done either by ROR or by CBR depending on what is most efficient at the time.
-If a large number of rows need refresh then CBR is selected, otherwise ROR is used.
+When a row is selected for refresh by the memory controller, DTail intercepts the refresh command and based on the Retention Time and Validity information either lets the refresh to be performed or cancels it.
+The method used for refreshing a row depends on what is the most efficient at the time, if a large number of rows need refresh then CBR is selected, otherwise ROR is used.
 
@@
 When the utilization is increased to $90\%$, they increase to $2\%$ and $15\%$, respectively.
+
+Comparing DTail with the somewhat similar RAIDR, Cui et al.
+achieves a bigger decrease in refresh operations compared to what Liu et al.
+does.
+This is due to DTail-R having a higher granularity as the refresh rate is kept per row and not per row group as in RAIDR.

### 2014-10-21 paper/include/dtail.tex
durr

@@
 When a row is selected for refresh by the memory controller, DTail intercepts the refresh command and based on the Retention Time and Validity information either lets the refresh to be performed or cancels it.
-The method used for refreshing a row depends on what is the most efficient at the time, if a large number of rows need refresh then CBR is selected, otherwise ROR is used.
+The command used for refreshing a row depends on what is the most efficient at the time, if a large number of rows need refresh then CBR is selected, otherwise ROR is used.
 

### 2014-10-21 paper/include/rioparis.tex
corrections

--- 2014-10-03 paper/include/rioparis.tex
@@
-RIO is a \textbf{R} based technique proposed by S.
+RIO is a Retention Time based technique proposed by Baek et al.
+\cite{rioparis} in which low retention time cells (bad cells) are identified and `deleted'.
+This is performed by changing the memory allocating modules in the OS to not allocate pages that have bad cells in them.
+Beak et al.
+implement RIO in a Linux kernel which is run and tested on a real hardware system and achieves $87.5\%$ reduction in refresh count with a performance increase of $4.5\%$ on average.
+It was found that aiming for a refresh period longer than $256~ms$ only yields a small performance and power improvement and therefore is the refresh period set to that.
+The implementation mostly relies on modifications to the kernel and there is no need to modify the hardware, so the technique would be easy to adopt in systems today.
+To keep fragmentation of the memory low, a maximum of $0.1\%$ of total memory pages gets `deleted'.
+Even though deletion of a larger amount of pages could decrease the amount of refreshes further it would cause trouble for the kernel which relies on being able to allocate large contiguous address spaces.
+As retention time varies hugely with temperature RIO continuously measures the system temperature and compensate the refresh periods for it.
+
 Baek et al.
-\cite{rioparis} in which they identify cells with low retention time (bad cells) and `deletes' them.
-This is done by changing the memory allocating modules in the OS to not allocate pages that have bad cells in them.
-They implement RIO in a Linux kernel which is run and tested on a real hardware system and is able to achieve 87.5\% reduction in refresh count with a performance increas of 4.5\% on average.
-It is shown that going for a refresh period longer than \(256ms\) only yields a small performance and power improvment and therefore they does not push it higher.
-Their implementaion almost only relies on changes in the kernel and there is no need to modify the hardware so this technique would be easy to adopt in systems today.
-To keep fragmentation of the memory low a maximum of 0.1\% of total memory pages gets `deleted'.
-Even though deletion of a larger amount of pages could decrease the amount of refreshes further it would cause trouble for the kernel which relies on being able to allocate large contiguous address spaces.
-As retentiontime varies hugely with temperature RIO continously measures the system temperature and compensate the refresh persiod for it.
+also propose PARIS \cite{rioparis}, which can be combined with RIO for further improvements.
+In PARIS, the goal is to avoid refreshing unused rows by using Validity information instead of prolonging the refresh rate.
+This is performed by keeping track of all page frames allocated and deallocated by the Linux kernel.
+The page frame data is translated to DRAM row data and kept in a bitmap with one full/empty bit per group of rows.
+A group of rows consist of any number of consecutive global rows.
+Larger groups give a smaller storage overhead for the bitmap, but lowers the refresh granularity.
+The bitmap is stored in the MC and as PARIS utilizes ROR.
 
-S.
-Baek et al.
-also propose PARIS which can be combined with RIO for further improvments.
-In PARIS the goal is not to make the refresh period as long as possible but to not refresh unused memory rows at all using \textbf{V} data.
-This is done by keeping track of all page frames allocated and deallocated by the linux kernel.
-The page frame data is translated to DRAM row data and kept in a bitmap with one full/empty bit per row group.
-A row group consist of rows from several DRAM chips.
-More rows per group gives a smaller storage overhead for the bitmap, but lowers the refresh granluarity.
-
-As these two techniques is orthogonal to each other they can be combined, meaning that PARIS removes unnecessary refreshes and RIO lowers the rate at which refreshes are performed.
-This is similar to what Z.
-Cui et al.
+As mentioned earlier, RIO and PARIS can be combined, which results in PARIS removing unnecessary refreshes and RIO lowering the rate at which the performed refreshes are done.
+This is similar to what Cui et al.
 does with DTail-RV.

### 2014-10-21 paper/include/dtail.tex
wrote

@@
 
-When a row is selected for refresh by the memory controller, DTail intercepts the refresh command and based on the Retention Time and Validity information either lets the refresh to be performed or cancels it.
+When a row is selected for refresh by the memory controller, DTail intercepts the refresh command and based on the Retention Time and Validity information, either lets the refresh to be performed or cancels it.
 The command used for refreshing a row depends on what is the most efficient at the time, if a large number of rows need refresh then CBR is selected, otherwise ROR is used.
+To be able to use CBR, Cui et al.
+proposes a new refresh command that increments a rank's internal address counter, which makes it possible for the internal address counter to be synchronized with MC's address counter.
 
 When only DTail-R is used, Cui et al.
-manages to decrease the number of refreshes with as much as $87.9\%$ and thereby only have $16.7\%$ energy overhead compared to a refresh less 32GB DRAM.
+manages to decrease the number of refreshes with as much as $87.9\%$ and thereby only have $16.7\%$ energy overhead compared to a refresh less \textit{32~GB} DRAM.
 With only DTail-V power consumption and performance overhead increase from $4\%$ and $3\%$ to $90\%$ and $23\%$, respectively, when going from $10\%$ to $90\%$ memory utilization compared to a refresh less 32 GB system.

### 2014-10-21 paper/include/dtail.tex
git push hej

@@
 
-When a row is selected for refresh by the memory controller, DTail intercepts the refresh command and based on the Retention Time and Validity information, either lets the refresh to be performed or cancels it.
-The command used for refreshing a row depends on what is the most efficient at the time, if a large number of rows need refresh then CBR is selected, otherwise ROR is used.
+The number of rows refreshed when a CBR is issued is denoted by Cui et al.
+as a super row.
+When the first row in a super row is selected for refresh, DTail decides if CBR or ROR should be used for all the consecutive rows in the super row.
+If a large number of rows need to be refreshed then CBR is selected, otherwise ROR.
 To be able to use CBR, Cui et al.
 proposes a new refresh command that increments a rank's internal address counter, which makes it possible for the internal address counter to be synchronized with MC's address counter.
+When ROR is used, the row's Retention Time and Validity information is considered; if the row is valid and the time since the last refresh is $n \times 64\:ms$ a refresh is performed.
 

### 2014-10-21 paper/include/background.tex
corrections
started with discussion

@@
 The second is called \textit{CAS before RAS refresh} (CBR).
-In this scheme, the MC has an internal address counter that hold which global row that will be refreshed.
-On a refresh cycle, the global row stated by the internal address counter is refreshed and the counter is then incremented.
-Contrary to ROR, CBR do not need provided addresses on the bus, which results in a lower power consumption and less bus usage.
-% During CBR, a row in each bank is updated, right?
-if so.
-write that in the text.
+In this scheme, each rank has an internal address counter that hold which global row that will be refreshed next.
+On a refresh cycle, a number of global rows are refreshed, beginning with the global row pointed to by the internal address counter, followed by an incrementation of the counter after each refresh of a global row.
+The number of global rows refreshed is dependent on DRAM device capacity.
+Contrary to ROR, CBR do not need provided addresses on the bus, which results in a lower power consumption due to less bus usage.
 

### 2014-10-21 paper/include/reducing_refresh_power.tex
flikker started
reordering of sections
reorder again
started with discussion

@@
 
+\subsubsection*{\textbf{RAIDR}}
+\label{par:raidr}
+\input{include/raidr}
+
 \subsubsection*{\textbf{DTail}}
@@
 \input{include/dtail}
-
-\subsubsection*{\textbf{RAIDR}}
-\label{par:raidr}
-\input{include/raidr}
 
@@
 
+\subsubsection*{\textbf{Flikker}}
+\label{par:flikker}
+\input{include/flikker}
+
 \subsubsection*{\textbf{Sparkk}}
@@
 
-\subsubsection*{\textbf{Flikker}}
-\label{par:flikker}
-\input{include/flikker}
-
-\subsubsection*{\textbf{CREAM}}
-\label{par:cream}
-\input{include/cream}
-
-\subsubsection*{\textbf{Energy Optimization in 3D MPSoCs with Wide-I/O DRAM
-Using Temperature Variation Aware Bank-wise Refresh}}
+\subsubsection*{\textbf{Temperature Variation Aware Bank-wise Refresh - TVABR}}
 \label{par:tempaware}
 \input{include/tempaware}
-
-\subsection{Differences and similarities of the approaches}
-\todo[inline]{Drafty text below!}
-
-Differences and similarities for ARTV approaches.
-
-What in the DRAM devices is modified? / Where is data placed.
-
-How much knowledge does the techniques have about the running software?
-
-Smart Refresh:
-Employed its solution to 2 GB ram.
-What about todays sizes?
-How effective is the technique now?
-- Only works effective if large parts the memory is accessed, i.e.
-large part of the memory has to be valid to begin with.
-Is it so today?
-
-Smart Refresh \& Refrint:
-Refrint is somewhat a extension of Smart Refresh.
-Smart Refresh's features can be laid somewhat equal to Refrint's time-based polices, whereas the data-based polices have no equivalence in Smart Refresh.
-
-They both modify the memory controller, and almost in the same way.
-But Refrint has no timers that is getting decremented the whole time, and instead copies a global "timer" to the line associated bits upon access.
-Of course Refrint's data-based polices also adds logic to the memory controller.
-
-Discuss energy savings. Hard to do eDRAM vs DRAM...

### 2014-10-22 paper/include/discussion.tex
made some placeholders for new subsections
wrote, added TODOs and sources

@@
 
-Talk about scheduling real accesses togheter with refresh ops
+\todo[inline]{Drafty text below!}
+\textbf{Broadly compare the techniques from different approaches}
 
-How do we cope with 3D DRAM?
+\textbf{Advantages and disadvantages for each approach}
+-What is modified? / Where is stuff stored?
+-How much knowledge does the techniques have about the running software?
+-Is the approach's benefits application-depended?
+-PARIS could have used CBR if sREF was implemented (as Dtail does), but this extends DDRx instruction set...
+
+Smart Refresh:
+Employed its solution to 2 GB ram.
+What about todays sizes?
+How effective is the technique now?
+- Only works effective if large parts the memory is accessed, i.e.
+large part of the memory has to be valid to begin with.
+Is it so today?
+
+Smart Refresh \& Refrint:
+Refrint is somewhat a extension of Smart Refresh.
+Smart Refresh's features can be laid somewhat equal to Refrint's time-based polices, whereas the data-based polices have no equivalence in Smart Refresh.
+
+They both modify the memory controller, and almost in the same way.
+But Refrint has no timers that is getting decremented the whole time, and instead copies a global `timer' to the line associated bits upon access.
+Of course Refrint's data-based polices also adds logic to the memory controller.
+
+Discuss energy savings. Hard to do eDRAM vs DRAM...
+
+\textbf{Which techniques can be combined?}
+
+... any of these techniques can be combined with `scheduling real accesses together with refresh ops' (can them???). Mention CREAM as an example.
+
+\textbf{Discuss how 3D DRAM affects everything}
+-Will any approach be benefited / dis-benefited by it? 
+-Which technique seems future proof?

### 2014-10-22 paper/include/rioparis.tex
durr. hata bäsk....

@@
 The implementation mostly relies on modifications to the kernel and there is no need to modify the hardware, so the technique would be easy to adopt in systems today.
+
 To keep fragmentation of the memory low, a maximum of $0.1\%$ of total memory pages gets `deleted'.
 Even though deletion of a larger amount of pages could decrease the amount of refreshes further it would cause trouble for the kernel which relies on being able to allocate large contiguous address spaces.
-As retention time varies hugely with temperature RIO continuously measures the system temperature and compensate the refresh periods for it.
+It would also cause a bigger loss of memory available to the system.
+As retention time varies hugely with temperature, RIO continuously measures the system temperature and compensate the refresh periods for it.
 
@@
 Larger groups give a smaller storage overhead for the bitmap, but lowers the refresh granularity.
-The bitmap is stored in the MC and as PARIS utilizes ROR.
+The bitmap is stored in the MC and the storage overhead is $0.0015\%$ in a system with a \textit{8 KB} row size and one row per group.
+The method used for refresh is ROR.
 
@@
 does with DTail-RV.
-\todo[inline]{compare to DTAIL-RV}
+Cui et al.
+manages to lower the number of refreshes somewhat more than Baek et al.
+does with RIO and PARIS.
+The difference comes from DTail-V performing better than PARIS due to DTail keeping Validity information per row instead of per group of rows.
+RIO and DTail-R performs similar but RIO have a bigger storage overhead.
+DTail requires changes to the MC while RIO and PARIS is implemented in software and thus are more easy to realize.

### 2014-10-22 paper/include/reducing_refresh_power.tex
bla

@@
 Retention time is a measurement of the time it takes before a cell loose its data if not refreshed (recharged).
-Variations in the manufacturing process causes fluctuation in the retention time of the DRAM cells on the same chip.
+Variations in the manufacturing process causes a exponentially distributed fluctuation \cite{katayama} in the retention time of the DRAM cells on the same chip.
 To accommodate for the cells with the lowest retention time manufacturers produce chips with average retention times much higher than possible minimum.
@@
 The retention time is also highly sensitive to temperature variations and becomes drastically lower at higher temperatures.
-For techniques using this approach the biggest advantage is that the improvements are independent of workload, except for the potential rise in temperature.
+For techniques using this approach the biggest advantage is that the improvements are independent of workload, except for a potential rise in temperature which would lower the retention time.
 
@@
 
-\subsubsection*{\textbf{ESKIMO}}
-\label{par:eskimo}
-\input{include/eskimo}
-
 \subsubsection*{\textbf{Flikker}}

### 2014-10-22 paper/include/smart_refresh.tex
disc

@@
 The modifications of the MC constitute of counter update logic, the counter array, and a pending refresh request queue.
-The memory footprint of the modifications is $0.0024\%$ of total device capacity when three-bits counters are used assuming a cache row size of \textit{16~KB}.
+The memory footprint of the modifications is $0.0048\%$ of total device capacity when three-bits counters are used assuming a cache row size of \textit{8~KB}.
 

### 2014-10-22 paper/include/raidr.tex
durr. hata bäsk....
derr

@@
 
-The implementation is realized by modifying the MC.
+The technique is realized by modifying the MC.
 To minimize the storage needed for keeping track of bin members, Bloom filters are used.
@@
 The latter gives the highest decrease in refresh operations but needs \textit{65.25 KB} of memory, while the first only needs \textit{64 B} but have much smaller impact on the number of refreshes performed.
-The default configuration is a trade of between these two extremes.
+The default configuration is a trade of between these two extremes, and with it, a performance improvement of 4.1\% and a decrease in power consumption by 8.3\% is achieved in a \textit{32 GB} system compared to CBR.

### 2014-10-22 paper/include/secret.tex
stuff
secret writelierite

@@
-With SECRET Chung-Hsiang Lin et al.
-\cite{secret} masks bad cells using Error Correction Pointers (ECP).
-The bad cells are maped with a one time profiling done before first system use and all cells that dont meet the retention time requirements gets a ECP.
-A ECP is basiclly a pointer to another cell that is healthy.
-The framwork which is implemented inside the memory controller checks whether there is any valid ECP for every address accessed.
-If there is, it replaces the bad cell(s) data with the data from ECP cell(s) instead.
-This method gives a high energy saving of up to 87.2\% in refresh power and a reduction of 18.57\% in total DRAM power consumption with the RCP cach overhead taken into account.
-The performance impact ranges from 1.3\% overhad att the worst test case to 1.4\% improvment at its best.
-The degredation of performance is due to the added memory access needed and the improvment is for the decreased congestion due to refreshes.
+With SECRET Lin et al.
+\cite{secret} masks cells with short retention time (bad cells) using Error Correction Pointers (ECP).
+The bad cells are mapped with a one time profiling done before first system use and all cells that do not meet the retention time requirements gets a ECP.
+A ECP is basically a pointer to another cell that is healthy.
+The framework, which is implemented inside the memory controller, checks whether there is any valid ECP for the address is accessed.
+If there is, it replaces the cell's data with the data from the cell pointed to by the ECP.
+The ECPs are stored in the DRAM during runtime and a cache in the memory controller is used to mask the delay of fetching the ECPs from DRAM.
+
+The framework divides the DRAM into regions of typically \textit{128 KB}.
+For every one of these regions a \textit{Directory Entry} is kept.
+These directories holds the number of bad cells in the region and a address to where the ECPs are stored.
+As the Directory Entries need to be stored in a contiguous address space and the probability of that existing without any bad cell three copies are stored.
+A majority vote is then performed to mask any bit error in the Directory Entry.
+The size of each entry is $36\:b$ and all entries in a \textit{4 GB} system thus generates a total of $3 \times 36\:b \times \frac{4\:GB}{128\:KB} =$ \textit{432 KB} $= 0.010\%$ storage overhead.
+This is without the space needed for the actual ECPs which requires $21\:b$ each and the total space thereby varies with the number of bad cells.
+
+This approach gives a high energy saving of up to $87.2\%$ in refresh power and a reduction of $18.57\%$ in total DRAM power consumption with the ECP cache overhead taken into account and the refresh rate set to $512\:ms$ instead of the default $64\:ms$.
+The performance impact ranges from $1.3\%$ degradation at the worst to a $1.4\%$ improvement at the best.
+The degradation of performance is due to the added memory access needed to fetch the ECPs, and the performance improvement comes from the decreased congestion due to less refresh operations performed.

### 2014-10-22 paper/include/discussion.tex
disc
wrote & corrections

@@
 
-\textbf{Discuss how 3D DRAM affects everything}
+Another proposed technique is ESKIMO by Isen and John \cite{eskimo}, which focus at reducing total DRAM power through optimizing the accesses to the DRAM.
+By using one bit of a cache row to denote whether it hold nonsense data, it becomes possible to resolve some memory accesses in the caches, which otherwise would have needed to access DRAM.
+A memory region is regarded as nonsense if it has been deallocated, or allocated but not written to.
+The ISA has to be extended to provide the allocation and deallocation information.
+If a block has to be replaced in a write-through cache and the replaced block is dirty but also deallocated, the cache can ignore writing the replaced block to memory.
+Similarly, a write miss to a newly allocated area results in a read access that can also be ignored.
+The latter optimization has support of a structure stored in DRAM which tracks nonsense rows.
+
+The 
+Combining ESKIMO with the selective refresh implementation proposed by Oshawa \cite{oshawa} (ESKIMO[31]), $39\%$ of total DRAM power was saved on average.
+
 -Will any approach be benefited / dis-benefited by it?

### 2014-10-22 paper/include/secret.tex
wrote & corrections

@@
 With SECRET Lin et al.
-\cite{secret} masks cells with short retention time (bad cells) using Error Correction Pointers (ECP).
-The bad cells are mapped with a one time profiling done before first system use and all cells that do not meet the retention time requirements gets a ECP.
-A ECP is basically a pointer to another cell that is healthy.
-The framework, which is implemented inside the memory controller, checks whether there is any valid ECP for the address is accessed.
+\cite{secret} masks cells with short retention time using Error Correction Pointers (ECP).
+The DRAM cells are mapped through profiling, which is performed the first time the system is in use, and all cells that do not meet the retention time requirements is assigned an ECP.
+An ECP is basically a pointer to a cell that met the retention time requirements.
+The framework logic, which is implemented inside the MC, checks whether there is any ECP for the address accessed.
 If there is, it replaces the cell's data with the data from the cell pointed to by the ECP.
-The ECPs are stored in the DRAM during runtime and a cache in the memory controller is used to mask the delay of fetching the ECPs from DRAM.
+The ECPs are stored in DRAM during runtime and are also cached in the MC to mask the delay of fetching them from DRAM.
 
 The framework divides the DRAM into regions of typically \textit{128 KB}.
-For every one of these regions a \textit{Directory Entry} is kept.
-These directories holds the number of bad cells in the region and a address to where the ECPs are stored.
-As the Directory Entries need to be stored in a contiguous address space and the probability of that existing without any bad cell three copies are stored.
-A majority vote is then performed to mask any bit error in the Directory Entry.
+For every region a \textit{Directory Entry} is kept.
+These directories holds the number of bad cells in a region and an address to where the ECPs are stored.
+As the Directory Entries need to be stored in a contiguous address space and the probability of that existing without any bad cell, three copies are stored.
+A majority vote is then performed to mask any bit error in the Directory Entries.
 The size of each entry is $36\:b$ and all entries in a \textit{4 GB} system thus generates a total of $3 \times 36\:b \times \frac{4\:GB}{128\:KB} =$ \textit{432 KB} $= 0.010\%$ storage overhead.
@@
 
-This approach gives a high energy saving of up to $87.2\%$ in refresh power and a reduction of $18.57\%$ in total DRAM power consumption with the ECP cache overhead taken into account and the refresh rate set to $512\:ms$ instead of the default $64\:ms$.
+This technique saves up to $87.2\%$ of refresh power consumption which corresponds to a total reduction of DRAM power consumption of $18.6\%$, when the ECP cache overhead is taken into account and the refresh rate set to $512\:ms$ instead of the default $64\:ms$.
 The performance impact ranges from $1.3\%$ degradation at the worst to a $1.4\%$ improvement at the best.
-The degradation of performance is due to the added memory access needed to fetch the ECPs, and the performance improvement comes from the decreased congestion due to less refresh operations performed.
+The degradation is due to the added memory accesses needed to fetch the ECPs, and the  improvement comes from the decreased DRAM access congestion due to less refresh operations performed.

### 2014-10-22 paper/include/smart_refresh.tex
wrote & corrections

@@
 The modifications of the MC constitute of counter update logic, the counter array, and a pending refresh request queue.
-The memory footprint of the modifications is $0.0048\%$ of total device capacity when three-bits counters are used assuming a cache row size of \textit{8~KB}.
+The memory footprint of the modifications is $0.0048\%$ of total device capacity when three-bits counters are used assuming a DRAM row size of \textit{8~KB}.
 

### 2014-10-22 paper/include/dtail.tex
corrections

@@
 manages to decrease the number of refreshes with as much as $87.9\%$ and thereby only have $16.7\%$ energy overhead compared to a refresh less \textit{32~GB} DRAM.
-With only DTail-V power consumption and performance overhead increase from $4\%$ and $3\%$ to $90\%$ and $23\%$, respectively, when going from $10\%$ to $90\%$ memory utilization compared to a refresh less 32 GB system.
+With only DTail-V power consumption and performance overhead increase from $4\%$ and $3\%$ to $90\%$ and $23\%$, respectively, when going from $10\%$ to $90\%$ memory utilization compared to a refresh less \textit{32~GB} system.
 The power of DTail shows when DTail-R and DTail-V are used together and at $10\%$ memory utilization, a performance overhead of only $1\%$ and close to $0\%$ increase in energy consumption is achieved.
@@
 
-Comparing DTail with the somewhat similar RAIDR, Cui et al.
+Comparing DTail-R with the somewhat similar RAIDR, Cui et al.
 achieves a bigger decrease in refresh operations compared to what Liu et al.

### 2014-10-22 paper/include/discussion.tex
corrections

@@
 \todo[inline]{Drafty text below!}
+
+\textbf{Mention some smart scheduling techniques}
+herpa herpa
+
+CREAM CREAM \cite{cream}
+
+Another proposed technique is ESKIMO by Isen and John \cite{eskimo}, which focus at reducing total DRAM power through optimizing the accesses to the DRAM.
+By using one bit of a cache row to denote whether it hold nonsense data, it becomes possible to resolve some memory accesses in the caches, which otherwise would have needed to access DRAM.
+A memory region is regarded as nonsense if it has been deallocated, or allocated but not written to.
+The ISA has to be extended to provide the allocation and deallocation information.
+If a block has to be replaced in a write-through cache and the replaced block is dirty but also deallocated, the cache can ignore writing the replaced block to memory.
+Similarly, a write miss to a newly allocated area results in a read access that can also be ignored.
+Combining ESKIMO with the selective refresh implementation proposed by Ohsawa \cite{ohsawa} (ESKIMO[31]), $39\%$ of total DRAM power was saved on average.
+
 \textbf{Broadly compare the techniques from different approaches}
@@
 
-... any of these techniques can be combined with `scheduling real accesses together with refresh ops' (can them???). Mention CREAM as an example.
+... any of these techniques can be combined with `scheduling real accesses together with refresh ops' (can them???)
 
-Another proposed technique is ESKIMO by Isen and John \cite{eskimo}, which focus at reducing total DRAM power through optimizing the accesses to the DRAM.
-By using one bit of a cache row to denote whether it hold nonsense data, it becomes possible to resolve some memory accesses in the caches, which otherwise would have needed to access DRAM.
-A memory region is regarded as nonsense if it has been deallocated, or allocated but not written to.
-The ISA has to be extended to provide the allocation and deallocation information.
-If a block has to be replaced in a write-through cache and the replaced block is dirty but also deallocated, the cache can ignore writing the replaced block to memory.
-Similarly, a write miss to a newly allocated area results in a read access that can also be ignored.
-The latter optimization has support of a structure stored in DRAM which tracks nonsense rows.
+\textbf{3D stacked DRAM discussion}
+-Will any approach be benefited / dis-benefited by it?
 
-The 
-Combining ESKIMO with the selective refresh implementation proposed by Oshawa \cite{oshawa} (ESKIMO[31]), $39\%$ of total DRAM power was saved on average.
-
--Will any approach be benefited / dis-benefited by it?
 -Which technique seems future proof?

### 2014-10-22 paper/include/rioparis.tex
discdisc

@@
 RIO is a Retention Time based technique proposed by Baek et al.
-\cite{rioparis} in which low retention time cells (bad cells) are identified and `deleted'.
-This is performed by changing the memory allocating modules in the OS to not allocate pages that have bad cells in them.
+\cite{rioparis} in which low retention time cells are identified and `deleted'.
+This is performed by changing the memory allocating modules in the OS to not allocate pages that have cells with low retention time in them.
 Beak et al.

### 2014-10-22 paper/include/raidr.tex
corrections

--- 2014-10-21 paper/include/raidr.tex
@@
 
-The implementation is realized by modifying the MC.
+The technique is realized by modifying the MC.
 To minimize the storage needed for keeping track of bin members, Bloom filters are used.
-Thanks to this, only \textit{1.25 KB} storage is needed for a \textit{32 GB} DRAM with the default configuration, which becomes 0.037 ppm.
+Thanks to this, only \textit{1.25 KB} storage is needed for a \textit{32 GB} (\textit{8 KB} row size) DRAM with the default configuration, which becomes \textit{0.031\%}.
 The storage requirement increases together with the number of bins used, while the granularity, thereby the decrease in refresh operations, of the technique becomes higher with more bins.
@@
 The latter gives the highest decrease in refresh operations but needs \textit{65.25 KB} of memory, while the first only needs \textit{64 B} but have much smaller impact on the number of refreshes performed.
-The default configuration is a trade of between these two extremes.
+The default configuration is a trade of between these two extremes, and with it, a performance improvement of $4.1\%$, a decrease in power consumption by $8.3\%$, and a total reduction of refresh operations by $74.6\%$, is achieved in a \textit{32 GB} system compared to CBR.

### 2014-10-23 paper/include/rioparis.tex
dtail addition

--- 2014-10-22 paper/include/rioparis.tex
@@
 does with RIO and PARIS.
-The difference comes from DTail-V performing better than PARIS due to DTail keeping Validity information per row instead of per group of rows.
+The difference comes from DTail-V performing better than PARIS both due to DTail keeping Validity information per row instead of per group of rows, and that DTail switches between CBR and ROR which PARIS does not do.
 RIO and DTail-R performs similar but RIO have a bigger storage overhead.

### 2014-10-23 paper/include/background.tex

@@
 
-\todo[inline]{change the next paragraph to introduce retention time.}
 The memory controller (MC) normally issues refresh operations periodically at an interval of \textit{64~ms} for each row, an interval that has been constant for the latest DRAM specifications [source] and called \textit{maximum refresh period}.

### 2014-10-23 paper/include/discussion.tex

@@
 Similarly, a write miss to a newly allocated area results in a read access that can also be ignored.
-Combining ESKIMO with the selective refresh implementation proposed by Ohsawa \cite{ohsawa} (ESKIMO[31]), $39\%$ of total DRAM power was saved on average.
+Combining ESKIMO with the selective refresh implementation proposed by Ohsawa \cite{ohsawa}, $39\%$ of total DRAM power was saved on average.
 
-\textbf{Broadly compare the techniques from different approaches}
+\textbf{Broadly compare the techniques}
+\begin{table*}[t]
+	\caption{\label{tbl:summary}Summarized stuff.}
+        \begin{center}
+			\normalsize
+			\begin{tabular}{ l r r r r r r}
+				\textbf{Technique} & \parbox{2cm}{ \textbf{Refresh \linebreak information*}} & \textbf{Modifies} & \parbox{1.7cm}{ \textbf{Refresh \linebreak reduction}} & \parbox{2.2cm}{ \textbf{DRAM power \linebreak reduction}} & \parbox{1.2cm}{\textbf{Storage \newline Overhead}} & \parbox{1.5cm}{\textbf{Performance impact}} \vspace{0.05cm} \\
+				\hline
+				\textit{Smart Refresh} & A & MC & 59\% & $12.13\%$ & $0.0048\%$  & Unknown  \\
+				\textit{Refrint} & A & MC, (?) & N/A & N/A  & $0.005\%$  & N/A  \\
+				\hline
+				\textit{RAIDR} & R & MC & $74.6\%$ & $8.3\%$  & $0.031\%$  & $4.1\%$  \\
+				\textit{DTail} & V, R & MC, DDRx & $88.2\% \to 98.9\% $ & $< 23\%$  & $0.006\%$ & R:  \\
+				\textit{RIO} & R & OS & $87.5\%$ & Unknown  & $0.1\%$  & $4.5\%$ \\
+				\textit{SECRET} & R & MC & $87.5\%$ & $18.6\%$  & $>0.01\%$  & $\pm 1.4\%$  \\
+				\hline
+				\textit{PARIS} & V & MC & Unknown & $-41.7\%$ DTail-V  & $0.0015\%$  & $-4.4\%$ DTail-V  \\
+				\hline
+				\textit{Flikker} & T & MC, OS, Apps & Unknown & $20\% \to 25\%$  & $0.005\%$  & $-1\%$   \\
+				\textit{Sparkk} & V, T  & MC, OS, Apps &  $+50\%$ Flikker & Unknown & Unknown  & Unknown \\
+				\hline
+			\end{tabular}
+		\end{center}
+\end{table*}
+
+Smart refresh
+Result shows that the refresh reduction was in average 59\%
+and refresh power consumption savings was on average 52\%,
+all without any performance degradation. 12.13\% reduction of total DRAM
+The memory
+footprint of the modifications is 0.0048\% of total device
+capacity when three-bits counters are used assuming a DRAM
+row size of 8 KB.
+
+Refrint
+N/A
+
+Flikker
+The partitioning scheme can potentially lower performance as
+locality of data elements changes, but the found performance
+degradation was 1\% or less. No loss of reliability was ex-
+perienced, and the technique saved between 20-25\% of total
+DRAM power consumption depending on application.
+In simulations, N was two bits and Count
+was 5 bits, giving a total overhead of 0.005\%, with a row size
+of 8 KB.
+
+Sparkk
+Sparkk produces higher peak SNR at half the
+average refresh rate used for Flikker.
+
+Modifies applications:
+Flikker, Sparkk
+
+Modifies OS/runtime system:
+ESKIMO, Flikker, Sparkk.
+
+Modifies MC:
+All but RIO and PARIS.
+
+Only in SW:
+RIO, PARIS.
+
+ROR doesn't work with $<$ 32 GB, cant keep up the req. speed.
 

### 2014-10-23 paper/include/secret.tex

@@
 With SECRET Lin et al.
-\cite{secret} masks cells with short retention time using Error Correction Pointers (ECP).
-The DRAM cells are mapped through profiling, which is performed the first time the system is in use, and all cells that do not meet the retention time requirements is assigned an ECP.
+\cite{secret} masks cells with short retention time (bad cells) using Error Correction Pointers (ECP).
+The DRAM cells are mapped through profiling, which is performed the first time the system is in use, and all bad cells are assigned an ECP.
 An ECP is basically a pointer to a cell that met the retention time requirements.

### 2014-10-23 paper/include/discussion.tex
discussion draft

--- 2014-10-22 paper/include/discussion.tex
@@
 
+\begin{table*}[t]
+	\caption{\label{tbl:summary}Summarized stuff.}
+        \begin{center}
+			\normalsize
+			\begin{tabular}{ l r r r r r r}
+				\textbf{Technique} & \parbox{2cm}{ \textbf{Refresh \linebreak information*}} & \textbf{Modifies} & \parbox{1.7cm}{ \textbf{Refresh \linebreak reduction}} & \parbox{2.2cm}{ \textbf{DRAM power \linebreak reduction}} & \parbox{1.2cm}{\textbf{Storage \newline Overhead}} & \parbox{1.5cm}{\textbf{Performance impact}} \vspace{0.05cm} \\
+				\hline
+				\textit{Smart Refresh} & A & MC & 59\% & $12.13\%$ & $0.0048\%$  & Unknown  \\
+				\textit{Refrint} & A & MC, (?) & N/A & N/A  & $0.005\%$  & N/A  \\
+				\hline
+				\textit{RAIDR} & R & MC & $74.6\%$ & $8.3\%$  & $0.031\%$  & $4.1\%$  \\
+				\textit{DTail-R} & R & MC, DDRx & $87.9\%$ & $\approx 23\%$  & $0.0045\%$ & $\approx$ \textit{RAIDR} \\
+				\textit{RIO} & R & OS & $87.5\%$ & Unknown  & $0.1\%$  & $4.5\%$ \\
+				\textit{SECRET} & R & MC & $87.5\%$ & $18.6\%$  & $\approx 0.01\%$  & $\pm 1.4\%$  \\
+				\hline
+				\textit{DTail-V} & V & MC, DDRx & $\approx 10\% \to 90\% $ & $41.7\%$ \textit{PARIS}  & $0.0015\%$ & $4.4\%$ \textit{PARIS}  \\
+				\textit{PARIS} & V & MC & $\approx 10\% \to 90\%$ & Unknown  & $0.0015\%$  & Unknown \\
+				\hline
+				\textit{Flikker} & T & MC, OS, Apps & Unknown & $20\% \to 25\%$  & $0.005\%$  & $-1\%$   \\
+				\textit{Sparkk} & V, T  & MC, OS, Apps &  $50\%$ \textit{Flikker} & Unknown & Unknown  & Unknown \\
+				\hline
+			\end{tabular}
+		\end{center}
+\end{table*}
+
 \todo[inline]{Drafty text below!}
 
-\textbf{Mention some smart scheduling techniques}
-herpa herpa
+\textbf{Advantages and disadvantages of each approach}
 
-CREAM CREAM \cite{cream}
+First of, we believe that the footprint of each surveyed technique is negligible when DRAM capacity increases.
 
-Another proposed technique is ESKIMO by Isen and John \cite{eskimo}, which focus at reducing total DRAM power through optimizing the accesses to the DRAM.
+Access Recency:
+Has low efficiency if the DRAM has a low access rate per row.
+$->$ scales bad with DRAM capacity (normally). The access rate will then be lower per row.
+When the access rate per DRAM row is high, the techniques are efficient. Provide two examples, Small embedded systems, large shared memory systems with small or no cache and many cores. 
+Access Recency is among the easier approaches to implement as it essentially only needs to track DRAM accesses and count the time between them.
+
+Retention Time:
+High efficiency independent of scaling, the running applications, and memory usage.
+Profiling has to be done somehow at/before first system startup.
+Temperature variations has to be accounted for, which some techniques do.
+All Retention Time techniques require profiling of the DRAM cells, a process that increases post-production test times.
+Other than that, the approach is not much harder to implement than Validity or Access Recency.
+
+Validity:
+Efficiency dependent on memory utilization; the lower the better. Copes well with scaling, as it is more probably that more rows are invalid if the DRAM capacity increases. 
+Validity is somewhat easy to implement, as the information required is easily gathered (validity bit).
+
+Data Tolerance:
+Efficiency dependent on application. If a large amount of data can be refreshed less often but still provide the necessary correctness, then Data Tolerance has a high potential. 
+To implement Data Tolerance based techniques, heavy modifications in several layers are required.
+
+Techniques that relies of ROR can have problems with DRAM capacities above \textit{32~GB}, as it then will be so many rows that need to be refreshed within the maximum refresh period.
+
+3D stacked DRAM will affect all approaches.
+The higher temperature increases the refresh rate and stretches the demands for ROR based techniques to lower the refresh rate in order to keep up.
+The temperature will also be more uneven across the device due to the stacking, which give higher requirements for tracking the variations and adapt to them.
+
+\textbf{Modifications required and future potential}
+
+Techniques which can be adopted right away is .. , which also got a good potential.
+
+little more modifications...
+
+Other techniques require more modifications to be adopted, but also has a higher potential. This and that technique.
+
+Techniques which we do not believe to have a larger future potential are the approaches based on Access Recency, alas it is quite simple to implement, the trend is to increase the DRAM capacity and cache structures, which speaks against the usage of this approach.
+But of course, in systems that does not follow these trends (as the previous mentioned ones), Access Recency can have a chance to be adopted.
+
+\textbf{Which techniques can be combined?}
+Can any of the techniques from different approaches be combined?
+
+... they can all be combined with techniques orthogonal to all approaches, as smart scheduling techniques.
+
+For example has Isen and John proposed the smart scheduling technique ESKIMO \cite{eskimo}, which focus at reducing total DRAM power through optimizing the accesses to the DRAM.
 By using one bit of a cache row to denote whether it hold nonsense data, it becomes possible to resolve some memory accesses in the caches, which otherwise would have needed to access DRAM.
@@
 Similarly, a write miss to a newly allocated area results in a read access that can also be ignored.
-Combining ESKIMO with the selective refresh implementation proposed by Ohsawa \cite{ohsawa} (ESKIMO[31]), $39\%$ of total DRAM power was saved on average.
-
-\textbf{Broadly compare the techniques from different approaches}
-
-\textbf{Advantages and disadvantages for each approach}
--What is modified? / Where is stuff stored?
--How much knowledge does the techniques have about the running software?
--Is the approach's benefits application-depended?
--PARIS could have used CBR if sREF was implemented (as Dtail does), but this extends DDRx instruction set...
-
-Smart Refresh:
-Employed its solution to 2 GB ram.
-What about todays sizes?
-How effective is the technique now?
-- Only works effective if large parts the memory is accessed, i.e.
-large part of the memory has to be valid to begin with.
-Is it so today?
-
-Smart Refresh \& Refrint:
-Refrint is somewhat a extension of Smart Refresh.
-Smart Refresh's features can be laid somewhat equal to Refrint's time-based polices, whereas the data-based polices have no equivalence in Smart Refresh.
-
-They both modify the memory controller, and almost in the same way.
-But Refrint has no timers that is getting decremented the whole time, and instead copies a global `timer' to the line associated bits upon access.
-Of course Refrint's data-based polices also adds logic to the memory controller.
-
-Discuss energy savings. Hard to do eDRAM vs DRAM...
-
-\textbf{Which techniques can be combined?}
-
-... any of these techniques can be combined with `scheduling real accesses together with refresh ops' (can them???)
-
-\textbf{3D stacked DRAM discussion}
--Will any approach be benefited / dis-benefited by it?
-
--Which technique seems future proof?
+Combining ESKIMO with the selective refresh implementation proposed by Ohsawa \cite{ohsawa}, $39\%$ of total DRAM power was saved on average.

### 2014-10-23 paper/include/summary.tex
Changed name on conclussion

\section{Summary} 
\label{sec:sum}
SumSumSUm!! my precious!

### 2014-10-23 paper/include/summary.tex
summary! woop woop

@@
 \label{sec:sum}
-SumSumSUm!! my precious!
+We have surveyed nine different techniques for lowering the number of refresh operations performed in a modern DRAM system.
+Because the portion of power consumption coming from refresh operations in DRAM is high and increasing, these techniques are both topical and highly considerable.
+To give a context to the techniques surveyed the basic operation of DRAM is elaborated on.
+
+All the techniques surveyed fits into one of the approaches based on four types of information, Retention Time, Access Recency, Data Tolerance, and Data Validity, mentioned in the paper.
+The relative performance of the techniques is compared, and some projections for the future concerning higher density and bigger capacity DRAMs are discussed.

### 2014-10-23 paper/include/discussion.tex
discussion draft
wrtie write

@@
 			\begin{tabular}{ l r r r r r r}
-				\textbf{Technique} & \parbox{2cm}{ \textbf{Refresh \linebreak information*}} & \textbf{Modifies} & \parbox{1.7cm}{ \textbf{Refresh \linebreak reduction}} & \parbox{2.2cm}{ \textbf{DRAM power \linebreak reduction}} & \parbox{1.2cm}{\textbf{Storage \newline Overhead}} & \parbox{1.5cm}{\textbf{Performance impact}} \vspace{0.05cm} \\
+				\textbf{Technique} & \parbox{2cm}{ \textbf{Refresh \linebreak information}} & \textbf{Modifies} & \parbox{1.7cm}{ \textbf{Refresh \linebreak reduction}} & \parbox{2.2cm}{ \textbf{DRAM power \linebreak reduction}} & \parbox{1.2cm}{\textbf{Storage \newline Overhead}} & \parbox{1.5cm}{\textbf{Performance impact}} \vspace{0.05cm} \\
 				\hline
@@
 				\hline
-				\textit{DTail-V} & V & MC, DDRx & $\approx 10\% \to 90\% $ & $41.7\%$ \textit{PARIS}  & $0.0015\%$ & $4.4\%$ \textit{PARIS}  \\
-				\textit{PARIS} & V & MC & $\approx 10\% \to 90\%$ & Unknown  & $0.0015\%$  & Unknown \\
+				\textit{DTail-V} & V & MC, DDRx, (OS) & $\approx 10\% \to 90\% $ & $41.7\%$ \textit{PARIS}  & $0.0015\%$ & $4.4\%$ \textit{PARIS}  \\
+				\textit{PARIS} & V & MC, OS & $\approx 10\% \to 90\%$ & Unknown  & $0.0015\%$  & Unknown \\
 				\hline
@@
 
-Techniques which can be adopted right away is .. , which also got a good potential.
+\todo[inline]{describe the table}
+-no area info
 
-little more modifications...
+Techniques which only require OS modifications is easier to adopt than those who need other changes.
+RIO is a good example of a technique which can be adopted after light kernel updates.
+The potential of RIO is significant compared to the modifications needed, as it reduces refreshes to the same degree as R based techniques that need hardware modifications at the cost of higher storage overhead.
+Moreover, all other surveyed techniques require specific support from at least the MC.
 
-Other techniques require more modifications to be adopted, but also has a higher potential. This and that technique.
+Among the techniques which modifies the MC, there are one group that in addition to their logic also have large tables in the MC; Smart Refresh, Refrint, RAIDR, and PARIS.
+Another group stores it's data in the DRAM instead of the MC; SECRET, and DTail.
+The advantage of keeping the refresh data in DRAM is lower MC area overhead and cheaper memory, whereas the downside is that the data has to be cached or prefetched to the MC to mitigate the longer access time of DRAM.
+DTail has additional need of a extension to the DDRx standard, which can complicate the adoption of the technique.
 
-Techniques which we do not believe to have a larger future potential are the approaches based on Access Recency, alas it is quite simple to implement, the trend is to increase the DRAM capacity and cache structures, which speaks against the usage of this approach.
+The techniques that requires the heavies modifications are those based on T.
+Even so, their potential is good and Sparkk's total DRAM power reduction is high, but the changes needed will not be transparent for the application programmer, who has to learn the concept of approximate data in order for the technique to be efficient.
+This is maybe to much to ask, therefor we do believe that T based techniques has lower future potential than R and V based techniques.
+
+We believe that the R and V approaches is the most promising for the future.
+This as techniques based on Access Recency, even though they are quite simple to implement, the trend is to increase the DRAM capacity and cache structures, which speaks against the usage of this approach.
 But of course, in systems that does not follow these trends (as the previous mentioned ones), Access Recency can have a chance to be adopted.
@@
 \textbf{Which techniques can be combined?}
-Can any of the techniques from different approaches be combined?
 
-... they can all be combined with techniques orthogonal to all approaches, as smart scheduling techniques.
+To achieve a larger DRAM power reduction, techniques from different approaches can be combined\footnote{Or you just shut off the computer and enjoy the autumn rain}.
+An example of this is RIO+PARIS, which increases the refresh reduction to $93.8\%$.
+If DTail-RV is used instead, the refresh reduction increases to $98.9\%$.
+DTail could be extended with an arbitrary T data acquisition technique for further improvements.
+All surveyed techniques target to reduce refresh power, but to achieve the greater goal of minimizing total DRAM power, smart scheduling of all DRAM accesses could advantageously be incorporated.
 

### 2014-10-23 paper/include/reducing_refresh_power.tex
wrtie write

@@
 
-\subsubsection*{\textbf{Access Recency}}
+\subsubsection*{\textbf{Access Recency (A)}}
 As mentioned, a DRAM access is performed in the same way as a refresh, which makes it possible to delay the next refresh operation to an accessed row without endangering the stored data.
@@
 
-\subsubsection*{\textbf{Retention Time}}
+\subsubsection*{\textbf{Retention Time (R)}}
 The need for refresh is as mentioned caused by leakage in the memory cells.
@@
 
-%Retention time refers to the period during which cells hold valid values as charge gradually leaks.
-Process variation [10][16] causes the retention time of DRAM cells to vary across the chip.
-Note that approaches that ex-ploit R information are insensitive to system workloads and global memory usage, which makes them attractive omponents for a refresh-optimized memory subsystem.
-
-\subsubsection*{\textbf{Data Tolerance}}
+\subsubsection*{\textbf{Data Tolerance (T)}}
 In the most approaches all data loss is seen as intolerable.
@@
 
-% Applications like games, media processing, machine learning, and unstructured information analysis tolerate errors in portions of their data and still produce acceptably accurate results.
-Approximate computation [31, 8] exploits this approximate data to realize tradeoffs among performance, energy, and accuracy.
-Cells containing such error-tolerant T data need not be refreshed as often as those containing critical data.
-How many cells fall into this category depends on application characteristics.
-
-\subsubsection*{\textbf{Validity}}
+\subsubsection*{\textbf{Validity (V)}}
 In a system that does not utilize all the available memory space it is unnecessary to refresh parts not used.
@@
 This approach is highly sensitive to memory utilization of the system and its effects on energy consumption and performance improvements will be diminishing on systems with high workload.
-
-%If the OS has not allocated the page frame, its data are meaningless, and refreshes to it are wasteful.
-Several software approaches thus attempt to trigger refreshes only for rows with valid data.
-The effectiveness of schemes using such V refresh information is sensitive to the total memory usage of the system.
 
@@
 \input{include/sparkk}
-
-\subsubsection*{\textbf{Temperature Variation Aware Bank-wise Refresh - TVABR}}
-\label{par:tempaware}
-\input{include/tempaware}

### 2014-10-23 paper/include/dtail.tex

@@
 \cite{dtail} propose DTail.
-They evaluate two versions of DTail referenced as DTail-R and DTail-V utilizing Retention Time and Validity information, respectively.
-Even so, their main idea is that more types of refresh information can be incorporated and stored together with the Retention time and Validity information.
+They evaluate two versions of DTail referenced as DTail-R and DTail-V utilizing R and V information, respectively.
+Even so, their main idea is that more types of  information can be incorporated and stored together with the R and V information.
 DTail can then choose the most efficient refresh schema depending on the current situation, i.e.
 workload and type of application.
-Refresh data is kept per row and is stored in the DRAM itself instead of storing it in the MC which many other solutions do \cite{raidr}\cite{smartrefresh}\cite{refrint}.
-To counter the added latency of a slower memory, Cui et al.
-utilizes spatial locality of the refresh data and performs prefetching into a FIFO buffer.
+How acquisition of the \textit{refresh data} is to be performed is left as future work.
+Refresh data is kept per row and is stored in the DRAM itself instead of in the MC which many other solutions do \cite{raidr}\cite{smartrefresh}\cite{refrint}.
+To counter the added latency of a slower memory (DRAM versus fast memory in the MC), Cui et al.
+utilizes spatial locality of the RV information and performs prefetching into a FIFO buffer.
 As the refresh data is kept on a row level DTail can save more refreshes than other techniques that have a lower granularity.

### 2014-10-23 paper/include/rioparis.tex

@@
 The difference comes from DTail-V performing better than PARIS both due to DTail keeping Validity information per row instead of per group of rows, and that DTail switches between CBR and ROR which PARIS does not do.
-RIO and DTail-R performs similar but RIO have a bigger storage overhead.
-DTail requires changes to the MC while RIO and PARIS is implemented in software and thus are more easy to realize.
+Looking at RIO and DTail-R instead, they performs similar but RIO have a bigger storage overhead.
+DTail requires changes to the MC and the DDRx standard whereas a combination of RIO and PARIS would need to modify the MC and the OS.

### 2014-10-23 paper/include/background.tex
disc text

@@
 
-The memory controller (MC) normally issues refresh operations periodically at an interval of \textit{64~ms} for each row, an interval that has been constant for the latest DRAM specifications [source] and called \textit{maximum refresh period}.
+The memory controller (MC) normally issues refresh operations periodically at an interval of \textit{64~ms} for each row, an interval that has been constant for the latest DRAM specifications \cite{ddr}\cite{ddr2}\cite{ddr3} and called \textit{maximum refresh period}.
 This interval is conservatively halved if the DRAM exceeds the specified working temperature of $85^{\circ}C$.

### 2014-10-23 paper/include/discussion.tex
disc text

@@
 
-\begin{table*}[t]
-	\caption{\label{tbl:summary}Summarized stuff.}
-        \begin{center}
-			\normalsize
-			\begin{tabular}{ l r r r r r r}
-				\textbf{Technique} & \parbox{2cm}{ \textbf{Refresh \linebreak information}} & \textbf{Modifies} & \parbox{1.7cm}{ \textbf{Refresh \linebreak reduction}} & \parbox{2.2cm}{ \textbf{DRAM power \linebreak reduction}} & \parbox{1.2cm}{\textbf{Storage \newline Overhead}} & \parbox{1.5cm}{\textbf{Performance impact}} \vspace{0.05cm} \\
-				\hline
-				\textit{Smart Refresh} & A & MC & 59\% & $12.13\%$ & $0.0048\%$  & Unknown  \\
-				\textit{Refrint} & A & MC, (?) & N/A & N/A  & $0.005\%$  & N/A  \\
-				\hline
-				\textit{RAIDR} & R & MC & $74.6\%$ & $8.3\%$  & $0.031\%$  & $4.1\%$  \\
-				\textit{DTail-R} & R & MC, DDRx & $87.9\%$ & $\approx 23\%$  & $0.0045\%$ & $\approx$ \textit{RAIDR} \\
-				\textit{RIO} & R & OS & $87.5\%$ & Unknown  & $0.1\%$  & $4.5\%$ \\
-				\textit{SECRET} & R & MC & $87.5\%$ & $18.6\%$  & $\approx 0.01\%$  & $\pm 1.4\%$  \\
-				\hline
-				\textit{DTail-V} & V & MC, DDRx, (OS) & $\approx 10\% \to 90\% $ & $41.7\%$ \textit{PARIS}  & $0.0015\%$ & $4.4\%$ \textit{PARIS}  \\
-				\textit{PARIS} & V & MC, OS & $\approx 10\% \to 90\%$ & Unknown  & $0.0015\%$  & Unknown \\
-				\hline
-				\textit{Flikker} & T & MC, OS, Apps & Unknown & $20\% \to 25\%$  & $0.005\%$  & $-1\%$   \\
-				\textit{Sparkk} & V, T  & MC, OS, Apps &  $50\%$ \textit{Flikker} & Unknown & Unknown  & Unknown \\
-				\hline
-			\end{tabular}
-		\end{center}
-\end{table*}
+% \begin{table*}[t]
+% 	\caption{\label{tbl:summary}Summarized stuff.}
+%         \begin{center}
+% 			\normalsize
+% 			\begin{tabular}{ l r r r r r r}
+% 				\textbf{Technique} & \parbox{2cm}{ \textbf{Refresh \linebreak information}} & \textbf{Modifies} & \parbox{1.7cm}{ \textbf{Refresh \linebreak reduction}} & \parbox{2.2cm}{ \textbf{DRAM power \linebreak reduction}} & \parbox{1.2cm}{\textbf{Storage \newline Overhead}} & \parbox{1.5cm}{\textbf{Performance impact}} \vspace{0.05cm} \\
+% 				\hline
+% 				\textit{Smart Refresh} & A & MC & 59\% & $12.13\%$ & $0.0048\%$  & Unknown  \\
+% 				\textit{Refrint} & A & MC, (?) & N/A & N/A  & $0.005\%$  & N/A  \\
+% 				\hline
+% 				\textit{RAIDR} & R & MC & $74.6\%$ & $8.3\%$  & $0.031\%$  & $4.1\%$  \\
+% 				\textit{DTail-R} & R & MC, DDRx & $87.9\%$ & $\approx 23\%$  & $0.0045\%$ & $\approx$ \textit{RAIDR} \\
+% 				\textit{RIO} & R & OS & $87.5\%$ & Unknown  & $0.1\%$  & $4.5\%$ \\
+% 				\textit{SECRET} & R & MC & $87.5\%$ & $18.6\%$  & $\approx 0.01\%$  & $\pm 1.4\%$  \\
+% 				\hline
+% 				\textit{DTail-V} & V & MC, DDRx, (OS) & $\approx 10\% \to 90\% $ & $41.7\%$ \textit{PARIS}  & $0.0015\%$ & $4.4\%$ \textit{PARIS}  \\
+% 				\textit{PARIS} & V & MC, OS & $\approx 10\% \to 90\%$ & Unknown  & $0.0015\%$  & Unknown \\
+% 				\hline
+% 				\textit{Flikker} & T & MC, OS, Apps & Unknown & $20\% \to 25\%$  & $0.005\%$  & $-1\%$   \\
+% 				\textit{Sparkk} & V, T  & MC, OS, Apps &  $50\%$ \textit{Flikker} & Unknown & Unknown  & Unknown \\
+% 				\hline
+% 			\end{tabular}
+% 		\end{center}
+% \end{table*}
+In the following section we compare all surveyed techniques more thoroughly and present some of our thoughts on their future potential.
 
-\todo[inline]{Drafty text below!}
+\subsection{Advantages and disadvantages of each approach}
+First of, we believe that the footprint of each surveyed technique is negligible when DRAM capacity increases. This as no technique has a larger storage overhead than $0.1\%$ of DRAM capacity.
 
-\textbf{Advantages and disadvantages of each approach}
+%\subsubsection*{Access Recency} 
+Access Recency techniques becomes less effective when DRAM capacity scales, as it is probable that the access rate per row will decrease.
+Among the approaches, this is one of the easier to implement, as it essentially only needs to track DRAM accesses and count the time between them.
+Techniques based on the other approaches can achieve the same or better results even when the access rate is low.
 
-First of, we believe that the footprint of each surveyed technique is negligible when DRAM capacity increases.
+%\subsubsection*{Retention Time}
+Using R as the refresh data gives a high efficiency independent of DRAM capacity, running applications, and memory utilization.
+A downside is that retention time profiling of the DRAM has to be performed in conjunction with first system startup.
+As retention time is highly dependent on temperature variations and has to be accounted for, which some techniques do; RAIDR, and RIO.
 
-Access Recency:
-Has low efficiency if the DRAM has a low access rate per row.
-$->$ scales bad with DRAM capacity (normally). The access rate will then be lower per row.
-When the access rate per DRAM row is high, the techniques are efficient. Provide two examples, Small embedded systems, large shared memory systems with small or no cache and many cores. 
-Access Recency is among the easier approaches to implement as it essentially only needs to track DRAM accesses and count the time between them.
+%\subsubsection*{Validity} 
+The V approach scales well, as it is more probably that more rows becomes invalid when DRAM capacity increases.
+It also produce high efficiency in many common systems.
+However, support on a OS level is needed to gather the V data, which increases the complexity.
 
-Retention Time:
-High efficiency independent of scaling, the running applications, and memory usage.
-Profiling has to be done somehow at/before first system startup.
-Temperature variations has to be accounted for, which some techniques do.
-All Retention Time techniques require profiling of the DRAM cells, a process that increases post-production test times.
-Other than that, the approach is not much harder to implement than Validity or Access Recency.
+%\subsubsection*{Data Tolerance} 
+The efficiency of T based techniques is dependent of memory utilization besides of data with low criticality, as T do not target invalid memory areas.
+Using T information can yield a high decrease of refresh operations, but heavy modifications in several layers are required.
 
-Validity:
-Efficiency dependent on memory utilization; the lower the better. Copes well with scaling, as it is more probably that more rows are invalid if the DRAM capacity increases. 
-Validity is somewhat easy to implement, as the information required is easily gathered (validity bit).
+\subsection{Modifications required and future potential}
 
-Data Tolerance:
-Efficiency dependent on application. If a large amount of data can be refreshed less often but still provide the necessary correctness, then Data Tolerance has a high potential. 
-To implement Data Tolerance based techniques, heavy modifications in several layers are required.
+%\todo[inline]{describe the table}
+%-no area info
 
-Techniques that relies of ROR can have problems with DRAM capacities above \textit{32~GB}, as it then will be so many rows that need to be refreshed within the maximum refresh period.
-
-3D stacked DRAM will affect all approaches.
-The higher temperature increases the refresh rate and stretches the demands for ROR based techniques to lower the refresh rate in order to keep up.
-The temperature will also be more uneven across the device due to the stacking, which give higher requirements for tracking the variations and adapt to them.
-
-\textbf{Modifications required and future potential}
-
-\todo[inline]{describe the table}
--no area info
+The method used for refresh can affect the possibility of a technique to scale.
+Techniques that relies on ROR can have problems with large DRAM capacities.
+This as the number of rows will be to large for ROR to refresh within the maximum refresh period.
+One example of this is PARIS, which fails to work on a system with \textit{32~GB} of memory \cite{dtail}.
 
@@
 We believe that the R and V approaches is the most promising for the future.
-This as techniques based on Access Recency, even though they are quite simple to implement, the trend is to increase the DRAM capacity and cache structures, which speaks against the usage of this approach.
-But of course, in systems that does not follow these trends (as the previous mentioned ones), Access Recency can have a chance to be adopted.
+This as techniques based on A, even though they are quite simple to implement, the trend is to increase the DRAM capacity and cache structures, which speaks against the usage of this approach.
+But of course, in systems that does not follow these trends, Access Recency can have a chance to be adopted.
+Such systems could for example be small embedded systems or large shared memory systems with small or no cache and many cores.
 
-\textbf{Which techniques can be combined?}
+3D stacked DRAM will affect all approaches.
+The higher temperature increases the refresh rate and stretches the demands for techniques using ROR to lower the refresh rate in order to keep up.
+The temperature will also be more uneven across the device due to the stacking, which give higher requirements for tracking the variations and adapt to them.
+One technique not surveyed in this paper, proposed by Sadri et al.
+\cite{tempaware}, tracks the temperature in different regions of a 3D DRAM and adapts the refresh rate accordingly in these areas.
+
+\subsection{Combining techniques}
 

### 2014-10-29 excersice 2/main.tex
ex 2.

\documentclass[a4paper]{article}

\renewcommand{\labelenumi}{(\alph{enumi})}

\usepackage{listings}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{fancyhdr}
\usepackage[margin=1in]{geometry}
\usepackage{color}
\renewcommand{\labelenumi}{(\alph{enumi})}
\pagestyle{fancy}

\lhead{Chalmers - EDA282}
\chead{Excercise 2}
\rhead{Dan Larsson}
\lfoot{Oktober 2014}

\begin{document}
%\textcolor{red}{\huge Changed stuff marked in red.}

\section*{Q1}
\large{\textbf{What is Amdahl’s law? (give the equation, but then explain in [your own]     words) 2 points\\}}
$S=\frac{1}{B + \frac{1}{n}\times(1-B)}$, $S$ is the speedup, $n$ the number of threads, $B$ the fraction of the program performed serially. Amdahl’s law is used to calculate the speedup gained from parallelization of a task, compared to the serial version. It can also be used for calculating the total speedup on a system from improving a subsystem.

\section*{Q2}
\large{\textbf{What is perfect scaling? Why do we fail to achieve perfect scaling     by doubling (a) the number of transistors, (b) the number of functional units, (c) the number of cores, or (d) the clock speed? Give a sentence or two of explanation for each of (a)-(d). 5 points\\}}
Com up with some shiet

\section*{Q3}
\large{\textbf{
What is a CHARM architecture? What two coordination mechanisms are needed in CHARM architectures? 5 points\\}}
Coordinated homogeneous arrays of RISC microprocessors (CHARM), a SIMD system with locally distributed memory, ...
%See architectural overview in hillis_SIMD_scalable_supercomputer for the coordination.

\section*{Q4}
\large{\textbf{
What family of processor architectures has been used in both data parallel and thread parallel machines? (not the same version) 2 points
The Connection Machine processor family. \\}}
%See hillis_SIMD_data_parallel_algorithms uses data parallelism, starting with the paragraph ending page 1.
%hillis_SIMD_scalable_supercomputer (CM-5) uses thread parallelism.

\section*{Q5}
\large{\textbf{
Why was the cache ineffective in the CM-5? 2 points\\}}
%Due to little spatial locality. page 3 right column in hillis_SIMD_scalable_supercomputer.

\section*{Q6}
\large{\textbf{
How does simultaneous multithreading (SMT) make a single core appear to the software to be multiple cores running multiple threads? 5 points\\}}
The processor keeps state for several threads and can dispatch instructions for the different threads interleaved. One thread is the main thread and gets the most execution time, but when the process running on that thread blocks, the other threads can execute and make some progress.

\section*{Q7}
\large{\textbf{
In an architecture that supports SMT, some resources are shared     completely, some are partitioned to limit each thread’s usage, and     some are duplicated to provide private copies. Indicate S, P, or D     and include an explanation (e.g., why or how). For shared and     partitioned resources, say how many logical threads can share them. (23 points -- 1 point for each S/P/D plus explanation) for the \\}}

\textbf{Intel Pentium:} two Hyperthreads.
\begin{enumerate}
\item    
\textbf{trace cache (the special instruction cache holding sequences of decoded uops)\\}
S - Shared by the logical processors. Stores decoded instructions.
\item
\textbf{ucode ROM\\}
S - Shared by the logical processors. Used to store complex IA-32 instructions.
\item
\textbf{instruction TLB (iTLB)\\}
D - Translates instructions addresses to physical addresses on TC misses.
\item
\textbf{instruction fetch unit\\}
    S - Shared by the logical processors.
\item
\textbf{instruction decoder logic\\}
S and P - The regular uop decoder is shared while the IA-32 decoder is partitioned.
\item
\textbf{instruction queues (memory op queue and “all other” op queue)\\}
P - Made sure so that not one thread occupy all entries except for when running self.
\item
\textbf{retirement logic\\}
S - Shared by the logical processors.
\item
\textbf{uop schedulers\\}
S - Shared by the logical processors. The op buffer is however partitioned so one logical processor can not occupy all places.
\item
\textbf{general purpose architected registers\\}
D - Each logical processor can utilize all architecture registers.
\item
\textbf{Register Alias Table (RAT) (used for expanding the few architected registers to which the ISA refers, mapping them onto a much larger physical register file)\\}
D - Each logical processor need to keep track of the complete architectural state.    
\item
\textbf{physical registers\\}
S - Register renaming makes sure each processor gets it’s own register.
\item
\textbf{branch predictor’s Return Stack Buffer (RBS)\\}
    D - Better predictions if done per thread.
\item
\textbf{branch predictor’s global history array\\}
    S - Entries tagged with logical processor id.
\item
\textbf{data cache (shared)\\}
S - Shared by the logical processors.
\item
\textbf{interrupt controller\\}
D - for each logical processor as it handles it’s own interrupts.

\textbf{Sun rock:}
    See RockIEEEmicro
\item
\textbf{instruction fetch unit\\}
\item
\textbf{instruction decoder logic\\}
\item
\textbf{instruction TLB (iTLB)\\}
\item
\textbf{retirement logic\\}
\item
\textbf{data cache\\}
\item
\textbf{FPU\\}

\textbf{Sun Niagara:}
    %see niagra_IEEE_MICRO
\item
\textbf{data TLB\\}
\item
\textbf{store buffers\\}
\end{enumerate}
\section*{Q8}
\large{\textbf{
How many threads can run on one physical Intel Pentium core? On one Sun Rock core? On one Niagara Sparc pipe? 3 points\\}}
The intel Pentium can run two Hyperthreads. ee papers from the Q above.

\section*{Q9}
\large{\textbf{
Name the three processors we’ve talked about (in papers or in lecture slides) that support Transactional Memory in hardware. Which ones are available in commercial products? 3 points\\}}
First commercial processors with TM support:
Rock (Sun)(2010) first design to include TM, but cancelled
BlueGeneQ (2011) some form of HTM, custom made by IBM
%Haswell, see SC13-TSX and haswell_analysis

\section*{Q10}
\large{\textbf{
Name one     advantage and one disadvantage of having SIMD ISA extension instructions and FP instructions share resources. Explain your answers. Which specific resources are shared in the Intel     Pentium/MMX architecture? In the Sun Rock/Vis architecture? 5 points\\}}

%See RockIEEEmicro and marr_hyperthreading.

\section*{Q11}
\large{\textbf{
What is a data race? What is atomicity? What is sequential     consistency? 3 points\\}}

Data race = see page 3 (upper right corner) in p48-boehm-adve-mem-model.
mutual exclusion = atomicity
SC = Lecture 4 slide 25

\section*{Q12}
\large{\textbf{
%Assume x is initially 999. Explain how, when executed by two threads simultaneously, the following code can result in x holding either 1 or 2000. 10 points// x is number variable represented in a way that requires two words// on the target machine. Assume x is stored in two parts, x_hi and x_lo,// each holding three decimal digits.// you can view highword() and lowword() as #define’s in C (syntactic sugar)// to access the hi and lo parts of a number variable// incr() does whatever is required to add one to its number variable argumentreg1 = highword(x);reg2 = lowword(x);x = incr(x);highword(x) = reg1;lowword(x) = reg2;    see page 3 in p48-boehm-adve-mem-model.
\\}}

\section*{Q13}
\large{\textbf{
Explain the difference between write invalidate and write update cache coherence protocols. 5 points\\}}
When using the Write-Invalidate CCP, a memory write to a shared address will cause invalidations to be sent to all caches holding that data. When using Write Update CCP, instead of invalidating the value in all the other caches, the new value will be distributed to update their values.

\section*{Q14}
\large{\textbf{
What are the possible cache line states in the Blue Gene/P coherence protocol? 2 points\\}}
The blue Gene/P uses write-through for both L1 and L2 and does not have any hardware CCP the cache lines can be either Valid or Invalid (Two states).

\section*{Q15}
\large{\textbf{
How does a snoop filter improve performance and save energy? (be specific about which coherence commands it filters when) 5 points\\}}
A snoop filter removes invalidates and snoops for a cache line that is not in the cache, thus removing unnecessary snoops and invalidates. The improvements gained by a snoop filter depends on both the type of filter used and of the cache it is used with. For a destination based snoop filter there is small energy savings and if the cache is multi-ported, snoops and invalidates does not have to compete with regular accesses, so the performance gain is also minimal. This as the snoops and invalidates still traverse the bus draining energy and causing congestion and the added hardware for the filter consumes energy. For a source based snoop filter the commands will not need to traverse the bus and will thus save more energy and cause less congestion (performance up).

\section*{Q16}
\large{\textbf{
What limitation of the BG/P L1 cache makes the snoop filter critical to performance? 2 points\\}}
The L1 cache in the BG/P only have one access port meaning that snoops and invalidates would compete with regular accesses. THis together with the caches being write-through and invalidations being sent to all caches every time a value is written makes the snoop filters very important.

\section*{Q17}
\large{\textbf{
Could a snoop filter be used to improve performance/efficiency of a bus-based write-update protocol? How, or why not? 2 points\\}}
With a source based snoop-filter, snoops for data that does not exist in a cache is removed and never put on the bus. This will therefore lower the traffic on the bus and thus increase performance. Also, in a single access-port memory the snoops would compete with the normal accesses and removing unnecessary snoops would therefore decrease congestion of the port and increase performance.

\section*{Q18}
\large{\textbf{
What is dimension-order routing? When/why is it used? 2 points\\}}
Dimension order routing is when a message is routed through a system one dimension at a time. If node N(0,0) sends a packet to N(10,7) the packet will go to node N(10,0) and then continue on to N(10,7). The same applies in a cubic interconnect, but simply with one more dimension.

\section*{Q19}
\large{\textbf{
What is a flit (in general, not in the context of any specific     network)? Is it different from a packet? Explain. 2 points\\}}
Flow control unit, flit, is the smallest part of a packet that get transmitted on a interconnect. A flit gets dropped in the case of congestion and thus has to be retransmitted.

\section*{Q20}
What network topology was used in the Connection Machine CM-5? Why? 5 points

%hillis_SIMD_scalable_supercomputer
Fat tree (figure 4).

\section*{Q21}
\large{\textbf{
In the Stanford DASH system, bus bandwidth imposed a greater limit on remote memory bandwidth than the network bisection bandwidth. Explain a) what this means (e.g., what’s the bisection     bandwidth?), and b) why was this the case? 5 points \\}}
Bisection bandwidth is the bandwidth through the path in a interconnect split into two parts at most narrow point. E.g. in a tree interconnect the bisection bandwidth would be 1u as we would cut the tree at the root were only one vertex exist. In DASH each nodes memory bus is connected to a global bus, and when a remote request is issued the local bus will be traversed twice (local->global->local) and thus the traffic on the local bus will be twice as high. This means that the local bandwidth will be more important than the bisection brandwidth.

\section*{Q22}
\large{\textbf{
In what types of application areas are VLIW processors employed today? 2 points\\}}
In processing of digital signals (DSPs) and multimedia.

\end{document}

### 2014-10-29 excersice 2/main.tex
asd

@@
 5 points\\}}
-
+One advantage is that there is no need for adding additional hardware for keeping the architectural state and thus simplifies the design, while as the drawback is that you can not us FP instructions in the same routine as the added instructions and corruption of the data can occur.
+For the MMX architecture, the added registers maps onto the FP registers meaning these are shared between the two.
 %See RockIEEEmicro and marr_hyperthreading.
@@
 \section*{Q20}
-What network topology was used in the Connection Machine CM-5? Why? 5 points
+\large{\textbf{What network topology was used in the Connection Machine CM-5? Why? 5 points\\}}
 
@@
 \section*{Q22}
-\large{\textbf{
-In what types of application areas are VLIW processors employed today? 2 points\\}}
+\large{\textbf{In what types of application areas are VLIW processors employed today? 2 points\\}}
 In processing of digital signals (DSPs) and multimedia.

### 2014-10-30 lab/report/include/lab1.tex
revision_1

@@
 
-In the producer consumer benchmark, the E state is beneficial if the writing thread accesses the data element for the first time or when it has been fetched from higher level caches.
-Then, the thread will transition the data element from I to M through E, when performing an increment.
+\textcolor{red}{In the producer consumer benchmark, the E state is beneficial if the writing thread accesses the data element before any other thread, during one iteration when the data element has been fetched from higher level caches.
+This happens on the first iteration and on every iteration when the element does not reside in cache due to an earlier conviction.
+If the writing thread succeed with accessing the data element first, it will transition the data element from state I to M through E, when performing an increment.
 This avoids the S state, which give an increased performance as no bus command has to be issued.
 However, the probability that the writing thread accesses the data first is low, $\frac{1}{nthreads}$, which is shown in \reffig{fig:lab1bars} where MSI and MESI seems to perform equally.
-As the working set size increases, the benefit of the E state increases due to more frequent reads from globally invalid cache lines.
-This can however not be seen in the simulation results, due to fluctuations.
+As the working set size increases, the benefit of the E state magnifies due to that the data elements more frequently will be convicted from cache before they are accessed again.
+The benefit can, however, not be seen in the simulation results, due to fluctuations.}
 
@@
 This makes MSI issue an invalidate broadcast for each increment and the accumulated overhead is easily spotted in \reffig{fig:lab1bars}.
+\textcolor{red}{Table \ref{tbl:lab1:task3} express the overhead in more detail when the work set size increase, where MSI's write misses and invalidation commands increase as much as MESI's write hits.}
 
@@
 The producer consumer benchmark benefit from the S state, and therefor produces MESI better results than MESI-MG.
-In the benchmark, one thread update an element whereas the other threads read it and by migrating the element from one private cache to the other is less optimal than letting the reading threads share the element and read it in parallel.
-This overhead will be a smaller part of the total cycles as the working set increases, as the misses increase in the private caches and also in the LLC, which is costly in terms of cycles.
+In the benchmark, one thread update an element whereas the other threads read it and migrating the element from one private cache to the other is less optimal than letting the reading threads share the element and read it in parallel.
+This overhead becomes a smaller part of the total cycles as the working set increases, because the misses increase in the private caches and also in the LLC, which is costly in terms of cycles.
 This trend is shown in \reffig{fig:lab1bars}.

### 2014-10-30 excersice 2/main.tex
stuff

--- 2014-10-29 excersice 2/main.tex
@@
 \begin{document}
+\section*{}
+I have discussed my solutions with Jonas Hemlin.
 %\textcolor{red}{\huge Changed stuff marked in red.}
@@
 5 points\\}}
-Com up with some shiet
+Amdahl's law can be used for all these cases and it would show the actual speedup gained.
+\begin{enumerate}
+\item
+One have to know what to do with all the new transistors.
+Simply adding more lines to the caches or widening the memory bus will double the performance as all parts in the system have to be improved on.
+\item
+The system will not be able to utilize all FU at the same time without also beefing up the units feeding them with data. 
+\item
+As the code that can be distributed for parallel execution in a program is not 100\%, the gain from increasing the number of cores will diminish.  
+\item
+When increasing just the clock speed, long latency operations like cache misses and I/O does not gain any speed improvements and thereby holds the system back.
+\end{enumerate}
 
@@
 What is a CHARM architecture? What two coordination mechanisms are needed in CHARM architectures? 5 points\\}}
-Coordinated homogeneous arrays of RISC microprocessors (CHARM), a SIMD system with locally distributed memory, ...
-%See architectural overview in hillis_SIMD_scalable_supercomputer for the coordination.
+CHARM stands for Coordinated Homogeneous Arrays of RISC Microprocessors and is a collection, ranging from a few to thousands, of simple processors working in parallel.
+They communicate by fast networks which scales well with the number of processors.
+The two coordination mechanisms needed in a CHARM architecture is, a low latency interconnect allowing processors to access other processors memories and data, and a mechanism for keeping global state between the processors in an efficient way.
 
@@
 \large{\textbf{
-What family of processor architectures has been used in both data parallel and thread parallel machines? (not the same version) 2 points
-The Connection Machine processor family. \\}}
-%See hillis_SIMD_data_parallel_algorithms uses data parallelism, starting with the paragraph ending page 1.
-%hillis_SIMD_scalable_supercomputer (CM-5) uses thread parallelism.
+What family of processor architectures has been used in both data parallel and thread parallel machines? (not the same version) 2 points\\}}
+The CM-5 processor architecture is thread parallel and data parallel, while the previous CM machines only is data parallel. Both belongs to the Connection Machine processor family by TMC.
 
@@
 Why was the cache ineffective in the CM-5? 2 points\\}}
-%Due to little spatial locality. page 3 right column in hillis_SIMD_scalable_supercomputer.
+Small RISC processors depend on caches to bridge the speed gap of the memory and the processor, but in most of the applications running on a CM-5 almost all of main memory is accessed during one program iteration meaning that the  data in the cache will be written over and the hit rate of the cache will be low.
 
@@
 The processor keeps state for several threads and can dispatch instructions for the different threads interleaved.
-One thread is the main thread and gets the most execution time, but when the process running on that thread blocks, the other threads can execute and make some progress.
+When a thread starts a long latency operation, the other threads can execute and make progress.
+The threads are seen as processors on their own by the OS, meaning that it simply schedules processes for them as it would in as system with many actual processors or singled threaded cores.
 
@@
 \textbf{ucode ROM\\}
-S - Shared by the logical processors. Used to store complex IA-32 instructions.
+S - Shared by the threads (logical processors). Used to store complex IA-32 instructions.
 \item
@@
 \textbf{instruction fetch unit\\}
-    S - Shared by the logical processors.
+S - Shared by the threads (logical processors).
 \item
@@
 \textbf{retirement logic\\}
-S - Shared by the logical processors.
+S - Shared by the threads (logical processors).
 \item
 \textbf{uop schedulers\\}
-S - Shared by the logical processors. The op buffer is however partitioned so one logical processor can not occupy all places.
+S - Shared by the threads (logical processors). The op buffer is however partitioned so one thread can not occupy all places.
 \item
 \textbf{general purpose architected registers\\}
-D - Each logical processor can utilize all architecture registers.
+D - Each thread (logical processor) can utilize all architecture registers.
 \item
 \textbf{Register Alias Table (RAT) (used for expanding the few architected registers to which the ISA refers, mapping them onto a much larger physical register file)\\}
-D - Each logical processor need to keep track of the complete architectural state.    
+D - Each thread (logical processor) keeps track of the complete architectural state.    
 \item
@@
 \textbf{branch predictor’s global history array\\}
-    S - Entries tagged with logical processor id.
+S - Entries tagged with thread (logical processor) id.
 \item
 \textbf{data cache (shared)\\}
-S - Shared by the logical processors.
+S - Shared by the threads (logical processors).
 \item
 \textbf{interrupt controller\\}
-D - for each logical processor as it handles it’s own interrupts.
+D - for each thread(logical processor) as it handles it’s own interrupts.
 
-\textbf{Sun rock:}
-    See RockIEEEmicro
+\textbf{Sun rock:} Two threads (16 cores, 32 threads per chip).  
 \item
 \textbf{instruction fetch unit\\}
+P - The IFU is shared among four cores, which also means that all threads in those cores share the IFU to (8 threads).
+The IFU fetches instruction in a round robin fashion meaning that no one thread can use all resources.
+
 \item
 \textbf{instruction decoder logic\\}
+P - One Decoder in each pipeline, which there are one per core (2 threads) and the instruction is decoded in a round robin fashion.
+
 \item
 \textbf{instruction TLB (iTLB)\\}
+S - There is one iTLB per iCache, and there is one iCache in each IFU, meaning four cores(8 threads) share one iTLB.
+
 \item
 \textbf{retirement logic\\}
+S - There is retirement logic in pipeline, which there are one for each core (2 threads).
+
 \item
 \textbf{data cache\\}
+S - One D-Cache is shared between two cores (4 threads), but each core have it's own tag array(2 threads).
+
 \item
 \textbf{FPU\\}
+S - One FPU is shared between two cores (4 threads).
 
-\textbf{Sun Niagara:}
-    %see niagra_IEEE_MICRO
+\textbf{Sun Niagara:} Four threads per Sparc Pipe (eight Sparc Pipes per chip, 32 threads).
 \item
 \textbf{data TLB\\}
+S - Each Sparc Pipe have it's own DTLB. (4 threads)
+
 \item
 \textbf{store buffers\\}
+D - Each thread have its own store buffers. The buffers is used to check for RAW hazards.
+
 \end{enumerate}
@@
 How many threads can run on one physical Intel Pentium core? On one Sun Rock core? On one Niagara Sparc pipe? 3 points\\}}
-The intel Pentium can run two Hyperthreads. ee papers from the Q above.
+The Intel Pentium can run two threads, Sun Rock two threads, and Sparc Niagara four threads per Sparc Pipe.
 
@@
 Name the three processors we’ve talked about (in papers or in lecture slides) that support Transactional Memory in hardware. Which ones are available in commercial products? 3 points\\}}
-First commercial processors with TM support:
-Rock (Sun)(2010) first design to include TM, but cancelled
-BlueGeneQ (2011) some form of HTM, custom made by IBM
-%Haswell, see SC13-TSX and haswell_analysis
+The Intel's Haswell uses HTM and is commercially available.
+Sun's Rock processor include HTM and was commercially available but got canceled.
+BlueGeneQ made by IBM is part of Top 500 supercomputers, thus commercial, and uses HTM.
 
@@
 5 points\\}}
-
-%See RockIEEEmicro and marr_hyperthreading.
+One advantage is that there is no need for adding additional hardware for keeping the architectural state and thus simplifies the design and saves hardware, while as the drawback is that you can not us FP instructions in the same routine as the added instructions without corruption of the data occurring.
+For the MMX architecture, the added registers maps onto the FP registers, meaning these are shared between the two.
+It is the registers that are shared in Sun Vis to.
 
@@
 What is a data race? What is atomicity? What is sequential consistency? 3 points\\}}
-
-Data race = see page 3 (upper right corner) in p48-boehm-adve-mem-model.
-mutual exclusion = atomicity
-SC = Lecture 4 slide 25
+Data race = When two or more threads accesses and writes to the same address without proper synchronization a data race arises.
+atomicity = A operation is performed mutually exclusive, that is, the operation is performed without any disturbance.  
+SC = Sequential Consistency is enforced if the end result of a parallel execution becomes the same as if the operations had been performed sequentially.
 
@@
 \large{\textbf{
-%Assume x is initially 999.
-Explain how, when executed by two threads simultaneously, the following code can result in x holding either 1 or 2000.
-10 points// x is number variable represented in a way that requires two words// on the target machine.
-Assume x is stored in two parts, x_hi and x_lo,// each holding three decimal digits.// you can view highword() and lowword() as #define’s in C (syntactic sugar)// to access the hi and lo parts of a number variable// incr() does whatever is required to add one to its number variable argumentreg1 = highword(x);reg2 = lowword(x);x = incr(x);highword(x) = reg1;lowword(x) = reg2;    see page 3 in p48-boehm-adve-mem-model.
-\\}}
+Assume x is initially 999. Explain how, when executed by two threads simultaneously, the following code can result in x holding either 1 or 2000. 10 points\\}}
+\textit{x is number variable represented in a way that requires two words on the target machine.
+Assume x is stored in two parts, x\_hi and x\_lo, each holding three decimal digits.
+you can view highword() and lowword() as \#define’s in C (syntactic sugar) to access the hi and lo parts of a number variable incr() does whatever is required to add one to its number variable argument.}
+\lstset{language=C,frame=single}
+\begin{lstlisting}
+reg1 = highword(x);
+reg2 = lowword(x);
+x = incr(x);
+highword(x) = reg1;
+lowword(x) = reg2; 
+\end{lstlisting}
+The first case, x becomes 1, happens when one thread first reads the high word equal to 0, and then gets context switched.
+The other thread then completes it's code and increases x to 1000.
+When the first thread then executes again it reads the low word, now equal to 0 instead of 999, increases it to 1, and writes it back.
+
+For the second case, almost the same thing happens, the first thread executes until it has stored the high word of x increased to 1, gets context switched, and the other thread reads 1999 instead of 999.
+Thread two then increments 1999 to 2000, gets switched out for the first thread that stores 1000, gets switched back and stores 2000.
 
@@
 What are the possible cache line states in the Blue Gene/P coherence protocol? 2 points\\}}
-The blue Gene/P uses write-through for both L1 and L2 and does not have any hardware CCP the cache lines can be either Valid or Invalid (Two states).
+The blue Gene/P uses write-through for both L1 and L2 and does not have any hardware CCP the cache lines can be Valid or Invalid (Two states).
 
@@
 What is a flit (in general, not in the context of any specific network)? Is it different from a packet? Explain. 2 points\\}}
-Flow control unit, flit, is the smallest part of a packet that get transmitted on a interconnect. A flit gets dropped in the case of congestion and thus has to be retransmitted.
+Flow control unit, flit, is the part of a packet that gets blocked in the case of congestion. In a cut-through switching a flit is the entire packet.
 
 \section*{Q20}
-What network topology was used in the Connection Machine CM-5? Why? 5 points
-
-%hillis_SIMD_scalable_supercomputer
-Fat tree (figure 4).
+\large{\textbf{What network topology was used in the Connection Machine CM-5? Why? 5 points\\}}
+The data network topology of the CM-5 is Fat tree which is used as it scales well with the number of nodes.
+For sending packets around the network a packet-switched router is used and adaptive load balancing that together minimizes the network congestion and makes it deadlock-free.
+A fat tree's bandwidth naturally increases with the number of nopes which makes it perfect for a scalable application.
+It also have good fault tolerant features.
 
@@
 in a tree interconnect the bisection bandwidth would be 1u as we would cut the tree at the root were only one vertex exist.
-In DASH each nodes memory bus is connected to a global bus, and when a remote request is issued the local bus will be traversed twice (local->global->local) and thus the traffic on the local bus will be twice as high.
-This means that the local bandwidth will be more important than the bisection brandwidth.
+In DASH each nodes memory bus is connected to a global bus, and when a remote request is issued the local bus will be traversed twice (local-$>$global-$>$local) and thus the traffic on the local bus will be twice as high.
+This means that the local bandwidth will be more important than the bisection bandwidth.
 
 \section*{Q22}
-\large{\textbf{
-In what types of application areas are VLIW processors employed today? 2 points\\}}
+\large{\textbf{In what types of application areas are VLIW processors employed today? 2 points\\}}
 In processing of digital signals (DSPs) and multimedia.

### 2014-10-30 lab/report/include/lab2.tex
corrections
stuff

@@
-\section{Lab 2}
+\section{Laboratory assignment 2}
 \label{sec:lab2}
@@
 To add the Owner (O) state to the existing MESI CCP, only small changes are needed.
-In \textit{readLine()} and \textit{remoteWriteAction()} no changes are needed.
+In \textit{readLine()} and \textit{writeRemoteAction()} no changes are needed.
 Turning to \textit{readRemoteAction()}, transitions from the E, M, and O states to the O state are added, and all of these transitions provide data to the calling cache.
@@
 For the cache access statistics there is no need for change.
+\textcolor{red}{One additional change needed to the MESI protocol is that when a cache line is in the O state and gets replaced the owner ship must be returned to the memory.}
+
+\textcolor{red}{We assume that the next level cache, the LLC, does not implement any CCP as there is no other caches on the same level that needs to be synchronized with.
+Because of this, change of ownership is not explicitly performed in any of the \textit{readLine(), writeLine(), readRemoteAction(), or writeRemoteAction()} functions.
+This as, when the line gets replaced, the state information gets overwritten with the replacing line's state information.}
 
 \par\noindent
+
 \begin{minipage}[t]{.5\textwidth}
-\begin{lstlisting}[language=C,frame=lrtb]
-In readLine()
+
+	\begin{lstlisting}[language=C,frame=lrtb,style=base]
+In ¤readRemoteAction()¤
 	if(MODIFIED) {...}
@@
 
-In writeRemoteAction()
+In ¤writeLine()¤
 	if(SHARED) {...}
@@
 \begin{minipage}[t]{.5\textwidth}
-\begin{lstlisting}[frame=lrtb]
-In readLine()
+
+	\begin{lstlisting}[language=C,frame=lrtb,style=base]
+In ¤readRemoteAction()¤
 	if(MODIFIED || EXCLUSIVE || OWNER) 
@@
 
-In writeRemoteAction()
+In ¤writeLine()¤
 	if(SHARED || OWNER) {...}

### 2014-10-30 lab/report/include/lab2.tex
corrections

@@
 For the cache access statistics there is no need for change.
-\textcolor{red}{One additional change needed to the MESI protocol is that when a cache line is in the O state and gets replaced the owner ship must be returned to the memory.}
-
-\textcolor{red}{We assume that the next level cache, the LLC, does not implement any CCP as there is no other caches on the same level that needs to be synchronized with.
-Because of this, change of ownership is not explicitly performed in any of the \textit{readLine(), writeLine(), readRemoteAction(), or writeRemoteAction()} functions.
-This as, when the line gets replaced, the state information gets overwritten with the replacing line's state information.}
+\textcolor{red}{One additional change needed to the MESI protocol is that when a cache line in the O state gets replaced, the ownership must be returned to the memory.
+The change of ownership is not explicitly performed in any of the \textit{readLine(), writeLine(), readRemoteAction(), or writeRemoteAction()} functions.
+This as, when the line gets replaced, the state information is overwritten with the replacing line's state information, and thus is the ownership transferred to the LLC.}
 

### 2014-10-31 paper/include/summary.tex
more summary

@@
 
-All the techniques surveyed fits into one of the approaches based on four types of information, Retention Time, Access Recency, Data Tolerance, and Data Validity, mentioned in the paper.
+All the techniques surveyed fits into one of the approaches based on four types of information, Retention Time, Access Recency, Data Tolerance, and Validity, mentioned in the paper.
 The relative performance of the techniques is compared, and some projections for the future concerning higher density and bigger capacity DRAMs are discussed.
+
+The future importance of trying to mitigate DRAM power is pressed upon, and refresh power are a large, and increasing, part of this.
+Some new techniques for increasing DRAM capacity, like 3D stacked DRAM, will have problems being competitive if refresh power is not decreased.

### 2014-10-31 paper/include/refrint.tex
new version

@@
 For the hot rows, an approach similar to Smart Refresh is employed.
-The differences are that Refrint does not maintain a counter for each row, but a snapshot of a global counter as well as the row's valid bit.
+The differences are that Refrint does not maintain a counter for each row, but a snapshot of a global counter's highest bits as well as the row's valid bit.
 When a row is accessed, the corresponding snapshot is updated with the current value of the global counter.
@@
 
-\todo[inline]{Drafty below. Ask Sally, confusing paper.}
+Refrint modifies the MC.
+The $N$ snapshot bits and a valid bit per DRAM row is stored in the MC and from experiments, Agrawal et al.
+found out that one or two are good values for $N$.
+To process the policies for the cold rows, decision logic as well as Count bits and State bits for each cache row is needed.
+Where the bits and logic resides is unclear to us, but they are placed outside the MC.
+In simulations, Agrawal et al.
+uses five Count bits, which give a negligible overhead and two bits for $N$.
+Of this, the total overhead became $0.005\%$, with a row size of \textit{8~KB}.
 
-Refrint modifies the MC and also stores data on chip(?).
-In an unmodified MC, there already exist a counter which can be used as the global counter and the authors have chosen to store the $N$ snapshot bits in the MC as well as the rows' valid bits in the MC.
-From experiments, the authors found out that one or two are good values for $N$.
-The logic to process the policies for the cold rows consists of decision logic, as well as Count bits and State bits for each cache row.
-This logic is located on chip(?)/in DRAM(?)(???) and has apparently a negligible overhead.
-In simulations, $N$ was two bits and Count was 5 bits, giving a total overhead of $0.6\%$, when cache row size is 64 bytes.
-
-Presents the results. Add that we can not really use the result as they are from eDRAM.
+Results indicated that Refrint lower the total power consumption by $70\%$ but increases execution time of a cache hierarchy when evaluated as eDRAM, compared to an ordinary cache hierarchy using SRAM.
+Even though Refrint speeds up the execution time compared to an eDRAM hierarchy without Refrint, the execution time is increased by $6\%$  with respect to the SRAM solution, due to the characteristics of eDRAM.

### 2014-10-31 paper/include/smart_refresh.tex
new version

@@
 The counters are decremented in such a fashion so the refreshes are distributed evenly over time and thus avoiding burst refresh.
-To achive this, the counters are initially staggered and divided into n logical blocks where the counters are accessed in a sequential order.
-Only $n$ counters will be accessed, set or decremented at the same point in time, which result in that maximum $n$ rows are refreshed simultainously.
+To achieve this, the counters are initially staggered and divided into n logical blocks where the counters are accessed in a sequential order.
+Only $n$ counters will be accessed, set or decremented at the same point in time, which result in that maximum $n$ rows are refreshed simultaneously.
 
@@
 A higher bit size results in that an access can be captured later in time, which effectively postpones the refresh further.
-The least optimal case for postponing the refresh is if the row access occurs a small time intervall before the counter is accessed.
+The least optimal case for postponing the refresh is if the row access occurs a small time interval before the counter is accessed.
 If a two-bit counter is used the least optimal row access result in that the row is refreshed after $3/4$ of the maximum refresh period.
@@
 The technique was simulated on \textit{2~GB} of main memory.
-Result shows that the refresh reduction was in average $59\%$ and refresh power consumption savings was on average $52\%$, all without any performance degradation.
+Results shows that the refresh reduction was on average $59\%$ and refresh power consumption savings became on average $52\%$.
+This produced a total DRAM power saving of $12\%$, without any performance degradation.

### 2014-10-31 paper/include/summary.tex
some summary. DTAIL POWER!!!!!

--- 2014-10-23 paper/include/summary.tex
@@
 \label{sec:sum}
-SumSumSUm!! my precious!
+*conducted survey
+
+*dram power importance
+
+*approaches
+
+*our ideas
+
+Of the techniques surveyed DTail by Cui et al.
+stands out by showing promise for being a platform on which combining many kinds approaches is possible.
+Even as some of the approaches brought up in this paper shows great potential for tackling the problem with refresh power, it is important to note that they are not the only possible way.
+Other ways not surveyed, like scheduling of DRAM access, could also play an important role.

### 2014-10-31 paper/include/summary.tex
hej

@@
 \label{sec:sum}
-*conducted survey
-
-*dram power importance
-
-*approaches
-
-*our ideas
-
-Of the techniques surveyed DTail by Cui et al.
-stands out by showing promise for being a platform on which combining many kinds approaches is possible.
-Even as some of the approaches brought up in this paper shows great potential for tackling the problem with refresh power, it is important to note that they are not the only possible way.
-Other ways not surveyed, like scheduling of DRAM access, could also play an important role.
+We have conducted a survey comprising of several proposed techniques to optimize DRAM refresh power.
+The survey has been performed  These techniques are based on one or several of the four approaches; Access Recency, Retention Time, Data Tolerance and Validity.
+DRAM refresh power is an increasing part of computing systems and to be able to mitigate this trend

### 2014-10-31 paper/include/summary.tex

@@
 \label{sec:sum}
-We have conducted a survey comprising of several proposed techniques to optimize DRAM refresh power.
-The survey has been performed  These techniques are based on one or several of the four approaches; Access Recency, Retention Time, Data Tolerance and Validity.
-DRAM refresh power is an increasing part of computing systems and to be able to mitigate this trend
+Due to the increasing importance of lowering power consumption in modern computer systems and as DRAM refresh power is an increasing part of computing systems many techniques targeting DRAM power has been proposed.
+We have conducted a survey comprising of several of these proposed techniques that optimize DRAM refresh power.
+The surveyed techniques are based on one or several of the four approaches, Access Recency, Retention Time, Data Tolerance and Validity.
+The approaches which we believe have higher future potential are Retention Time and Validity.
+This opinion is based on the fact that those approaches are more capable to handle the DRAM trends and provide good power reduction potential compared to the approaches' design complexity.
+
+Findings from the survey shows that the best way to optimize DRAM refresh is to combine several of the approaches.
+Of the techniques surveyed, DTail by Cui et al.
+stands out by showing promise by providing a framework where it is possible to combine different approaches.
+Even though some of the approaches brought up in this paper shows great potential for tackling the problem with refresh power, it is important to note that their use only are a part of the solution.
+Other ways not surveyed, like scheduling of DRAM access, will also play an important role.

### 2014-10-31 paper/include/background.tex
corrections

@@
 All ranks on one channel can operate in parallel, but are constrained by the channel bandwidth, which is shared among the ranks.
-Further, the banks within the ranks operate in parallel, but are constrained by both the shared channel bandwidth and resources shared among the banks.
+Further, the banks within the ranks operate in parallel, but are constrained by both the shared channel bandwidth and resources shared among the banks, such as power.
 
@@
 When the word line is activated, the transistors connect the capacitors to the bit lines.
-When the capacitors connect to the bit lines, charge will either flow from or to the capacitors depending on whether they were charged or not, respectively.
+Then, charge will either flow from or to the capacitors depending on whether they were charged or not, respectively.
 For this to work, the bit lines have to be precharged to $V_{DD}/2$.
-The sense amplifiers then detect the change of voltage, which makes it possible to interpret the stored data.
+The sense amplifiers then detect the change in voltage, which makes it possible to interpret the stored data.
 The charge stored on the capacitors has, as mentioned, left the capacitor and a read operation is thus destructive.
-To maintain the data the capacitors are recharged by the bit lines, which are driven to either $V_{DD}/2$ or $0\:V$ by the sense amplifiers depending on the former voltage detected.
+To maintain the data the capacitors are recharged by the bit lines, which are driven to either $V_{DD}$ or $0\:V$ by the sense amplifiers depending on the former voltage detected.
 Finally, when another row on the bank is accessed, the open row has to be closed, which disconnects the capacitors from the bit lines.
@@
 
-As mentioned, the capacitors leak charge over time and have to be recharged in order to preserve the data.
-The refresh is performed by opening a row to let the sense amplifiers magnify the detected voltage change and thus recharge the cells to either $V_{DD}/2$ or $0\:V$.
+As mentioned, the capacitors leak charge over time and have to be refreshed in order to preserve the data.
+The refresh is performed by opening a row to let the sense amplifiers magnify the detected voltage change and thus recharging the cells to either $V_{DD}$ or $0\:V$.
 This follows the same procedure as a read operation and thus is the row always refreshed when a read is performed.

### 2014-10-31 paper/include/introduction.tex
corrections

@@
 This refreshing has to be done as DRAM's store data as a charge in a capacitor, and these charges leak over time.
-To increase DRAM performance and lower access latencies, 3D stacked DRAM has been proposed, but stacked design have trouble dissipating heat.
+To increase DRAM performance and lower access latencies, 3D stacked DRAM has been proposed, but a stacked design have trouble dissipating heat.
 The higher temperature increases leakage, and the DRAM will have to be refreshed more often.
@@
 As such, one straightforward way to mitigate DRAM power consumption is to reduce the number of unnecessary refresh operations.
-Here we survey some proposed approaches to avoid or delaying refresh.
+Here we survey some proposed techniques to avoid or delaying refresh.
 Smarter scheduling, while promising interesting, is beyond the scope of this paper.

### 2014-10-31 paper/include/reducing_refresh_power.tex
disc text
corrections

@@
 \label{sec:red}
-There are several approaches to decrease the refresh power by reducing unnecessary refreshes and these can be categorized depending on with sort of information the scheduling decisions are based upon.
+There are several approaches to decrease the refresh power by reducing unnecessary refreshes and these can be categorized depending on which sort of information the scheduling decisions are based upon.
 The four types of information are: access recency, cell retention time, error tolerance of the data, and line validity.
@@
 Retention time is a measurement of the time it takes before a cell loose its data if not refreshed (recharged).
-Variations in the manufacturing process causes a exponentially distributed fluctuation \cite{katayama} in the retention time of the DRAM cells on the same chip.
-To accommodate for the cells with the lowest retention time manufacturers produce chips with average retention times much higher than possible minimum.
+Variations in the manufacturing process causes a exponentially distributed fluctuation \cite{katayama} in the retention time of the DRAM cells.
+To accommodate for the cells with the lowest retention time, manufacturers produce chips with average retention times much higher than possible minimum.
 This gives room for optimization using variable refresh rate for different cells.
@@
 \subsubsection*{\textbf{Data Tolerance (T)}}
-In the most approaches all data loss is seen as intolerable.
-In this approach data is categorized depending on how critical it is and e.g.
-approximate computations, games, or media processing can tolerate some bit errors while still producing a acceptable result.
-This yields that cells containing data from such applications may be refreshed more seldom.
-Using this approach the energy and performance savings achieved is highly depending on what kind of applications is most commonly run on the system.
+In the other approaches all data loss is seen as intolerable.
+In this approach data is categorized depending on how critical it is, which yields that cells containing data with a lower criticality may be refreshed more seldom.
+Applications such as approximate computations, games, or media processing can tolerate some bit errors while still producing a acceptable result.
+Using this approach, the energy and performance savings achieved is highly depending on what kind of applications most commonly are run on the system.
 
@@
 In a system that does not utilize all the available memory space it is unnecessary to refresh parts not used.
-Therefore, cells containing data that are not valid does not need to be refreshed.
+Therefore, cells that do not keep valid data do not need to be refreshed.
+The allocation and deallocation of the cells have to be tracked to get the information needed for implementation.
 This approach is highly sensitive to memory utilization of the system and its effects on energy consumption and performance improvements will be diminishing on systems with high workload.

### 2014-10-31 paper/include/dtail.tex
iFinished! yaya!

@@
 workload and type of application.
-How acquisition of the \textit{refresh data} is to be performed is left as future work.
-Refresh data is kept per row and is stored in the DRAM itself instead of in the MC which many other solutions do \cite{raidr}\cite{smartrefresh}\cite{refrint}.
+The R and V information, called\textit{refresh data}, is kept per row and is stored in the DRAM itself instead of in the MC which many other solutions do \cite{raidr}\cite{smartrefresh}\cite{refrint}.
 To counter the added latency of a slower memory (DRAM versus fast memory in the MC), Cui et al.
@@
 As the refresh data is kept on a row level DTail can save more refreshes than other techniques that have a lower granularity.
+How to actually implement the acquisition of the \textit{refresh data} is not stated, but Cui et al.
+proposes that a few modifications to the OS would work well, as the OS can manage the refresh data residing in DRAM.
 

### 2014-10-31 paper/include/refrint.tex
corrections

@@
 \textit{WB(n,m)} refreshes a dirty row $n$ times before flushing it and changing the row state to valid clean, then the row is refreshed $m$ times after the last access before it is invalidated.
-Moreover, \textit{WB} is implemented using so called \textit{Count} bits, which is set either to $n$ or $m$ and is decremented upon refresh.
-% Mention the State bits as well?
+Moreover, \textit{WB} is implemented using so called \textit{Count} bits, which is set either to $n$ or $m$ and is decremented upon refresh, as well as a \textit{State} bit.
 
@@
 found out that one or two are good values for $N$.
-To process the policies for the cold rows, decision logic as well as Count bits and State bits for each cache row is needed.
-Where the bits and logic resides is unclear to us, but they are placed outside the MC.
+To process the policies for the cold rows, decision logic as well as Count bits and State bit for each cache row is is needed.
+The decision logic is placed outside the MC, loosely coupled with it, whereas the Count and State bits are stored in eDRAM logically with the data.
 In simulations, Agrawal et al.
-uses five Count bits, which give a negligible overhead and two bits for $N$.
-Of this, the total overhead became $0.005\%$, with a row size of \textit{8~KB}.
+set Count to five bits and $N$ to two bits, resulting in a total overhead of $0.015\%$ if row size is \textit{8~KB}.
 

### 2014-10-31 paper/include/dtail.tex
layout

@@
 workload and type of application.
-The R and V information, called\textit{refresh data}, is kept per row and is stored in the DRAM itself instead of in the MC which many other solutions do \cite{raidr}\cite{smartrefresh}\cite{refrint}.
+The R and V information, called \textit{refresh data}, is kept per row and is stored in the DRAM itself instead of in the MC which many other solutions do \cite{raidr}\cite{smartrefresh}\cite{refrint}.
 To counter the added latency of a slower memory (DRAM versus fast memory in the MC), Cui et al.
